---
title: "Time series change points"
date: "`r format(file.mtime(knitr::current_input()), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time series change points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
if (requireNamespace("pkgload", quietly = TRUE) && pkgload::is_dev_package("mcp")) {
  pkgload::load_all(quiet = TRUE)
}
bayesplot::bayesplot_theme_set(bayesplot::theme_default(base_size = 11, base_family = "sans"))
```

Serial dependence is common in time series. Add autoregressive terms with `ar(p)`, moving-average terms with `ma(q)`, or both in the segment formulas:

``` r
model = list(y ~ 1 + x + ar(1) + ma(1))
```

The most common use case is still `ar(1)`. AR and MA components are active only where declared, so omitting a component turns it off; `ar(0)` and `ma(0)` remain accepted explicit turn-off forms. In contrast, a positive-order `ar(p, 0)` or `ma(q, 0)` joins the preceding intercept/local-$x$ level for matching lags. An `ar(p)` or `ma(q)` declaration replaces that whole component: if `ar(1)` follows `ar(2)`, the lag-2 coefficient is removed in the later segment. Both accept a regression formula, such as `ar(1, 1 + x)` or `ma(1, 0 + x)`.

## GARMA definition

`mcp` implements AR and MA as a generalized ARMA (GARMA) recurrence on the response-family link scale. If $\eta^{\text{reg}}_t$ is the ordinary regression predictor from the segment formulas and $\eta_t$ is the predictor including serial dependence, then

$$\begin{aligned}
\text{AR}_t &= \sum_{j=1}^{p} \phi_{j,t} \left[g(y^*_{t-j}) - \eta^{\text{reg}}_{t-j}\right] \\
\text{MA}_t &= \sum_{k=1}^{q} \theta_{k,t} \left[g(y^*_{t-k}) - \eta_{t-k}\right] \\
\eta_t &= \eta^{\text{reg}}_t + \text{AR}_t + \text{MA}_t
\end{aligned}$$

where $\phi_{j,t}$ is the lag-$j$ autoregressive (AR) coefficient at time $t$, $\theta_{k,t}$ is the lag-$k$ moving-average (MA) coefficient at time $t$, $g(\cdot)$ is the link function ($g(\mu_t) = \eta_t$, so $\mu_t = g^{-1}(\eta_t)$), and $y^*_t$ is the threshold-constrained observation with threshold constant $c$ (set via `threshold = 0.1` by default). Thus `ar()` uses lagged link-scale residuals relative to the ordinary regression, while `ma()` uses lagged one-step innovations.

**`mcp` does regression with change points on the coefficients** $\phi_{j,t}$ **and** $\theta_{k,t}$. See below for applied examples.

Unavailable lags at the beginning of the series contribute zero. This same recurrence is used by JAGS, fitted values, predictions, log likelihoods, and fresh-series simulation.

A key feature of this recurrence is that AR/MA memory flows continuously across change points:

- For an order-$N$ component in a new segment, the last $N$ observations *before* the change point are input into the first $\eta_t$ in the new segment, weighted by the new segment's AR/MA parameters.

- AR/MA lags do not reset between segments; memory flows continuously from one segment into the next. Lags only reset at the very beginning of the dataset, or between independent series when using `series = "column_name"` (see [Multiple time series](#multiple-series)).

- Because change point locations $\tau$ are estimated with posterior uncertainty, the change point where AR/MA parameters switch varies conditionally across MCMC draws.

GARMA currently supports only the default links for `gaussian()` (identity), `binomial()` (logit), `bernoulli()` (logit), `poisson()` (log), and `negbinomial()` (log). Non-default links are rejected for now.

**Warning:** The AR and MA coefficients are modeled directly and are not jointly constrained to the stationary or invertible regions. With higher-order terms such as `ar(2)`, independently bounded coefficients can violate the usual root conditions. Coefficients that change across predictors or segments, such as `ar(1, 1 + x)` or `model = list(y ~ ar(1), ~ ar(1))`, instead define a time-varying process to which the usual constant-coefficient conditions do not directly apply. The same applies to `ma()`.

### Observation threshold

The transformed observation $y^*_t$ keeps log and logit residuals finite when observations lie on the domain boundary of the link function, using a threshold constant $c$ (default `threshold = 0.1`). For Poisson and negative-binomial models, zero counts are replaced by $c$, so $y^*_t = \max(y_t, c)$. For binomial models, observed success counts are constrained to $[c, n_t - c]$ before conversion to a rate ($n_t = 1$ for `bernoulli()`, constraining $y^*_t$ to $[0.1, 0.9]$). It has no effect for Gaussian models.

The default should usually be left unchanged. If needed, set it on either term, for example `ar(1, threshold = 0.01) + ma(1)`. AR and MA share one threshold within a segment, and the threshold may differ between segments. Supplying different AR and MA thresholds in the same segment is an error.

# Simple example

Let's simulate some data using the `mcp_example()` function with the "ar" settings:

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic

ar_data = mcp_example_data("ar")
head(ar_data)
```

See how this was generated in `fit$example_code`. We model this as a plateau (`1`) with a second-order autoregressive residual (`ar(2)`) followed by a joined slope (`0 + time`) with a negative first-order autoregressive residual (`ar(1)`):

```{r, results = FALSE, warning=FALSE, message=FALSE}
model = list(
  price ~ 1 + ar(2),  # Intercept_1, ar1_1, ar2_1
  ~ 0 + time + ar(1)  # time_2, ar1_2; replaces AR(2) with AR(1)
)
fit = mcp(model, ar_data, seed = 42)
```

Let's plot it and we see that AR was strong in the first segment and weaker-but-negative in the second:

```{r}
set.seed(42)
plot(fit)
```

We can summarise the inferred coefficients:

```{r, R.options = list(width = 9999)}
summary(fit)
```

The naming syntax is `[component][order]_[segment]` for intercepts. For example, `ar1_2` is the first-order autoregressive coefficient and `ma1_2` the first-order moving-average coefficient in segment 2. Slopes use the usual `mcp` parameter names, e.g., `ar1_x_3` for a slope on AR(1) in segment 3.

Comparing the columns `mean` and `sim` we see that the AR coefficients are reasonably recovered. In fact, the posterior mean is almost always exactly the same as `arima(data, order = c(N, 0, 0))` (see below), so the non-perfect fits are due to randomness in the simulation - not in the fit.

For Gaussian GARMA models, `sigma` describes the *innovations*, i.e., the part of the residuals not explained by AR and MA coefficients. `sd(fit$data$price)` is therefore generally higher. In this case, the SD of raw data in the plateau is `r round(sd(fit$data$price[1:120]), 2)`. As always, it is good to assess posteriors and convergence more directly:

```{r, fig.height=9, fig.width=8}
set.seed(42)
plot_pars(fit, nvariables = NULL)
```

Sometimes, the trace plot shows that the change point (`cp_1`) is not well identified with this model and data. As discussed in the article on [tips, tricks, and debugging](../articles/tips.html), you could combine a more informative prior with more samples (`mcp(..., warmup = 10000, iter = 10000)`), if this is a problem.

You can do hypothesis testing with GARMA models using `hypothesis()`. [Read more here](../articles/comparison.html) or scroll down for an applied example.

# Tips, comments, and warnings

AR/MA must be declared in every active segment. Repeating `ar(2)` estimates new lag coefficients; `ar(2, same(1))` shares both lag intercepts from the preceding segment, and errors if either lag lacks an active source. An explicit `as` can select an earlier source across a disabled segment. `ar(2, 0)` joins the preceding level; `ar(2, 0 + same(x))` also shares the local slope for each lag. These declarations retain observation and innovation history across segment transitions.

The AR and MA terms apply to link-scale *residuals* from the ordinary regression. In time-series jargon, this is a *dynamical* regression model where the ordinary regression parameters make up the *deterministic* structure. See further comments in the section on priors below.

To control the *direction* of a change, put a one-sided prior on the corresponding coefficient, e.g., `ar1_x_1 = "dnorm(0, 1) T(0, )"`. See recommendations in [the section on priors](#ar_prior).

## Regression on AR/MA coefficients

You can specify how the coefficients change with $x$ using `ar(p, formula)` and `ma(q, formula)`. For example, `ar(1, 1 + time)` models a steady change in AR(1) strength. These regression formulas use an identity link, so slopes can easily produce non-stationary or non-invertible coefficients. See [the section on priors](#ar_prior) for ways to constrain them. The same formula is applied separately to every order in the term.

Population-level [`mcp` formula syntax](../articles/formulas.html) is available inside both terms, so you could use `ar(3, 1 + I(x^2) + exp(z))` or `ma(1, 0 + group)`. Group-level terms such as `(1|id)` are not supported inside `ar()` or `ma()`. Transformed predictors must stay within the transformation's domain; for example, values passed to `log()` and `sqrt()` must be non-negative.

## Combining ar(), ma(), and sigma()

You can combine `ar()` and `ma()` with any regression model and with [group-level change points](../articles/group_effects.html). For Gaussian models, `sigma()` controls the innovation standard deviation:

```{r}
model = list(
  y ~ 1 + ar(1) + ma(1) + sigma(1),  # First order ARMA
  ~ 0 + x + ar(1) + sigma(1),        # New AR coefficient; no MA
  ~ 1  # No ARMA modelled in this segment
)
```

## Row ordering and multiple time series (`series`) {#multiple-series}

AR and MA lags apply strictly to the *order of rows* in your data frame, without taking into account the numerical distance between values of `x`. This has two important implications:

- **Sort by time:** Adjacent rows are treated as lag-1 neighbors. Always ensure your data is sorted by time before fitting: `data = data[order(data$x), ]`.
- **Uneven spacing:** Points recorded seconds apart vs. months apart will be modeled with the exact same autocorrelation if they are placed in consecutive rows.

### Panel and longitudinal data (`series`)

When data contains repeated measurements from multiple participants, sessions, or locations, you do *not* want residual autocorrelation from the end of one individual to carry over to the beginning of the next. Use the `series` argument inside `ar()` or `ma()` to define where independent series start and end:

```{r, message=FALSE}
# 1. Model: declare series = id once
model = list(
  y ~ 1 + ar(1, series = id),  # Segment 1: AR(1) within series
    ~ 0 + time + ar(1)          # Segment 2: new AR(1); series = id applies globally
)

# 2. Simulate panel data for 3 subjects (must be contiguous and sorted by time)
df_panel = data.frame(
  id = rep(c("Subject_A", "Subject_B", "Subject_C"), each = 100),
  time = rep(1:100, times = 3),
  y = 0
)

empty = mcp(model, data = df_panel, sample = FALSE)
set.seed(42)
df_panel$y = empty$simulate(
  empty, df_panel,
  cp_1 = 30,         # Change point location
  Intercept_1 = 10,  # Intercept in segment 1
  time_2 = 0.1,      # Slope for the mean in segment two
  ar1_1 = 0.3,       # First-order AR in segment 1
  ar1_2 = 0.8,       # First-order AR in segment 2
  sigma_1 = 1.5      # Residual sigma; sigma(0) is implicit in segment 2 where not declared
)

# 3. Fit the model
fit = mcp(model, data = df_panel)
print(summary(fit))

# Plot fit and ar(1) parameter
library(patchwork)
gg_trend = plot(fit, color_by = "id")
gg_dpar  = plot_dpar(fit, "ar1", q_fit = TRUE)
gg_trend / gg_dpar
```

Within each series, AR/MA memory flows continuously across change points. Between series, memory resets to zero so individuals remain completely independent.

# Simulating autocorrelated change point data

Assessing the correctness of autocorrelation is less intuitive than seeing e.g. a mean fit. However, we can verify `mcp` up against more tested-and-tried functions such as `arima()` in base R. Let us simulate a single AR(3) segment, i.e., without change points, and see if it fits:

```{r}
# Model
model = list(response ~ 1 + ar(3))

# Simulate data
df = data.frame(time = 1:200, response = 0)
empty = mcp(model, df, sample = FALSE, par_x = "time")
set.seed(42)  # For consistent "random" results
df$response = empty$simulate(
    empty,
    df,
    Intercept_1 = 20,
    ar1_1 = 0.7, 
    ar2_1 = 0.2, 
    ar3_1 = -0.4, 
    sigma_1 = 8)

# Base arima AR(3)
arima(df$response, order = c(3, 0, 0))
```

OK, we can see that the `ar` coefficients and sigma ($sigma = sqrt(sigma^2)$) are simulated correctly, if taking `arima()` as ground truth. Inferring with `mcp` is straightforward:

```{r, results = FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, df, par_x = "time", seed = 42)
```

The Bayesian parameter estimates are in perfect correspondence with `arima()`, even where they deviate a tiny bit from the simulation parameters (due to the inherent randomness in simulating data):

```{r, R.options = list(width = 9999)}
summary(fit)
```

# Inferring an autocorrelation-only change

One "side-effect" of the `mcp` implementation of autocorrelation using `ar()` *in the formulas* is that you can infer when autocorrelation parameters and structures change.

Let's simulate a change point in autocorrelation and see if we can infer it later:

```{r, message=FALSE}
# The model
model = list(
  y ~ 1 + x + ar(1),
  ~ 0 + same(x) + ar(1)  # retain the same slope
)

# Get predictions
df = data.frame(x = seq(0, 100, length.out = 200), y = 0)
empty = mcp(model, df, sample = FALSE)
set.seed(42)
df$y = empty$simulate(
  empty,
  df,
  cp_1 = 60,
  Intercept_1 = 20,
  x_1 = 1,
  ar1_1 = 0.8, ar1_2 = 0.2,
  sigma_1 = 5)
```

We use `same(x)` in segment 2 to keep the slope identical across segments, so the change point is strictly in autocorrelation. We use `sample = "both"` because we will do a Savage-Dickey test later.

```{r, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, iter = 8000, sample = "both", seed = 42)
```

Let's plot the full model prediction using `plot(fit)`. You could use `plot(fit, geom_data = "line")` for a more classical line plot of the time series data. Set `plot(fit, arma = FALSE)` to omit all AR and MA effects.

We plot it together with the change in the `ar1` parameter using `plot_dpar(fit)`:

```{r}
library(patchwork)
set.seed(42)
plot(fit) /  # Patchwork syntax to show on separate rows
  plot_dpar(fit, dpar = "ar1", lines = 100)
```

We recovered the parameters, including the change point (see `mean` but also the helpful `sim` and `match` columns):

```{r, R.options = list(width = 9999)}
summary(fit)
```

We can also plot some of the parameters. As usual, we see that the change point is not well defined by any known distribution. The fact that the posterior mean is around 60 does not (necessarily) mean that there is a high credence in this value. Usually, I find that any bi- or N-modality on the posterior matches well with what you would guess from looking at the raw data. As they say: Bayesian inference is common sense applied to data.

```{r}
set.seed(42)
plot_pars(fit, regex_pars = "cp_1|ar_*")
```

As usual, we can test hypotheses ([read more here](../articles/comparison.html)). We can also ask how much more likely it is (relative to the prior) that the two autocorrelations are equal compared to them differing. Because we sampled both the prior and posterior (`mcp(..., sample = "both")`), we can do a Savage-Dickey density ratio test:

```{r, R.options = list(width = 9999)}
hypothesis(fit, "ar1_1 = ar1_2")
```

Notice the warning. We used default priors, so the Savage-Dickey ratio is quite meaningless because the prior does not represent our actual pre-likelihood beliefs. In any case, if it did, we would conclude that evidence for equality is very small so it was rarely visited by the sampler and hence the precision is low (as the second warning show).

Of course, we can also do directional tests. For example, what is the evidence that `ar1_1` is more than 0.3 greater than `ar1_2`? Answer: More than 100 to one.

```{r, R.options = list(width = 9999)}
hypothesis(fit, "ar1_1 - 0.3 > ar1_2")
```

# Priors on AR and MA coefficients {#ar_prior}

The default prior on each AR and MA intercept is a truncated normal distribution:

``` r
dnorm(0, 0.5) T(-1, 1)
```

It is symmetric around zero and gently shrinks away from the $[-1, 1]$ bounds: its central 95% interval is approximately $[-0.84, 0.84]$. This is a regularizing prior on each direct coefficient, not a joint stationarity or invertibility constraint for orders above one.

`brms` also models AR and MA coefficients directly within $[-1, 1]$, but uses flat default priors over that range. Thus `mcp` keeps the familiar parameter interpretation while adding mild regularization toward zero; fits should generally be similar when the likelihood is informative.

Categorical contrasts default to `dnorm(0, 0.25)`. Numeric slopes are also normal and scaled so that one representative predictor change has prior SD 0.25. Thus approximately 95% of the prior change lies within $\pm 0.5$.

The defaults express no assumption about the sign. For a time series where positive first-order autocorrelation is expected, alternatives include `ar1_1 = "dunif(0, 1)"` or `ar1_1 = "dnorm(0.5, 0.5) T(0, 1)"`. [Read more about specifying and checking priors](../articles/priors.html).

Here is a complete list of the (default) priors in the model above:

```{r, R.options = list(width = 9999)}
prior_summary(fit)
```

We can also visualize the priors because we sampled the prior. `prior = TRUE` works in most `mcp` functions, including `plot()` and `summary()`.

```{r, fig.height=10, fig.width=7}
set.seed(42)
plot_pars(fit, prior = TRUE, nvariables = NULL)
```

Notice that the plot smooths the posteriors at sharp cutoffs, slightly misrepresenting the true distribution.

Let's inspect the priors for a more advanced AR model, since you would often have to inform these:

```{r, R.options = list(width = 9999)}
model = list(
  y ~ 1 + ar(2, 1 + x),
  ~ 0 + ar(1, 1 + I(x^2))
)
empty = mcp(model, data = data.frame(y = 1:10, x = 1:10), sample = FALSE)
prior_summary(empty)
```

AR and MA coefficient regressions use an identity link, while their residual inputs use the supported response-family link described above. Careful priors become especially important with higher orders or coefficient slopes, where individual values in [-1, 1] do not ensure a stable recurrence. After fitting, `mcp()` checks up to 500 posterior draws at up to 100 observed predictor rows and warns if the checked-draw violation rate exceeds the corresponding `ar` or `ma` threshold in `diagnostics` (10% by default). `fit$simulate()` checks the supplied coefficient trajectory before generating a fresh series. For predictor- or segment-varying coefficients these are local smoke tests, not proofs of global stationarity or invertibility.

Here are a few ways in which you may want to inform the AR or MA parameters:

- For a constant AR(2) process, stationarity requires $-1 < \phi_2 < 1$ and $\phi_2 - 1 < \phi_1 < 1 - \phi_2$. One dependent-prior construction that stays in this region is `ar2_1 = "dunif(-1, 1)"` together with `ar1_1 = "dunif(ar2_1 - 1, 1 - ar2_1)"`. The simpler suggestion `ar2_1 = "dunif(0, ar1_1)"` is not sufficient. Higher orders require the corresponding joint root condition.
- Slopes can quickly make AR coefficients non-stationary or MA coefficients non-invertible. Constrain their magnitude with reference to a plausible change over the predictor range. For example, a shallow negative change over the observed x-span could use `"dnorm(0, 0.1 / (max(x) - min(x))) T(, 0)"`.

# Notes on observed data vs simulated data

JAGS, `fitted()`, `predict()`, and `log_lik()` evaluate a finite, conditional recurrence: the response history supplies earlier residuals, and unavailable residuals before the start of the series are set to zero.

- `fitted()` and `predict()` use posterior imputations only to complete missing response histories. `predict()` generates fresh outcomes at all rows, even where the original response was missing.
- `log_lik()` is unavailable if the data has missingness before some observed data.
- `predict(..., conditional = FALSE)`, `posterior_predict(..., conditional = FALSE)`, and `pp_check()` generate each series recursively from the model posteriors without conditioning on the original response history. This makes serial summaries of posterior replications, such as their autocorrelation and run lengths, meaningful.

# JAGS code

Here is the JAGS code for the second simulation example, i.e., the one with a single slope going from AR(2) to AR(1). You can print `fit$simulate` and see that it runs much of the same code.

```{r}
fit$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
