---
title: "Binomial change point analysis with mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binomial change point analysis with mcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`mcp` aims to implement Generalized Linear Models in a way that closely mimics that of [brms::brm](https://github.com/paul-buerkner/brms). You can set the family and link functions using the `family` argument.

First, let us specify a toy model with three segments:

```{r}
model = list(
  y | trials(N) ~ 1,  # constant rate
  ~ 0 + year,  # joined changing rate
  ~ 1 + year  # disjoined changing rate
)
```


# Simulate data
If you already have data, you can safely skip this section. 

We run `mcp` with `sample = FALSE` to get what we need to simulate data.

```{r}
library(mcp)
empty = mcp(model, family = binomial(), sample = FALSE)
```

Now we can simulate. First, let us see the model parameters.

```{r}
empty$pars
```

 * It takes two intercepts (`int_*`), for segments 1 and 3.
 * It takes two slopes (`year_*`), for segment 2 and 3.
 * It takes two change points (`cp_*`) - one between each segment. 

`empty$simulate` is now a function that can predict data given these parameters. If you are in a reasonable R editor, type `empty$simulate(` and press TAB to see the required arguments. I came up with some values below, including change points at $year = 25$ and $year = 65$. Notice that because `binomial()` defaults to the link function `link = "logit"`, the intercept and slopes are on a [logit scale](https://en.wikipedia.org/wiki/Logit). Briefly, this extends the narrow range of binomial rates (0-1) to an infinite logit scale from minus infinity to plus infinity. This will be important later when we set priors.

```{r}
df = data.frame(
  year = 1901:2000,  # evaluate for each of these
  N = sample(10:20, size = 100, replace = TRUE)  # number of trials
)
df$y = empty$simulate(
  df$year, df$N, 
  cp_1 = 1925, cp_2 = 1965, 
  int_1 = 2, int_3 = 0, 
  year_2 = -0.2, year_3 = 0.05)

head(df)
```

Visually:

```{r}
plot(df$year, df$y)
```


# Check parameter recovery
The next sections go into more detail, but let us quickly see if we can recover the parameters used to simulate the data.

```{r, cache = TRUE, results=FALSE, message=FALSE, warning=FALSE}
fit = mcp(model, data = df, family = binomial(), adapt = 5000)
```


We can use `summary` to see that it recovered the parameters to a pretty good precision. Again, recall that intercepts and slopes are on a `logit` scale.

```{r}
summary(fit)
```

`summary` uses 95% highest density intervals (HDI) by default, but you can change it using `summary(fit, width = 0.80)`. If you have [varying effects](../articles/varying.html), use `ranef(fit)` to see them.

Plotting the fit confirms good fit to the data, and we see the discontinuities at the two change points:

```{r}
plot(fit)
```

These lines are just `fit$simulate` applied to a random draw of the posterior samples. In other words, they represent the joint distribution of the parameters. You can change the number of draws (lines) using `plot(fit, lines = 50)`.

Notice for binomial models it defaults to plot the *rate* (`y / N`) as a function of x. The reason why is obvious when we plot on "raw" data by toggling `rate`:

```{r}
plot(fit, rate = FALSE)
```

These lines are jagged because `N` varies from year to year. Although there is close too 100% success rate in the years 1900 - 1920, the number of trials varies, as you can see in the raw data. However, using `rate = FALSE` will be great when the number of trials is constant for extended periods of time, as `y` is more interpretable then.

Of course, these plots work with [varying effects](../articles/varying.html) as well.


# Model diagnostics and sampling options
Already in the default `plot` as used above, it will be obvious if there was poor convergence. A more direct assessment is to look at the posterior distributions and trace plots:

```{r, fig.height = 10, fig.width = 6}
plot_pars(fit)
```

Convergence is perfect here as evidenced by the overlapping trace plots that look like fat caterpillars (Bayesians love fat caterpillars). Notice that the posterior distribution of change points can be quite non-normal and sometimes even bimodal. Therefore, one should be careful not to interpret the HDI as if it was normal.

`plot()` and `plot_pars()` can do a lot more than this, so check out their documentation.


# Priors for binomial models
`mcp` uses priors to achieve a lot of it's functionality. See [how to set priors](../articles/priors.html), including how to share parameters between segments and how to fix values. Here, I post a few notes about the binomial-specific default priors.

The default priors in `mcp` are set so that they are reasonably broad to cover most scenarios, though also specific enough to sample effectively. They are not "default" as in "canonical". Rather, they are "default" as in "what happens if you do nothing else". All priors are stored in `fit$prior` (also `empty$prior`). We did not specify `prior` above, so it ran with default priors:

```{r}
cbind(fit$prior)
```

The priors on change points are discussed extensively in the prior vignette. The priors on slopes and intercepts are normals with standard deviation of "3" logits. This corresponds to quite extreme binomial probabilities, yet not so extreme as to be totally flat. Here are visualization of priors `dnorm(0, 1)` (red), `dnorm(0, 2)` (black, mcp default), and a `dnorm(0, 5)` (blue) prior, and the correspondence between logits and probabilities:

```{r}
inverse_logit = function(x) exp(x) / (1 + exp(x))

# Start the plot
library(ggplot2)
ggplot(data.frame(logits = 0), aes(x = logits)) + 
  
  # Plot normal prior. Set parameters in "args"
  stat_function(fun=dnorm, args = list(mean=0, sd = 1), lwd=2, col="red") +
  stat_function(fun=dnorm, args = list(mean=0, sd = 3), lwd=2, col="black") +
  stat_function(fun=dnorm, args = list(mean=0, sd = 5), lwd=2, col="blue") +
  
  # Set the secondary axis
  scale_x_continuous(breaks = -7:7,limits = c(-7, 7), sec.axis = sec_axis(~ inverse_logit(.), name = "Probability", breaks = round(inverse_logit(seq(-7, 7, by = 2)), 3)))
```

Please keep in mind that when these priors combine through the model, the joint probability may be quite different. 

Returning to the priors, the `3 / (MAXX - MINX)` on slopes mean that this change in probability occurs over the course of the observed X.


# JAGS code
Here is the JAGS code for the model used in this article.

```{r}
cat(fit$jags_code)
```

