---
title: "Binomial change point analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binomial change point analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`mcp` aims to implement Generalized Linear Models in a way that closely mimics that of [brms::brm](https://github.com/paul-buerkner/brms). You can set the family and link functions using the `family` argument.

First, let us specify a toy model with three segments:

```{r}
model = list(
  y | trials(N) ~ 1,  # constant success probability
  ~ 0 + year,  # joined changing success probability
  ~ 1 + year  # disjoined changing success probability
)
```


# Simulate data
If you already have data, you can safely skip this section. 

We run `mcp` with `sample = FALSE` to get what we need to simulate data.

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic

df = data.frame(
  year = 1901:2020,  # evaluate for each of these
  N = sample(10:20, size = 120, replace = TRUE),  # number of trials
  y = 1
)

empty = mcp(model, data = df, family = binomial(), sample = FALSE)
```

Now we can simulate. First, let us see the model parameters.

```{r}
mcp_pars(empty)
```

 * It takes two intercepts (`Intercept_*`), for segments 1 and 3.
 * It takes two slopes (`year_*`), for segment 2 and 3.
 * It takes two change points (`cp_*`) - one between each segment. 

`empty$simulate` is now a function that can predict data given these parameters (signature: `empty$simulate(fit, newdata, ...)`). If you are in a reasonable R editor, type `empty$simulate(` and press TAB to see the required arguments. I came up with some values below, including change points at $year = 25$ and $year = 65$. Because `binomial()` defaults to `link = "logit"`, the intercept and slopes are on the [logit scale](https://en.wikipedia.org/wiki/Logit), which maps success probabilities between 0 and 1 to the real line. This will be important later when we set priors.

```{r}
df$y = empty$simulate(
  empty, df, 
  cp_1 = 1925, cp_2 = 1975, 
  Intercept_1 = 2, Intercept_3 = -1, 
  year_2 = -0.1, year_3 = 0.1)

head(df)
```

Visually:

```{r}
plot(df$year, df$y)
```


# Check parameter recovery
The next sections go into more detail, but let us quickly see if we can recover the parameters used to simulate the data.

```{r, cache = TRUE, results=FALSE, message=FALSE, warning=FALSE}
fit = mcp(model, data = df, family = binomial(), adapt = 5000)
```


We can use `summary` to see that it recovered the parameters to a pretty good precision. Again, recall that intercepts and slopes are on a `logit` scale.

```{r}
summary(fit)
```

`summary` uses 95% central posterior intervals by default, but you can change it using `summary(fit, width = 0.80)`. If you have [group-level effects](../articles/varying.html), use `ranef(fit)` to see their deviations.

Plotting the fit confirms good fit to the data, and we see the discontinuities at the two change points:

```{r}
plot(fit)
```

These lines are just `fit$simulate` applied to a random draw of the posterior samples. In other words, they represent the joint distribution of the parameters. You can change the number of draws (lines) using `plot(fit, lines = 50)`.

For binomial models, the default plot shows the observed success proportion (`y / N`) and fitted success probability as functions of x. Compare this with success counts by toggling `rate`:

```{r}
plot(fit, rate = FALSE)
```

These lines are jagged because `N` varies from year to year. Although the success probability is close to 1 in 1900--1920, the number of trials varies. Using `rate = FALSE` can be useful when the number of trials is constant for extended periods and the success count `y` is directly interpretable.

Speaking of alternative visualizations, you can also plot this on the logit scale, where the linear trends are modeled:

```{r, warning=FALSE, message=FALSE}
plot_dpar(fit)
```


These plots work with [group-level effects](../articles/varying.html) as well.


# Model diagnostics and sampling options
Already in the default `plot` as used above, it will be obvious if there was poor convergence. A more direct assessment is to look at the posterior distributions and trace plots:

```{r, fig.height = 10, fig.width = 6}
plot_pars(fit)
```

Convergence is perfect here as evidenced by the overlapping trace plots that look like fat caterpillars (Bayesians love fat caterpillars). Notice that the posterior distribution of change points can be quite non-normal and sometimes even bimodal. Therefore, one should be careful not to interpret the interval as if it was normal.

`plot()` and `plot_pars()` can do a lot more than this, so check out their documentation.


# Priors for binomial models
`mcp` uses priors to achieve a lot of it's functionality. See [how to set priors](../articles/priors.html), including how to share parameters between segments and how to fix values. Here, I post a few notes about the binomial-specific default priors.

The default priors in `mcp` are set so that they are reasonably broad to cover most scenarios, though also specific enough to sample effectively. They are not "default" as in "canonical". Rather, they are "default" as in "what happens if you do nothing else". All priors are stored in `fit$prior` (also `empty$prior`). We did not specify `prior` above, so it ran with default priors:

```{r}
cbind(fit$prior)
```

The priors on change points are discussed extensively in the prior vignette. With the default logit link, intercepts and categorical contrasts use `dt(0, 1.5, 3)`. Its central 95% interval is approximately -4.8 to 4.8 logits, corresponding to probabilities from 0.008 to 0.992, while its heavy tails retain support for more extreme values. The figure compares Student-t priors with scales 1 (red), 1.5 (black, the `mcp` default), and 2.5 (blue), together with their correspondence to probabilities. The probit link currently uses the same numerical default for simplicity.

Compared with the `dt(0, 2.5, 3)` intercept and flat coefficient defaults in
`brms`, these proper priors mildly reduce extreme predictions in short
segments while remaining broad and heavy-tailed.

```{r}
inverse_logit = function(x) exp(x) / (1 + exp(x))
scaled_t = function(x, scale) dt(x / scale, df = 3) / scale

# Start the plot
library(ggplot2)
ggplot(data.frame(logits = 0), aes(x = logits)) + 
  
  # Plot Student-t priors. Set scale in "args"
  stat_function(fun=scaled_t, args = list(scale = 1), lwd=2, col="red") +
  stat_function(fun=scaled_t, args = list(scale = 1.5), lwd=2, col="black") +
  stat_function(fun=scaled_t, args = list(scale = 2.5), lwd=2, col="blue") +
  
  # Set the secondary axis
  scale_x_continuous(breaks = -7:7,limits = c(-7, 7), sec.axis = sec_axis(~ inverse_logit(.), name = "Probability", breaks = round(inverse_logit(seq(-7, 7, by = 2)), 3)))
```

Please keep in mind that when these priors combine through the model, the joint probability may be quite different. 

Numeric coefficient scales are divided by a representative change in their model-matrix column: its range when it has two values, and two standard deviations otherwise. Terms involving local `par_x` additionally use the expected segment width, `(max(x) - min(x)) / n_segments()`. The implied change across a typical segment therefore has the same 1.5-logit scale, even when the model contains several change points.


# JAGS code
Here is the JAGS code for the model used in this article.

```{r}
fit$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
