---
title: "Change point model comparison with mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Change point model comparison with mcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

There are three main ways of doing model comparison in `mcp`:

 * Compare any N `mcp` models using leave-One-Out cross validation (LOO-CV). Check out `loo(fit)`, `loo::loo_compare(fit1$loo, fit2$loo, ...)`, and `loo::loo_model_weights(fit1$loo, fit2$loo, ...)`.
 * Flexible directional tests using `hypothesis(fit, cp_1 > 40)` or `hypothesis(fit, cp_1 > 40 & x_2 > x_1)`.
 * Point Bayes Factors tests using the Savage-Dickey density ratio, e.g., `hypothesis(fit, cp_1 = 50)` or `hypothesis(fit, x_2 = x_1)`.


# Some models to play with
We know quite a bit about human short term memory. Namely, the average human has almost perfect recall when presented with 1-4 items (we remember them all), and then errors begins intruding when presented with more items [(Cowan, 2001)](https://www.ncbi.nlm.nih.gov/pubmed/11515286). We also know that memory is not infinite, so as $N \rightarrow \infty$, it has to plateau.

In other words, on "easy" trials with $N <= capacity$ items to be recalled, we expect a constant high binomial rate. When $N > capacity$, the rate declines. We specify a prior reflecting our a priori knowledge.

```{r}
model = list(
  recalled | trials(items) ~ 1,
  ~ 0 + items
)
prior = list(
  int_1 = "dnorm(2, 1)",  # high recall for easy trials
  cp_1 = "dnorm(4, 1)",  # performance dicontinuity around 4
  items_2 = "dnorm(-0.4, 1) T( , -0.2)"  # performance deteriorates considerably
)

# A very simple model
model_simple = list(recalled | trials(items) ~ 1 + items)
```

Notice that `items` is used both as `trials` and x-axis. No problem. 


Simulate some data using `fit$simulate()`:

```{r}
library(mcp)
options(mc.cores = 3)  # Speed up sampling!

empty = mcp(model, family = binomial(), sample = FALSE)

set.seed(42)
df = data.frame(items = rep(1:9, each = 40))
df$recalled = empty$simulate(
  df$items, 
  cp_1 = 4,
  int_1 = 2.5, 
  items_2 = -0.4, 
  sigma = 0.5)
```


Now fit it with and without the informative prior. We set `iter` fairly high because this model is not sampled effectively. We will also add `sample = "both"` to sample *both* prior and posterior. We will need both to compute point Bayes Factors using `hypothesis()` later.

```{r, cache = TRUE, message = FALSE, warning = FALSE, results=FALSE}
fit_default = mcp(model, data = df,             family = binomial(), sample = "both", iter = 10000)
fit_info = mcp(model, data = df, prior = prior, family = binomial(), sample = "both", iter = 10000)
fit_simple = mcp(model_simple, data = df,       family = binomial(), sample = "both", iter = 10000)
```

We plot them and add a few `ggplot2` layers. We jitter the raw data in the middle plot, just to give a sense of the densities.

```{r}
library(patchwork)
library(ggplot2)

plot(fit_default) + 
  ggtitle("Default") +
  
plot(fit_info) + 
  ggtitle("Informed") + 
  geom_jitter(height = 0.05, color = "red", size = 0.2) + 
  
plot(fit_simple) + 
  ggtitle("Simple")
```


# Bayes Factors using Savage-Dickey density ratios
You can compute probabilities and Bayes Factors for various hypotheses using `hypothesis()`. For example, let's test the idea that the change point in recall occurs at four items:

```{r}
hypothesis(fit_default, "cp_1 = 4")
```

Let's unpack:

 * `hypothesis`: For internal convenience, `hypothesis` always re-arranges to test against zero. 
 * `mean`: When subtracting $4$, the posterior distribution is very close, but somewhat dispersed. 
 * `lower` and `uppwer`: The interval width defaults to 95% highest density interval, but you can change it using `hypothesis(..., width = 0.8)`. It is two-sided, so if you want the bound for a one-sided 5% test, just do `width = 0.975`.
 * `BF`: The Savage-Dickey density ratio, which is a Bayes Factor. This is the factor by which the density increases from the prior to the posterior at `cp_1 = 4`. A $BF > 1$ means that we now believe *more* in this value, $BF = 1$ means that our credence to this value is *unchanged*, and $BF < 1$ means that we believe *less* in it.
 * `p`: Because `BF` is an odds ratio, `p` can just be computed from it and contains no new information. For example, $p = 0.95$ corresponds to $BF = 19$. 

Because `BF` and `p` depends directly on the prior in Savage-Dickey test, it is as much an expression of the prior as the posterior. Computing it for the higher-prior-density-at-cp_1=4 `fit_info`, we see a substantially lower BF because we already had a substantial prior belief in `cp_1 = 4`:
 
```{r}
hypothesis(fit_info, "cp_1 = 4")
```

[The Dirichlet prior on change points](../docs/articles/priors.html) may be better than the default prior when testing point hypotheses on change points, but an informed prior beats both.



# Directional and combinatoric tests
Maybe we just want to test a few directional hypothesis. For example:

 * Does recall begin to deteriorate when `cp_1 > 3`?
 * Is the change point in the interval `cp_1 > 3.5 & cp_1 < 4.5`?
 * The only constraint is your imagination. How about all hypotheses we may have at once? `cp_1 > 3.5 & cp_1 < 4.5 & items_2 < -0.4 & int_1 > 2.5` against it's inverse `(cp_1 < 3.5 | cp_1 > 4.5) & items_2 > -0.4 & int_1 < 2.5`

```{r}
hypothesis(fit_info, c(
  "cp_1 > 3",
  "cp_1 > 3.5 & cp_1 < 4.5",
  "cp_1 > 3.5 & cp_1 < 4.5 & items_2 < -0.4 & int_1 > 2.5",
  "(cp_1 < 3.5 | cp_1 > 4.5) & items_2 > -0.4 & int_1 < 2.5"
))
```

Because Bayes Factors are transitive, we can divide them to get the relative evidence for two hypotheses. Here, we get approximately $0.40 / 0.04 = 10$ in favor of hypothesis 3 over hypothesis 4.

There are more examples [in the documentation for `hypothesis`](../docs/reference/hypothesis.html), including how to test varying effects. 

`mcp` does nothing fancy for the directional tests, and you could do them yourself. Just takes the hypothesis verbatim into `mutate` and `summarise` the proportion of samples that "passes" this test. These are identical:

```{r, eval = FALSE}
hypothesis(fit, "cp_1 > 3.5 & cp_1 < 4.5")

# ... is identical to
library(dplyr)
result = tidybayes::tidy_draws(fit$mcmc_post) %>%
  mutate(test = cp_1 > 3.5 & cp_1 < 4.5) %>%
  summarise(p = sum(test) / n())

print(c(p = result$p, 
        BF = result$p / (1 - result$p)))
```



# Cross Validation
We can use the cross-validation from the `loo` package to compare the predictive performance of `mcp` models. Use loo to compute Widely Applicable Information Criterion (WAIC) or Estimated Log Predictive Density (ELPD) for each model, and then compare them using `loo::loo_compare()`. 

The strength of LOO-CV is that you can compare any N models, as long as they are models of same data. In general, LOO-CV is the only inferential solution in `mcp` the models are not nested in each other. For example:

 * The existence of one or several change points. The [article on Poisson change points](../docs/articles/poisson.html) contain an example of comparing a change-point model to a model without change points.
 * Models that differ by several parameters.
 * Comparing different priors.


## What is LOO-CV?
You can read more about Leave-One-Out Cross Validation elsewhere, but briefly, it does this:

 1. Computes the posteriors using all data *less one data point* (hence "leave one out"). 
 2. Computes the posterior density at the left-out data point (out-of-sample data), i.e., what probability the model assigned to it occurring. For example, if the posterior is a normal distribution, a data point near the mean of the posterior has higher density (it is less "surprising") than if it is at $z = -3$. Better predictions means higher densities, i.e., less surprisal.
 3. Repeats step 2 for all observed data and multiplies these densities to get the combined predictive densities at unobserved data. Multiplying is the same as summing in log-space, and the latter has the advantage of being computationally much more feasible. I hope that the name "Estimated Log Predictive Density" (ELPD) makes sense now. The higher the ELPD, the better.

As with Bayes Factors, you can obtain positive evidence for null models. The reason simpler models can be preferred in Bayes is that each new parameter increases the prior predictive space, and thereby comes with a greater "risk" of making way-off predictions. If the parameter does too little to "make up" for this by increasing the likelihood (making a difference), it is a net negative for predictive accuracy, and the simpler model will be preferred. So *narrower priors are simpler models*.

Also, if the data is reasonably informative (it almost always is), LOO-CV is much less influenced by priors than Bayes factors. For LOO, the priors can more be thought of as a regularization (a way to ensure that sampling is efficient) than as the point of departure which all our inferences are relative to (Bayes Factors).


## Applied example
We can look at the loo for one model:
```{r}
loo(fit_info)
```

This is not terribly informative in and of itself. $looic = -2 * elpd_loo$ as is the corresponding SEs, so that is just a matter of scale. What ELPD tells you is that the product of the densities of all left-out data points is approximately $exp(-335.2) ~ 10 ^ {-146}$, a vanishingly small number because we multiply many small numbers. This is hard to interpret because the density at a given point is a strange thing. Furthermore, it depends on the size of the dataset (the more small densities you multiply, the smaller the ELPD).

What is interesting is the *relative* differences in these predicted densities. We can compare the models using `loo::loo_compare()`. We save the results in `fit$loo` to keep things together, which will also be useful if you want to `save(fit)` the objects for later use.

```{r, cache = TRUE}
fit_default$loo = loo(fit_default)
fit_info$loo = loo(fit_info)
fit_simple$loo = loo(fit_simple)

loo::loo_compare(fit_default$loo, fit_info$loo, fit_simple$loo)
```

Aha, so the second model ("model2", i.e., `fit_info`) passed to `loo_compare()` was preferred as the other models had smaller ELPDs. As a first rough conclusion, we've learned that informed priors and a plateau improved out-of-sample predictions.

Again, the absolute number (`elpd_diff`) is hard to relate to. More interpretable is the `elpd_diff`/`se_diff` ratio. This is almost like a z-score, i.e., a ratio of 1.96 corresponds to 95% probability that one model has superior predictive accuracy. This is not an *effect size* but more a measure of *certainty* that this is not compatible with equal predictive properties. Just like Bayes Factors, a nice property over frequentist p-values is that you can quantify the relative evidence for and against any pair of models without having to assume one of them as null.

Note that the `loo` developers have advised that this ratio is too optimistic in practice, but that one can be confident if it exceeds 5 [Citation needed]. So to conclude, none of these models above stand out as particularly good or bad.

You could do the same using `waic`. This is computationally lighter at the cost of robustness to influential data points. The results are almost always practically identical.

```{r, cache = TRUE}
fit_default$waic = waic(fit_default)
fit_info$waic = waic(fit_info)
fit_simple$waic = waic(fit_simple)

loo::loo_compare(fit_default$waic, fit_info$waic, fit_simple$waic)
```

## Stacking
The `loo` package contains many other useful functions. For example, if prediction is your goal, it is often optimal to *combine* models rather than letting the "winner-takes-it-all" do all the predicting. This is also called *stacking*:

```{r, eval = FALSE}
loo::stacking_weights(fit_default$loo, fit_info$loo, fit_simple$loo)
```

What this means is that once model2 (`fit_info`) has done it's "predicting", the others add very little over and above that. If you want to learn about how well they predict relative to each other, use `loo::peudobma_weights()` which weights proportional to the ELPD of each model.
