---
title: "Variance change points and other distributional parameters"
date: "`r format(file.mtime(knitr::current_input()), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance change points and other distributional parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: none
---

`mcp` supports **distributional regression**, where any parameter of a response distribution can be regressed on predictors and have change points across segments. In standard regression, formulas model only the conditional mean `mu()`. Distributional regression extends this by allowing formulas for distributional parameters such as residual standard deviation (`sigma`), negative-binomial overdispersion (`shape`) - and auxilliary parameters for time-series autocorrelation (`ar` / `ma`).

Here, we exemplify regression on the standard deviation `sigma`. A plateau mean with an increasing standard deviation can be written as `y ~ 1 + sigma(1 + x)`. Read more about mcp formulas [here](../articles/formulas.html).

# Priors for distributional parameters

Aligning with `brms` convention, without an explicit `sigma()` formula, the default is the intercept-only prior `dt(0, max(2.5, round(mad(y), 1)), 3) T(0, )` directly on the residual SD, i.e., an **identity link**. The response-calibrated scale adapts to the units of `y`, with a minimum of 2.5 to avoid an accidentally narrow prior. This matches `brms`.

With an explicit `sigma()` formula, the intercept instead defaults to `dt(0, 2.5, 3)` on **log-SD**, with scaled Student-t coefficient priors.

# Simple example: a change in standard deviation

Let us model a simple change in residual standard deviation with a plateau mean:

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic

model = list(
  y ~ 1,  # sigma(1) is implicit in the first segment
  ~ 0 + sigma(1)  # a new intercept on sigma, but not on the mean
)
```

We can simulate data with an abrupt change from a low to a high residual standard deviation at $x = 50$:

```{r}
df = data.frame(x = 1:100, y = 1)
empty = mcp(model, data = df, sample = FALSE, par_x = "x")

set.seed(42)
df$y = empty$simulate(
  empty, df, 
  cp_1 = 50, Intercept_1 = 20, 
  sigma_1 = log(5), sigma_2 = log(20))

head(df)
```

Now we fit the model to the simulated data.

```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, par_x = "x", seed = 42)
```

We plot a posterior predictive interval to show the effect of the changing standard deviation, which is not immediately visible in the default plot of the expected response:

```{r}
set.seed(42)
plot(fit, q_predict = TRUE, ndraws = Inf)  # Inf to get high precision
```

We can see all parameters are well recovered (compare `sim` to `mean`). Like the other parameters, the `sigma`s are named after the segment where they were instantiated. There will always be a `sigma_1`.

```{r, R.options = list(width = 9999)}
summary(fit)
```

# Advanced example

We can model changes in `sigma` alone or together with changes in the mean. The following deliberately complex model demonstrates that flexibility:

```{r}
model = list(
  # Increasing standard deviation.
  y ~ 1 + sigma(1 + x),
  
  # Abrupt change in mean and standard deviation.
  ~ 1 + sigma(1),
  
  # Joined slope on mean; log-SD changes as a second-order polynomial.
  ~ 0 + x + sigma(0 + x + I(x^2)),
  
  # Continue slope on mean, but plateau standard deviation (no sigma() term).
  ~ 0 + x
)

# The slope in segment 4 is just a continuation of 
# the slope in segment 3, as if there was no change point.
prior = list(
  x_4 = "x_3"
)
```

Notice a few things here:

- The prior ensures that the mean slope remains constant between segment 3 and 4. So only `sigma` changes there. Sharing the slope through the prior effectively removes the change point in the mean (read more about [priors in mcp](../articles/priors.html)).
- Segment 4: Without a new `sigma()` term, segment 4 and later segments inherit the preceding standard-deviation trajectory at the value where it was left.

In general, the log-SD parameters are named `sigma_[normalname]`, where "normalname" is the usual parameter names in mcp (see more [here](../articles/formulas.html)). For example, the log-SD slope on `x` in segment 3 is `sigma_x_3`. However, `sigma_Intercept_i` is just too verbose, so log-SD intercepts are simply called `sigma_i`, where i is the segment number.

## Simulate data

We simulate some data from this model, setting all parameters. As always, we can fit an empty model to get `fit$simulate`, which is useful for simulation and predictions from this model.

```{r}
# Set up model
df = data.frame(x = 1:200, y = 1)
empty = mcp(model, data = df, sample = FALSE)

# Simulate data
set.seed(42)
df$y = empty$simulate(
  empty, df,
  cp_1 = 50, cp_2 = 100, cp_3 = 150,
  Intercept_1 = -20, Intercept_2 = 0,
  sigma_1 = log(3), sigma_x_1 = log(2) / 50,
  sigma_2 = log(10),
  sigma_x_3 = -0.03,
  sigma_xE2_3 = 0.0003,
  x_3 = 1, x_4 = 1)
```

## Fit it and inspect results

Fit it:

```{r, cache = TRUE, results = FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, prior = prior, warmup = 6000, iter = 10000, seed = 42)
```

Plotting a posterior predictive interval is an intuitive way to see how the residual standard deviation is estimated:

```{r}
set.seed(42)
plot(fit, q_predict = TRUE)
```

We can also plot the `sigma_` parameters directly. Now the y-axis is `sigma`:

```{r}
set.seed(42)
plot_dpar(fit, dpar = "sigma", q_fit = TRUE)
```

`summary(fit)` shows that the parameters are well recovered and a posterior predictive check (`pp_check(fit)`) looks good too. The last change point is estimated with greater uncertainty than the others because its only signal is a stop in standard-deviation growth.

```{r, R.options = list(width = 9999)}
summary(fit)
pp_check(fit)
```

The bulk and tail effective sample sizes (`ess_bulk` and `ess_tail`) are fairly low, indicating poor mixing for these parameters. `rhat` is acceptable at \< 1.1, indicating good convergence between chains. Let us verify this by taking a look at the posteriors and trace. For now, we just look at the sigmas:

```{r, fig.height=7, fig.width = 6}
set.seed(42)
plot_pars(fit, regex_pars = "sigma_")
```

This confirms the impression from `rhat`, `ess_bulk`, and `ess_tail`. Setting `mcp(..., iter = 10000)` would be advisable to increase the effective sample size. Read more about [tips, tricks, and debugging](../articles/tips.html).

# JAGS code for distributional parameters

Printing the underlying JAGS code, you can see that the regression model is the same for the mean `mu` (see under "\# Formula for mu") and the standard deviation `sigma` (see under "\# Formula for sigma").

This is explained more in [Understanding mcp formulas](../articles/formulas.html).

```{r}
fit$jags_code
```

# Group-level change points and standard deviation

The standard-deviation model also applies with group-level change-point deviations. Here we model a by-person change in `sigma` on a flat mean.

```{r}
model = list(
  # intercept mean. sigma(1) is implicit in first segment if not specified
  y ~ 1,
  
  # Changed standard deviation, with a group-level change point.
  1 + (1|id) ~ 0 + sigma(1)
)
```

Simulate data:

```{r}
# Set up model
df = data.frame(
  x = 1:270,
  id = rep(1:9, times = 30),
  y = 1
)
empty = mcp(model, data = df, par_x = "x", sample = FALSE)

# Simulate data
set.seed(42)
df$y = empty$simulate(
  empty, df, 
  cp_1 = 130, cp_1_sd = 40,
  Intercept_1 = 20,
  sigma_1 = log(10), sigma_2 = log(25))
```

Fit it:

```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, par_x = "x", seed = 42)
```

Plot it:

```{r}
set.seed(42)
plot(fit, q_predict = TRUE, facet_by = "id", ndraws = Inf)
```

As usual, we can get the individual change points:

```{r, R.options = list(width = 9999)}
ranef(fit)
```

... and verify via posterior predictive checks:

```{r}
pp_check(fit, facet_by = "id")
```

# Generalizing to other distributional parameters (dpars)

`sigma` is just one instance of a distributional parameter in `mcp`. In general, every response family defines a set of supported `dpars` (distributional parameters). The default linear predictor in any segment formula models the location parameter `mu` (or probability/rate under the family's link function). Other `dpars` can be regressed on predictors or given change points using wrapper functions named after the `dpar`.

Here is an overview of standard distributional parameters across response families:

- **`mu`**: Conditional mean or location (e.g. mean for `gaussian()`, success probability for `binomial()`, event rate for `poisson()`). Modeled by default without any wrapper.
- **`sigma`**: Residual standard deviation for `gaussian()` (modeled on the log-SD scale via `sigma(...)`).
- **`shape`**: Overdispersion / shape parameter for `negbinomial()` (modeled on the log-shape scale via `shape(...)`).
- **`ar()` / `ma()`** Autoregressive and moving-average serial dependence terms for time-series models ([read more here](../articles/arma.html)). They are not really distributional parameters, but they are similar in terms of the `mcp` model syntax.

### Inspecting and working with dpars

You can inspect the parameters in a fitted model or family object using `mcp_pars(fit)` or `print(fit$prior)`.

Many core `mcp` functions take a `dpar` argument to extract or visualize specific distributional parameters:

- **`fitted(fit, dpar = "sigma")` or `fitted(fit, dpar = "shape")`**: Returns fitted values for the target `dpar` evaluated on the natural/response scale.
- **`predict(fit, dpar = "sigma")`**: Generates posterior predictive draws for the specified distributional parameter.
- **`plot_dpar(fit, dpar = "sigma")` or `plot_dpar(fit, dpar = "shape")`**: Plots the trajectory of the distributional parameter over $x$, complete with credible intervals across segments.
