---
title: "Variance change points and other distributional parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance change points and other distributional parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`mcp` supports **distributional regression**, where any parameter of a response distribution can be regressed on predictors and have change points across segments. In standard regression, formulas model only the conditional mean ($\mu$). Distributional regression extends this by allowing formulas for auxiliary parameters such as residual standard deviation (`sigma`), negative-binomial overdispersion (`shape`), or time-series autocorrelation (`ar` / `ma`).

For Gaussian models, the most common distributional parameter is the residual standard deviation (`sigma`). For example, a plateau mean with an increasing standard deviation can be written as `y ~ 1 + sigma(1 + x)`. An explicit `sigma()` formula uses a log link, so its linear predictor is log-SD and $\text{sigma} = \exp(\eta_{\text{sigma}})$, matching `brms`. This guarantees a positive, smoothly varying standard deviation. Formula syntax allowed outside `sigma()` (where it applies to the mean) also works inside it (where it applies to log-SD). For example, `~ 1 + sigma(1 + z + I(x^2))` is valid. Read more about mcp formulas [here](../articles/formulas.html).

There is one intentional distinction for the common model without any `sigma()` term. Its implicit constant `sigma_1` is the residual SD itself, not log-SD, and its default response-scale half-Student-t prior matches `brms`. As soon as any segment contains an explicit `sigma()` term, all `sigma_*` coefficients in that model are on the log-SD scale. Thus simulation values use, for example, `sigma_1 = log(5)`, while `fitted(fit, dpar = "sigma")` still returns an SD of 5 on the response scale.

# Priors for distributional parameters

Without an explicit `sigma()` formula, the default is the intercept-only prior `dt(0, max(2.5, round(mad(y), 1)), 3) T(0, )` directly on the residual SD. The response-calibrated scale adapts to the units of `y`, with a minimum of 2.5 to avoid an accidentally narrow prior. This matches `brms`.

With an explicit `sigma()` formula, the intercept instead defaults to `dt(0, 2.5, 3)` on log-SD, with scaled Student-t coefficient priors. `brms` uses the same log-SD parameterization and intercept prior but flat coefficients; the proper `mcp` defaults support prior sampling and mildly regularize short segments.

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic
```


# Simple example: a change in standard deviation
Let us model a simple change in residual standard deviation with a plateau mean:

```{r}
model = list(
  y ~ 1,  # sigma(1) is implicit in the first segment
  ~ 0 + sigma(1)  # a new intercept on sigma, but not on the mean
)
```

We can simulate data with an abrupt change from a low to a high residual standard deviation at $x = 50$:

```{r}
set.seed(40)
df = data.frame(x = 1:100, y = 1)
empty = mcp(model, data = df, sample = FALSE, par_x = "x")
set.seed(42)
df$y = empty$simulate(
  empty, df, 
  cp_1 = 50, Intercept_1 = 20, 
  sigma_1 = log(5), sigma_2 = log(20))

head(df)
```


Now we fit the model to the simulated data.

```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, par_x = "x", seed = 42)
```
We plot a posterior predictive interval to show the effect of the changing standard deviation, which is not immediately visible in the default plot of the expected response:

```{r}
set.seed(42)
plot(fit, q_predict = TRUE)
```


We can see all parameters are well recovered (compare `sim` to `mean`). Like the other parameters, the `sigma`s are named after the segment where they were instantiated. There will always be a `sigma_1`.

```{r}
summary(fit)
```


# Advanced example
We can model changes in `sigma` alone or together with changes in the mean. The following deliberately complex model demonstrates that flexibility:

```{r}
model = list(
  # Increasing standard deviation.
  y ~ 1 + sigma(1 + x),
  
  # Abrupt change in mean and standard deviation.
  ~ 1 + sigma(1),
  
  # Joined slope on mean; log-SD changes as a second-order polynomial.
  ~ 0 + x + sigma(0 + x + I(x^2)),
  
  # Continue slope on mean, but plateau standard deviation (no sigma() term).
  ~ 0 + x
)

# The slope in segment 4 is just a continuation of 
# the slope in segment 3, as if there was no change point.
prior = list(
  x_4 = "x_3"
)
```

Notice a few things here:

 * Segments 3 and 4: The residual standard deviation changes while the mean slope continues. Sharing the slope through the prior effectively removes the change point in the mean (read more about [priors in mcp](../articles/priors.html)).
 * Segment 4: Without a new `sigma()` term, segment 4 and later segments inherit the preceding standard-deviation trajectory at the value where it was left.

In general, the log-SD parameters are named `sigma_[normalname]`, where "normalname" is the usual parameter names in mcp (see more [here](../articles/formulas.html)). For example, the log-SD slope on `x` in segment 3 is `sigma_x_3`. However, `sigma_int_i` is just too verbose, so log-SD intercepts are simply called `sigma_i`, where i is the segment number.



## Simulate data
We simulate some data from this model, setting all parameters. As always, we can fit an empty model to get `fit$simulate`, which is useful for simulation and predictions from this model.

```{r}
set.seed(40)
df = data.frame(x = 1:200, y = 1)
empty = mcp(model, data = df, sample = FALSE)
set.seed(42)
df$y = empty$simulate(
  empty, df,
  cp_1 = 50, cp_2 = 100, cp_3 = 150,
  Intercept_1 = -20, Intercept_2 = 0,
  sigma_1 = log(3), sigma_x_1 = log(2) / 50,
  sigma_2 = log(10),
  sigma_x_3 = -0.03,
  sigma_xE2_3 = 0.0003,
  x_3 = 1, x_4 = 1)
```


## Fit it and inspect results
Fit it in parallel, to speed things up:

```{r, cache = TRUE, results = FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, prior = prior, seed = 42)
```

Plotting a posterior predictive interval is an intuitive way to see how the residual standard deviation is estimated:

```{r}
set.seed(42)
plot(fit, q_predict = TRUE)
```


We can also plot the `sigma_` parameters directly. Now the y-axis is `sigma`:

```{r}
set.seed(42)
plot_dpar(fit, dpar = "sigma", q_fit = TRUE)
```


`summary()` shows that the parameters are well recovered (compare `sim` to `mean`). The last change point is estimated with greater uncertainty than the others because its only signal is a stop in standard-deviation growth.

```{r}
summary(fit)
```

The bulk and tail effective sample sizes (`ess_bulk` and `ess_tail`) are fairly low, indicating poor mixing for these parameters. `rhat` is acceptable at < 1.1, indicating good convergence between chains. Let us verify this by taking a look at the posteriors and trace. For now, we just look at the sigmas:

```{r, fig.height=7, fig.width = 6}
set.seed(42)
plot_pars(fit, regex_pars = "sigma_")
```

This confirms the impression from `rhat`, `ess_bulk`, and `ess_tail`. Setting `mcp(..., iter = 10000)` would be advisable to increase the effective sample size. Read more about [tips, tricks, and debugging](../articles/tips.html).




# Group-level change points and standard deviation
The standard-deviation model also applies with group-level change-point deviations. Here we adapt the [group-level effects example](../articles/varying.html) and add a by-person change in `sigma`. The second joined slope has a different residual standard deviation, giving the model more information about the change-point location.

```{r}
model = list(
  # intercept + slope
  y ~ 1 + x,
  
  # Joined slope and increased standard deviation, with a group-level change point.
  1 + (1|id) ~ 0 + x + sigma(1)
)
```

Simulate data:
```{r}
set.seed(40)
df = data.frame(
  x = 1:180,
  id = rep(1:6, times = 30),
  y = 1
)
empty = mcp(model, data = df, sample = FALSE)
set.seed(42)
df$y = empty$simulate(
  empty, df, 
  cp_1 = 70, cp_1_sd = 30,
  Intercept_1 = 20, x_1 = 1, x_2 = -0.5,
  sigma_1 = log(10), sigma_2 = log(25))
```

Fit it:
```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, seed = 42)
```

Plot it:
```{r}
set.seed(42)
plot(fit, facet_by = "id")
```


As usual, we can get the individual change points:
```{r}
mcp::ranef(fit)
```


# Generalizing to other distributional parameters (dpars)

`sigma` is just one instance of a distributional parameter in `mcp`. In general, every response family defines a set of supported `dpars` (distributional parameters). The default linear predictor in any segment formula models the location parameter `mu` (or probability/rate under the family's link function). Other `dpars` can be regressed on predictors or given change points using wrapper functions named after the `dpar`.

Here is an overview of standard distributional parameters across response families:

 * **`mu`**: Conditional mean or location (e.g. mean for `gaussian()`, success probability for `binomial()`, event rate for `poisson()`). Modeled by default without any wrapper.
 * **`sigma`**: Residual standard deviation for `gaussian()` (modeled on the log-SD scale via `sigma(...)`).
 * **`shape`**: Overdispersion / shape parameter for `negbinomial()` (modeled on the log-shape scale via `shape(...)`).
 * **`ar()` / `ma()`**: Autoregressive and moving-average serial dependence terms for time-series models ([read more here](../articles/arma.html)).

### Inspecting and working with dpars

You can inspect the parameters in a fitted model or family object using `mcp_pars(fit)` or `cbind(fit$prior)`.

Many core `mcp` functions take a `dpar` argument to extract or visualize specific distributional parameters:

 * **`fitted(fit, dpar = "sigma")` or `fitted(fit, dpar = "shape")`**: Returns fitted values for the target `dpar` evaluated on the natural/response scale.
 * **`predict(fit, dpar = "sigma")`**: Generates posterior predictive draws for the specified distributional parameter.
 * **`plot_dpar(fit, dpar = "sigma")` or `plot_dpar(fit, dpar = "shape")`**: Plots the trajectory of the distributional parameter over $x$, complete with credible intervals across segments.

### Example: Modeling `shape()` in negative-binomial regression

For a `negbinomial()` model, `shape()` controls the overdispersion ($\operatorname{Var}(y) = \mu + \mu^2 / \text{shape}$). You can model a change in overdispersion across segments just like `sigma()`:

```{r, eval=FALSE}
model_nb = list(
  y ~ 1 + x + shape(1),     # Segment 1: intercept on shape
  ~ 1 + x + shape(1 + x)    # Segment 2: changing shape (overdispersion) over x
)
fit_nb = mcp(model_nb, data = df_count, family = negbinomial(), seed = 42)

# Visualize the shape trajectory across segments
set.seed(42)
plot_dpar(fit_nb, dpar = "shape")
```

Just like `sigma()`, an explicit `shape()` formula models log-shape so that the parameter remains strictly positive. All formula features (intercepts, continuous slopes, categorical contrasts, polynomial terms, and group-level deviations) work identically for any `dpar`.


```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
