---
title: "Families and link functions in mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Families and link functions in mcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(mcp)
```


`mcp 0.3` added support for more response families and link functions. For example, you can now do

```{r, eval=FALSE}
fit = mcp(model, data = df, family = gaussian("log"))
fit = mcp(model, data = df, family = binomial("identity"))
fit = mcp(model, data = df, family = exponential("identity"))
```


This is an ongoing effort and more will be added. This table shows the current status:

[[[[[INSERT XLSX HERE]]]]]

Color codes:

 * **Green:** supported and the default priors are unlikely to change.
 * **Yellow:** supported but the default priors may change.
 * **White:** not currently supported. [Raise a GitHub issue](https://github.com/lindeloev/mcp/issues) if you need it.
 * **Red:** will not be supported.

See the "GLM" menu above for more details on using GLM with `mcp`.


# General remarks
Some link functions are *default* in GLM for good reasons: they have proven computationally convenient and interpretable. When using a non-default link function, you risk predicting impossible values, at which point `mcp` will error (as it should) - hopefully with informative error messages. For example a `bernoulli("identity")` family with model `prob ~ 1 + x` (i.e., a slope directly on the probability of success) can easily exceed 1 or 0 and there are, of course, no such thing as probabilities below 0% or above 100%. One way to ameliorate such problems is by setting informative [priors](../articles/priors.html) (e.g., via truncation) to prevent the sampler from visiting illigal combinations of such values.

In short: think carefully and proceed at your own risk.


# An example

Reviving the example from the article on [binomial models in mcp](../articles/binomial.html), we can model it using an `identity` link function:

```{r}
model = list(
  y | trials(N) ~ 1 + x,  # Intercept and slope on P(success)
  ~ 1 + x                 # Disjoined slope on P(success)
)

fit = mcp(model, data = ex_binomial, family = binomial(link = "identity"))
```

Oops, the sampler visited impossible values! Likely `P(success) < 0%` or `P(success) > 100%`. Let's help it along with some more informative priors. For this data and model, the main problem is that the slope of the second segment has too great posterior probability of surpasses 100% with the default `mcp` priors. So let's set some more informative priors render a long (early `cp_1`) and steep (high `x_2`) slope unlikely:

```{r, results=FALSE, message=FALSE, warning=FALSE}
prior = list(
  x_2 = "dnorm(0, 0.002)",         # Slope is unlikely to be steep
  cp_1 = "dnorm(30, 10) T(20, )"   # Slope starts not-too-early
)

fit = mcp(model, data = ex_binomial, family = binomial(link = "identity"), prior = prior)
```

```{r, fig.width=6, fig.height=4}
plot(fit)
```

Sampling succeeded! This is a bad model of this data. But it illustrates the necessary considerations and steps to ameliorate problems when going beyond default link functions. 

Because of the identity-link, the regression coefficients are interpretable as intercepts and slopes on `P(success)` in contrast to the "usual" log-odds fitted when `family = binomial(link = "logit")`. For example, `int_1` is inferred probability of success at `x = 0` and likewise for `int_2` at `x = cp_1`.

```{r}
summary(fit)
```


# Specific remarks

 * The mean ($1 / {inverse\_link}(\lambda)$) is plotted for `family = exponential()` and output by `fit$simulate(..., fype = "fitted")`. Multiply by `log(2)` to get the median.

