---
title: "About mcp formulas and models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About mcp formulas and models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`mcp` takes a list of formulas, and defines the change point as the point on the x-axis where the data shifts from being generated by one formula to the next. So with `N` formulas, you have `N - 1` change points. A list with just one formula thus correspond to normal regression with 0 change points.

The formulas are called "segments" because they divide the (observed) x-axis into `N` segments.

# Formula format for segments
The three formula parts are called **response**, **cp**, and **predictor**. The general format is `response ~ cp ~ predictor`, except for the first segment, which has no change point and therefore uses `response ~ predictor`. Here `cp` names a formula part; it is not a literal variable name. Distributional terms such as `sigma()` live in the predictor part.


 * **The response** is just the name of a column in your data.
 * **The cp part** can be `1` for a population-level change point or `1 + (1|group)` for [group-level change-point deviations](articles/varying.html).
 * **The predictor part** can contain `0` (no intercept), `1` (an intercept), columns from your data whose effects you want to model, and [group-level effects](articles/varying.html), such as `(1|group)` and `(factor||group)`.
 * **Distributional predictors**, such as those inside `sigma()`, have [their own article](../articles/variance.html), but follow the same predictor rules, including group-level effects.

Predictor group-level effects use familiar `lme4`/`brms` syntax. For example, `y ~ 1 + (1|id)` gives each `id` a deviation from the population-level intercept. Use `||` for several independent group coefficients: `(1 + x||id)` adds an intercept and numeric slope, while `(factor||id)` adds an intercept and treatment-coded factor contrasts. `(0 + factor||id)` instead adds one independent coefficient for each factor level. Each coefficient has its own population-level SD.

A group-level term carries into later segments until another term for the same distributional parameter and grouping factor replaces the whole block; use `(0|id)` to turn it off. Correlated multi-coefficient `|` terms, such as `(1 + x|id)`, and group-level terms inside `ar()` or `ma()` are not yet supported.

Transformations in formulas are evaluated on the original predictor values, as
in `lm()` and `glm()`. The change-point predictor is the exception: bare
`par_x` and polynomial bases such as `I(par_x^2)` use distance from the segment
onset so that they can describe joined segment shapes.

For convenience, you can omit the response and cp parts in segment 2+. The previous response and an intercept-only population-level change point are then assumed. When you call `summary(fit)`, it will show the explicit representation. Let us see this in action for this model where we predict `score` as a function of `time` in three segments, i.e., with two change points:

```{r}
library(mcp)
model = list(
  score ~ 1,  # intercept
  score ~ 1 ~ 0 + time,  # joined slope
  ~ time  # disjoined slope. "score ~ 1 ~ 1 + time" is implicit here.
)

# Interpret, but do not sample.
fit = mcp(model, data = data.frame(score = 1:10, time = 1:10), sample = FALSE)
summary(fit)
```

Notice how it added the response and cp parts to the last segment?

`mcp` is heavily inspired by `brms` which again is inspired by [`lme4::lmer`](https://cran.r-project.org/web/packages/lme4/index.html). [Here is a bit of history](https://twitter.com/jonaslindeloev/status/1117760777249853440) on that.


# Parameter names
`mcp` automatically assigns names to the parameters in the format `type_i` where `i` is the segment number. Specifically:

 * `Intercept_i` is the intercept in the ith segment.
 * `year_i` is the slope in on the data column `year` in the ith segment. `x_i` is the slope on the data column `x` in the ith segment. The slope takes name after the data it is regressed on.
 * Multi-column R bases identify both the basis column and segment. For example, `poly_x_2_basis1_1` is the first column of `poly(x, 2)` in segment 1.
 * `cp_i` is the ith change point. Notice that `cp_i` is specified in segment `i + 1`. `cp_1` occurs when there are two segments, and `cp_2` when there are three segments, etc. *OBS: future versions may start at cp_2.*.
 * `cp_i_group` contains the group-level deviations from population-level `cp_i`. See [group-level change points in mcp](articles/varying.html).
 * `cp_i_sd` is the population-level standard deviation governing those group-level deviations.
 * `sigma_*` describe the Gaussian residual standard deviation; [read more here](../articles/variance.html). With an explicit `sigma()` formula these are log-SD coefficients. If no `sigma()` is supplied, the implicit `sigma_1` is instead the common response-scale SD initiated in the first segment.
 * `arj_i` are autocorrelation coefficients of order `j` for segment `i` ([read more here](../articles/arma.html)).


Use `mcp_pars()` to inspect the parameters, including their roles,
distributional parameters, and segments. Use `mcp_columns()` to inspect the
resolved data-column roles, including the change-point predictor. Let us
specify a somewhat complex model:

```{r}
model = list(
  # Intercept_1
  score ~ 1,
  
  # cp_1, cp_1_sd, cp_1_id, year_2
  1 + (1|id) ~ 0 + year,
  
  # cp_2, cp_2_sd, cp_2_id, Intercept_3, year_3
  1 + (1|id) ~ 1 + year
)

# Interpret, but do not sample.
fit = mcp(model, data = data.frame(score = 1:10, year = 1:10, id = 1:10), sample = FALSE)
mcp_pars(fit)
```



# Modeling intercept change points
A change point is simply like an `ifelse` statement or multiplying with indicators (0s and 1s):

```{r}
# Model parameters
x = 1:20
cp_1 = 12
Intercept_1 = 5
Intercept_2 = 10

# Ifelse version
y_ifelse = ifelse(x <= cp_1, yes = Intercept_1, no = Intercept_2)

# Indicator equivalent using dummy helpers
cp_0 = -Inf
cp_2 = Inf
y_indicator = (x > cp_0) * (x <= cp_1) * Intercept_1 +  # Between cp_0 and cp_1
              (x > cp_1) * (x <= cp_2) * Intercept_2  # Between cp_1 and cp_2

# Show it
par(mfrow = c(1,2))
plot(x, y_ifelse, main = "ifelse(x <= cp_1)")
plot(x, y_indicator, main = "(x > cp_1) * Intercept_2")
```

The magic of (Bayesian) MCMC sampling is that it can actually infer the change point from this simple formulation. We let `mcp` write the JAGS code for this simple two-plateaus model and see how it uses the indicator formulation of change points:

```{r}
model = list(y ~ 1, ~ 1)
fit = mcp(model, data = data.frame(y = 1:10, x = 1:10), sample = FALSE, par_x = "x")
fit$jags_code
```

Look at the section called `# Fitted value` which is the (automatically generated) model that was discussed above. Some unnecessary stuff is added to segment 1 just because it makes the code easier to generate. (`x[i_] >= cp_0` when `cp_0` is the smallest value of `x` is, of course, always true).


# Modeling slope change points
We can use the same principle to model change points on slopes. However, we have to "take off" where the previous slope left us on the y-axis. That is, we have to regard whatever y-value the previous segment ended with as a kind of intercept-at-x=0 in the frame of the new segment. The intercept of segment 2 is `cp_1 * slope_1` and the slope in segment 2 is `x * (slope_2 - cp_1)`.

```{r}
# Model parameters
x = 1:20
cp_1 = 12
slope_1 = 2
slope_2 = -1

# Ifelse version
y_ifelse = ifelse(x <= cp_1, 
            yes = slope_1 * x,
            no = cp_1 * slope_1 + slope_2 * (x - cp_1))

# Indicator version. pmin() is a vectorized min()
cp_0 = -Inf
y_indicator = (x > cp_0) * slope_1 * pmin(x, cp_1) + 
              (x > cp_1) * slope_2 * (x - cp_1)

# Show it
par(mfrow = c(1,2))
plot(x, y_ifelse, main = "ifelse(x <= cp_1)")
plot(x, y_indicator, main = "(x > cp_1) * Intercept_2")
```

Let us see this in action:

```{r}
model = list(y ~ 0 + x,
                ~ 0 + x)
fit = mcp(model, data = data.frame(y = 1:10, x = 1:10), sample = FALSE)
fit$jags_code
```

Again, look at the `#Fitted value` to see the indicator-version in action. And again, `mcp` adds something about `cp_0 = -Inf` and `cp_2 = Inf`, just for internal convenience.

You will find the exact same formula for `y_ = ...` if you do `print(fit$simulate)`, though this function contains a whole lot of other stuff too.


# Modeling relative slopes and intercepts
Relative parameters and changepoints can be modeled by specifying prior relationships or parameter offsets. Relative slopes are easy: just replace `x_2` with `x_1 + x_2`. You could do the same if all segments are intercept-only. However, if the previous segment had a slope, we want the intercept to be relative to where that "ended". The `mcp` solution is to only "turn off" the "hanging intercept" from that slope's ending (`pmin(x, cp_i)`) when the model encounters an absolute intercept. An indicator does this.

```{r}
# Model parameters
x = 1:20
cp_1 = 12
Intercept_1 = 5
Intercept_2 = 3  # let's model this as relative

# Indicator version.
cp_0 = -Inf
y_indicator = (x > cp_0) * Intercept_1 +  # Note: no (x < cp_1)
              (x > cp_1) * (Intercept_2)

# Plot it
plot(x, y_indicator, main = "Relative intercept")
```

You can look at `fit$jags_code` and `fit$simulate()` to see this in action.
