---
title: "About mcp formulas and models"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About mcp formulas and models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`mcp` takes a list of formulas, and defines the change point as the point on the x-axis where the data shifts from being generated by one formula to the next. So with `N` formulas, you have `N - 1` change points. A list with just one formula thus correspond to normal regression with 0 change points.

The formulas are called "segments" because they divide the (observed) x-axis into `N` segments.

# Formula format for segments
The format of each segment is generally `response ~ changepoint ~ predictors + sigma(variance_predictors)`, except for the first segment where there is no change point (`response ~ predictors + sigma(variance_predictors)`). Here is a simple and a complex formula


 * **The response** is just the name of a column in your data.
 * **The change point** can be `1` for a population-level change point or `1 + (1|group)` [varying change points](articles/varying.html) sampled around the population-level change point.
 * **The predictors** can be `0` (no change in intercept), `1` (change in intercept), and any column in your data which you want to model a slope on.
 * **The variance predictors** has [it's own article](../articles/variance.html), but they follow the same rules as the **predictors**.

For convenience, you can omit the response and the change point in segment 2+, in which case the former response and an intercept-only (as opposed to random/varying) change point is assumed. When you call `summary(fit)`, it will show the explicit representation. Let us see this in action for this model where we predict `score` as a function of `time` in three segments, i.e., with two change points:

```{r}
library(mcp)
segments = list(
  score ~ 1,  # intercept
  score ~ 1 ~ 0 + time,  # joined slope
  ~ time  # disjoined slope. "score ~ 1 ~ 1 + time" is implicit here.
)

# Interpret, but do not sample.
fit = mcp(segments, sample = FALSE)
summary(fit)
```

Notice how it added the response and the change point to the last segment?

`mcp` is heavily inspired by `brms` which again is inspired by [`lme4::lmer`](https://cran.r-project.org/web/packages/lme4/index.html). [Here is a bit of history](https://twitter.com/jonaslindeloev/status/1117760777249853440) on that.


# Parameter names
`mcp` automatically assigns names to the parameters in the format `type_i` where `i` is the segment number. Specifically:

 * `int_i` is the intercept in the ith segment.
 * `year_i` is the slope in on the data column `year` in the ith segment. `x_i` is the slope on the data column `x` in the ith segment. The slope takes name after the data it is regressed on.
 * `cp_i` is the ith change point. Notice that `cp_i` is specified in segment `i + 1`. `cp_1` occurs when there are two segments, and `cp_2` when there are three segments, etc. *OBS: future versions may start at cp_2.*.
 * `cp_i_group` is the varying *deviations* from `cp_i`. See [varying change points in mcp](articles/varying.html).
 * `cp_i_sd` is the population-level standard deviation of the varying effects.
 * `sigma_*` are variance parameters about which you can [read more here](../articles/variance.html)). Note that for `family = gaussian()` and other families with an SD residual, there will always be an `sigma_1`, i.e., the common sigma initiated in the first segment.
 * `phi_j` are autocorrelation coefficients ([read more here](../articles/autocorrelation.html)). There is one per autoregressive order, i.e., `phi_1` and `phi_2` for `mcp(..., autocor = "AR(2)")`.


These parameter names are saved in `fit$pars`. Let us specify a somewhat complex model to show off some parameter names:

```{r}
segments = list(
  # int_1
  score ~ 1,
  
  # cp_1, cp_1_sd, cp_1_id, x_2
  1 + (1|id) ~ 0 + year,
  
  # cp_2, cp_2_sd, cp_2_condition, int_2, x_2
  1 + (1|condition) ~ 1 + rel(year)
)

# Intepret, but do not sample.
fit = mcp(segments, sample = FALSE)
fit$pars
```



# Modeling intercept change points
A change point is simply like an `ifelse` statement or multiplying with indicators (0s and 1s):

```{r}
# Model parameters
x = 1:20
cp_1 = 12
int_1 = 5
int_2 = 10

# Ifelse version
y_ifelse = ifelse(x <= cp_1, yes = int_1, no = int_2)

# Indicator equivalent using dummy helpers
cp_0 = -Inf
cp_2 = Inf
y_indicator = (x > cp_0) * (x <= cp_1) * int_1 +  # Between cp_0 and cp_1
              (x > cp_1) * (x <= cp_2) * int_2  # Between cp_1 and cp_2

# Show it
par(mfrow = c(1,2))
plot(x, y_ifelse, main = "ifelse(x <= cp_1)")
plot(x, y_indicator, main = "(x > cp_1) * int_2")
```

The magic of (Bayesian) MCMC sampling is that it can actually infer the change point from this simple formulation. We let `mcp` write the JAGS code for this simple two-plateaus model and see how it uses the indicator formulation of change points:

```{r}
segments = list(y ~ 1, ~ 1)
fit = mcp(segments, sample=FALSE, par_x = "x")
cat(fit$jags_code)
```

Look at the section called `# Fitted value` and you should recognize it. Some unnecessary stuff is added to segment 1 just because it makes the code easier to generate. (`x[i_] >= cp_0` when `cp_0 = -10^100` is a way of writing `x >= -Inf` in JAGS and this is, of course, always true).


# Modeling slope change points
We can use the same principle to model change points on slopes. However, we have to "take off" where the previous slope left us on the y-axis. That is, we have to regard whatever y-value the previous segment ended with as a kind of intercept-at-x=0 in the frame of the new segment. The intercept of segment 2 is `cp_1 * slope_1` and the slope in segment 2 is `x * (slope_2 - cp_1)`.

```{r}
# Model parameters
x = 1:20
cp_1 = 12
slope_1 = 2
slope_2 = -1

# Ifelse version
y_ifelse = ifelse(x <= cp_1, 
            yes = slope_1 * x,
            no = cp_1 * slope_1 + slope_2 * (x - cp_1))

# Indicator version. pmin() is a vectorized min()
cp_0 = -Inf
y_indicator = (x > cp_0) * slope_1 * pmin(x, cp_1) + 
              (x > cp_1) * slope_2 * (x - cp_1)

# Show it
par(mfrow = c(1,2))
plot(x, y_ifelse, main = "ifelse(x <= cp_1)")
plot(x, y_indicator, main = "(x > cp_1) * int_2")
```

Let us see this in action:

```{r}
segments = list(y ~ 0 + x,
                ~ 0 + x)
fit = mcp(segments, sample = FALSE)
cat(fit$jags_code)
```

Again, look at the `#Fitted value` to see the indicator-version in action. And again, `mcp` adds something about `cp_0 = -Inf` and `cp_2 = Inf`, just for internal convenience.

This exact code is also used in `fit$simulate`. See under `"Fitted value`:

```{r}
print(fit$simulate)
```


# Modeling relative slopes and intercepts
`mcp` allows for specifying relative intercepts and change points through `rel()`. Relative slopes are easy: just replace `x_2` with `x_1 + x_2`. You could do the same if all segments are intercept-only. However, if the previous segment had a slope, we want the intercept to be relative to where that "ended". The `mcp` solution is to only "turn off" the "hanging intercept" from that slope's ending (`pmin(x, cp_i)`) when the model encounters an absolute intercept. An indicator does this.

```{r}
# Model parameters
x = 1:20
cp_1 = 12
int_1 = 5
int_2 = 3  # let's model this as relative

# Indicator version.
cp_0 = -Inf
y_indicator = (x > cp_0) * int_1 +  # Note: no (x < cp_1)
              (x > cp_1) * (int_2)

# Plot it
plot(x, y_indicator, main = "Relative intercept")
```

You can look at `cat(fit$jags_code)` and `fit$simulate` to see this in action.
