---
title: "Change Point Packages in R: A Practical Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Change Point Packages in R: A Practical Guide}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: none
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

The R ecosystem for change point analysis spans three primary areas of application:

1. **Segmented Regression**: Estimating the parameters of explicitly specified piecewise models (slopes, plateaus, GLM likelihoods, random effects, or prior constraints).
2. **Time Series Decomposition & Structural Breaks**: Isolating unknown trend shifts, periodic seasonality, and structural instability in time series.
3. **Change Point Detection**: Finding unknown shift points algorithmically in large data streams with minimal runtime.

This guide evaluates how packages perform on common benchmark problems to compare workflow, parameter recovery, and uncertainty estimation.


# Overview of Package Capabilities

The table below outlines the primary functional focus of each package across core modeling dimensions:

<style>
.full-width-table {
  width: 100vw;
  position: relative;
  left: 50%;
  right: 50%;
  margin-left: -50vw;
  margin-right: -50vw;
  padding: 0 1.5rem;
  box-sizing: border-box;
  overflow-x: auto;
}
.full-width-table table {
  margin: auto;
  width: 100%;
  max-width: 1150px;
}
</style>

::: full-width-table
| Package | Paradigm | Segmented Regression | GLM Likelihoods | Group / Random Effects | Transition Modeling | Location Uncertainty | Autoregression & Heteroscedasticity |
|:--------|:--------|:-------:|:-------:|:-------:|:-------:|:--------|:-------:|
| **`mcp`** | Bayesian | High | High (Gaussian, Binomial, Poisson, NegBin) | High (on CPs & slopes) | Abrupt (discontinuous or joined) | Posterior densities & credible intervals (ETI / HDI) | High (`sigma()`, `ar()`, `ma()`) |
| **`smoothbp`** | Bayesian | High | - (Gaussian only) | Moderate (on CPs & slopes) | Smooth (logistic transition) | Posterior densities & PIPs | - |
| **`bcp`** | Bayesian | - | - | - | Abrupt (step changes) | Pointwise posterior break probabilities $P(B_t = 1)$ | - |
| **`segmented`** | Frequentist | High | High (`glm()`, `coxph()`) | Moderate (`lme()` integration) | Continuous joined lines only | Frequentist SE & CIs (delta, score, boot) | Basic (`Arima()`) |
| **`Rbeast`** | Bayesian | Moderate (univariate trends) | - | - | Abrupt & gradual | Posterior changepoint probabilities | Seasonal decomposition |
| **`strucchange`** | Frequentist | Moderate | Basic | - | Abrupt | Asymptotic confidence intervals (Bai 1997) | Empirical fluctuation tests |
| **`changepoint`** | Algorithmic | Basic (mean/var) | - | - | Abrupt | Point estimates only | - |
:::


## The Two Axes of Change Point Analysis

The packages in R trade off **model complexity** (functional regression flexibility, generalized likelihoods, random effects, and custom priors) against **computational scalability** (handling high data volume and discovering dozens of unknown change points):

```
                            ▲ High Model Complexity
                            │ (Slopes, GLMs, Random Effects, Priors, ARMA)
             [ mcp ]        │   
           [ smoothbp ]     │      [ segmented ]
                            │   
          [ strucchange ]   │                         [ Rbeast ]
  Low Data Volume ──────────┼──────────────────────────────► High Data Volume
  & Few CPs (1-5)           │             [ bcp ]             & Many CPs (10-50+)
                            │      [ changepoint / PELT ]
                            │
                            ▼ Low Model Complexity
                              (Piecewise constant mean / variance)
```


# Benchmark Regression Data

To compare how different packages handle change points and parameter recovery, we generate a synthetic dataset ($N = 100$) with two true change points at $x = 30$ and $x = 70$:

- **Segment 1 ($x \in [1, 30]$)**: Plateau at $y = 10$.
- **Segment 2 ($x \in [31, 70]$)**: Joined linear slope of $+0.50$ per unit $x$, starting from $y = 10$.
- **Segment 3 ($x \in [71, 100]$)**: Disjoined step jump to intercept $20$ with a downward slope of $-0.30$ per unit $x$.
- **Residual error**: Gaussian noise with $\sigma = 2$.

```{r}
set.seed(42)
df = data.frame(
  x = seq(1, 100),
  y = c(
    rnorm(30, mean = 10, sd = 2),
    10 + 0.5 * (1:40) + rnorm(40, sd = 2),
    20 - 0.3 * (1:30) + rnorm(30, sd = 2)
  )
)
plot(df$x, df$y, xlab = "x", ylab = "y", pch = 16, col = "gray40", main = "Benchmark Regression Data")
abline(v = c(30, 70), col = "firebrick", lty = 2, lwd = 1.5)
```


# Application 1: Segmented & Piecewise Regression

Segmented regression packages estimate interpretable regression parameters (slopes, intercepts, and change point locations) along continuous predictors.


## `mcp`: Generative Regression with Multiple Change Points

`mcp` allows specifying different functional forms for each segment (e.g. plateaus, joined slopes, and disjoined steps).

```{r, message = FALSE, warning = FALSE, results = 'hide'}
library(mcp)

# 3-segment model: plateau -> joined slope -> disjoined step and slope
model = list(
  y ~ 1,        # Segment 1: plateau (Intercept_1)
  ~ 0 + x,      # Segment 2: joined slope (x_2) at cp_1
  ~ 1 + x       # Segment 3: disjoined step & slope at cp_2
)

fit_mcp = mcp(model, data = df, seed = 42, adapt = 500, iter = 1000)
```

```{r}
summary(fit_mcp)
set.seed(42)
plot(fit_mcp, q_fit = TRUE)
```

`mcp` recovers both change points (`cp_1 = 33.0 [29.2, 36.2]`, `cp_2 = 70.5 [70.0, 71.0]`) and accurately estimates the segment parameters (`Intercept_1 = 10.3`, `x_2 = 0.56`, `Intercept_3 = 19.5`, `x_3 = -0.28`, `sigma_1 = 2.12`) within their 95% highest-density intervals.

**Key capabilities:**

- **Arbitrary per-segment formulas**: Segments can independently combine intercepts, continuous slopes, categorical contrasts, polynomial terms, and fixed values.
- **Generalized linear models**: Supports `gaussian()`, `binomial()`, `bernoulli()`, `poisson()`, and `negbinomial()`.
- **Group-level (random) effects**: Hierarchical deviations on change points (`1 + (1|id) ~ ...`) and on segment slopes and intercepts.
- **Variance & Autocorrelation regression**: Models heteroscedasticity (`~ sigma(1 + x)`) and serial dependence (`~ ar(1)`) per segment.
- **Inference & Hypothesis testing**: Full posterior distributions for change points, Savage-Dickey Bayes factors, and `loo` cross-validation.
- **Speed on typical models**: JAGS has zero compilation overhead and fits standard 1–3 change point models in under 1–5 seconds.

**Constraints:**

- Computation uses MCMC sampling (JAGS), requiring more compute time on large datasets (>20,000 observations) or models with $\ge 10$ change points than closed-form alternatives.
- Selecting the number of change points is performed via model comparison rather than within a single model fit.


## `smoothbp`: Smooth Transitions and Breakpoint Selection

`smoothbp` fits Bayesian piecewise linear regression using a compiled Rust sampler, modeling continuous transitions between slopes with variable sharpness. Because default priors assume a standardized predictor scale, setting priors scaled to the predictor range ($x \in [1, 100]$) ensures proper calibration:

```{r, eval = requireNamespace("smoothbp", quietly = TRUE), message = FALSE, warning = FALSE}
library(smoothbp)

priors_sbp = smoothbp_priors(
  b0 = prior_normal(10, 20),
  b1 = prior_normal(0, 5),
  deltas = list(prior_normal(0, 5), prior_normal(0, 5)),
  omega = list(prior_normal(30, 20, lb = 0), prior_normal(70, 20, lb = 0)),
  rho = list(prior_normal(2, 5, lb = 0), prior_normal(2, 5, lb = 0))
)

# Fit smooth change-point model with 2 breakpoints
fit_sbp = smoothbp(
  formula = y ~ x,
  b0 = ~ 1,                        # Baseline intercept
  b1 = ~ 1,                        # Baseline slope
  deltas = list(~ 1, ~ 1),         # Slope changes at each breakpoint
  omega = list(~ 1, ~ 1),          # Breakpoint locations
  rho = list(~ 1, ~ 1),            # Transition sharpness parameters
  data = df,
  priors = priors_sbp,
  iter = 1000,
  warmup = 500,
  chains = 2,
  seed = 42,
  .verbose = FALSE
)

summary(fit_sbp)
```

`smoothbp` estimates `omega1 = 30.09` (true: 30) and `omega2 = 63.86`. Change point posterior distributions can be visualized directly using `plot(..., type = "density")`:

```{r, eval = requireNamespace("smoothbp", quietly = TRUE), message = FALSE, warning = FALSE}
plot(fit_sbp, type = "density", pars = c("omega1_(Intercept)", "omega2_(Intercept)"))
```

Notice the structural difference between `smoothbp`'s $\omega_1$ posterior and `mcp`'s `cp_1`: in `mcp`, `cp_1` is an instantaneous, sharp non-differentiable corner, capturing all observational location uncertainty into a single parameter. In `smoothbp`, $\omega_1$ represents the inflection midpoint of a continuous logistic transition curve whose width is parameterized by sharpness $\rho_1$. Additionally, `plot.smoothbp_fit()` renders separate overlaid density curves per MCMC chain, which reflect chain-specific exploration along the non-linear $\omega \leftrightarrow \rho$ likelihood ridge.

**Key capabilities:**

- **Gradual transitions**: Replaces sharp corners with estimated logistic transition widths ($\rho$).
- **Covariates on structural parameters**: Regresses change-point locations ($\omega$) and transition sharpness ($\rho$) on covariates.
- **Spike-and-slab selection (`smoothbp_ss`)**: Computes posterior inclusion probabilities (PIPs) for candidate breakpoints within a single model fit.
- **Hierarchical breakpoints**: Supports random intercepts on change-point locations (`(1|group)`).

**Constraints:**

- Limited to Gaussian response models. Non-Gaussian GLMs (Poisson, Binomial) and ARMA residual structures are not supported.
- Models continuous slope transitions; discontinuous intercept step changes across segments are not directly parameterized.
- While the sampler is compiled in Rust, evaluating multiple logistic transitions across high-dimensional parameter spaces ($\boldsymbol{\omega} \times \boldsymbol{\rho}$) is computationally demanding and does not outpace `mcp` on typical models.


## `segmented`: Frequentist Piecewise Regression

`segmented` estimates breakpoints by extending standard frequentist regression models (`lm`, `glm`, `coxph`, `Arima`).

```{r, eval = requireNamespace("segmented", quietly = TRUE), message = FALSE, warning = FALSE}
library(segmented)

# Fit base linear model, then estimate 2 change points along x
fit_lm = lm(y ~ x, data = df)
fit_seg = segmented(fit_lm, seg.Z = ~ x, npsi = 2)

# Breakpoint estimates and standard errors
fit_seg$psi

# 95% Confidence intervals for breakpoints
confint(fit_seg)

# Plot fitted piecewise regression lines
plot(fit_seg, conf.level = 0.95, rug = FALSE, main = "segmented fit")
points(df$x, df$y, pch = 16, col = "gray60")
lines(fit_seg, col = "blue", lwd = 2)
```

`segmented` estimates `psi1.x = 29.40 (95% CI: [24.93, 33.86])` and `psi2.x = 64.00 (95% CI: [61.55, 66.45])`.

**Important limitation: `segmented` is designed for continuous (joined) piecewise lines only.** Because `segmented()` enforces continuity across segments, it cannot model a simultaneous slope change and discontinuous vertical jump. To bridge the step down at $x = 70$, `segmented` places `psi2.x` earlier (around 64) with a steep connecting slope. If your data contains discontinuous step jumps between segments, use `mcp` or `strucchange`.

**Key capabilities:**

- **Workflow integration**: Works directly on fitted `lm`, `glm`, and `coxph` objects.
- **Execution speed**: Uses iteratively reweighted least squares for fast estimation on large sample sizes.
- **Standard frequentist output**: Returns point estimates, standard errors, and confidence intervals via `confint()`.

**Constraints:**

- Enforces continuous joined slopes; does not parameterize discontinuous step jumps between slope segments.
- Does not support Bayesian priors, exact posterior distributions on change points, or varying group-level change-point distributions.


## Other Segmented Regression Packages on CRAN

- **`segmentr`**: Exact and heuristic segmentation algorithms under customizable segment loss functions.
- **`BayesPieceHazSelect`**: Bayesian piecewise constant hazard rate modeling with variable selection.
- **`BreakPoints`**: Multiple structural change point estimation in linear and generalized linear models.


# Application 2: Time Series Decomposition & Structural Breaks

Decomposition and structural break packages analyze time-ordered series to separate long-term trends, seasonal components, and parameter instability without requiring predefined segment boundaries.

To illustrate time series modeling, we use the classic annual Nile river streamflow dataset (1871–1970, $N = 100$), which features a well-documented abrupt downward shift around 1898 following the construction of the Aswan dam:

```{r}
data("Nile")
df_nile = data.frame(
  year = as.numeric(time(Nile)),
  flow = as.numeric(Nile)
)

plot(df_nile$year, df_nile$flow, type = "l", lwd = 1.5, xlab = "Year", ylab = "River Flow (10^8 m^3)", main = "Nile River Annual Flow (1871-1970)")
abline(v = 1898, col = "firebrick", lty = 2, lwd = 1.5)
```

While the Nile dataset reflects a single level shift, real-world time series often feature richer dynamics such as **variance regime changes** (e.g. macroeconomic volatility shocks), **shifts in serial persistence** (modeled via `mcp`'s `~ ar(1)` or `~ ar(2)`), and **seasonal cycles coupled with abrupt policy interventions** (e.g. monthly road casualty drops following mandatory seatbelt legislation in `Seatbelts`).


## `mcp` on Time Series

`mcp` can model time series change points in intercepts, slopes, autoregressive parameters (`~ ar(1)`), or residual variance (`~ sigma()`):

```{r, message = FALSE, warning = FALSE, results = 'hide'}
# 2-segment plateau model for Nile flow
model_nile = list(
  flow ~ 1,      # Pre-break mean
  ~ 1            # Post-break mean
)

fit_mcp_nile = mcp(model_nile, data = df_nile, par_x = "year", seed = 42, adapt = 500, iter = 1000)
```

```{r}
summary(fit_mcp_nile)
set.seed(42)
plot(fit_mcp_nile)
```

`mcp` dates the change point to **1898.3 [1896.4, 1899.7]**, estimating a pre-1898 mean flow of $1094$ and a post-1898 mean flow of $851$.


## `strucchange`: Econometric Testing for Parameter Stability

`strucchange` provides formal hypothesis tests and dating algorithms for structural changes in linear regression models:

```{r, eval = requireNamespace("strucchange", quietly = TRUE), message = FALSE, warning = FALSE}
library(strucchange)

# Identify optimal breakpoints minimizing BIC and test stability
bp_nile = breakpoints(Nile ~ 1)
summary(bp_nile)
confint(bp_nile)
```

`strucchange` identifies the optimal breakpoint date at **1898 (95% CI: [1895, 1902])**, corresponding to observation 28. Because `strucchange` allows independent intercept and slope parameters across segments, it detects abrupt shifts with exact asymptotic confidence intervals.

**Key capabilities:**

- Formal econometric fluctuation tests (CUSUM, MOSUM, and supremum F-tests) for testing the null hypothesis of parameter constancy.
- Identifies optimal breakpoint dates minimizing information criteria (BIC, LWZ) with asymptotic confidence intervals via Bai (1997).
- Monitoring algorithms for real-time detection of parameter shifts in incoming data.

**Constraints:**

- Limited to linear models with Gaussian errors. Does not provide hierarchical varying effects or generalized response distributions.


## `Rbeast`: Bayesian Trend and Seasonal Decomposition

`Rbeast` (Bayesian Estimator of Abrupt change, Seasonal change, and Trend) uses Bayesian model averaging over piecewise linear trends, harmonic seasonal components, and changepoints. It averages over competing numbers of trend and seasonal changes and can represent nonlinear trends.

```{r, eval = requireNamespace("Rbeast", quietly = TRUE), message = FALSE, warning = FALSE}
library(Rbeast)

# Decompose series into trend and changepoint probabilities
out_beast = beast(Nile, season = "none", quiet = TRUE)
plot(out_beast)
```

`Rbeast` isolates the primary trend break at **1899**, generating a full posterior probability profile for changepoints across time without requiring pre-specification of break dates.

For a comparable `mcp` analysis, the two-plateau model fitted above is the closest match to the Nile level-shift question. It recovers a single, interpretable change point and the mean flow on each side, but it fixes the number of change points and does not decompose a seasonal or nonlinear latent trend. Thus, the two fits answer different versions of the same exploratory question: `Rbeast` asks which trend decomposition is supported by the data, whereas `mcp` estimates the parameters of a chosen step model.

**Key capabilities:**

- Decomposes regular and irregular time series into trend, periodic, and abrupt change components simultaneously.
- Infers the number and locations of changepoints via model averaging without pre-specifying candidate locations.
- Accommodates missing observations in the time series.

**Constraints:**

- Designed for univariate time series decomposition rather than general multi-predictor regression grammar.


## `CausalImpact`: Known Interventions with Control Series

`CausalImpact` answers a different question from change-point detection: given a known intervention date and unaffected control series, what would the response have been without the intervention? The following simulated example follows its recommended setup, where the intervention begins after observation 70 while the control series remains unaffected.

```{r}
set.seed(42)
df_impact = data.frame(
  time = 1:100,
  control = as.numeric(arima.sim(model = list(ar = 0.8), n = 100, sd = 3))
)
df_impact$y = 5 + 1.2 * df_impact$control + rnorm(100, sd = 3)
df_impact$y[71:100] = df_impact$y[71:100] + 12
```

```{r, eval = requireNamespace("CausalImpact", quietly = TRUE), message = FALSE, warning = FALSE}
library(CausalImpact)

fit_impact = CausalImpact(
  data = df_impact[, c("y", "control")],
  pre.period = c(1, 70),
  post.period = c(71, 100),
  model.args = list(niter = 1000)
)
plot(fit_impact)
summary(fit_impact)
```

The closest `mcp` model lets the intercept change once, while constraining the response-control relationship to remain constant. It estimates the date rather than treating it as fixed:

```{r, message = FALSE, warning = FALSE, results = 'hide'}
model_impact = list(
  y ~ 1 + control,
  ~ 1 + control
)
fit_mcp_impact = mcp(
  model_impact,
  data = df_impact,
  par_x = "time",
  prior = list(control_2 = "control_1"),
  seed = 42,
  adapt = 500,
  iter = 1000
)
```

```{r}
summary(fit_mcp_impact)
set.seed(42)
plot(fit_mcp_impact)
```

`mcp` recovers the change point near 70 and the post-intervention intercept increase. That is useful descriptive evidence, but it is not a causal effect: `mcp` has no counterfactual control-series assumption. Conversely, `CausalImpact` estimates the counterfactual effect but depends critically on controls being unaffected by the intervention and retaining a stable relation to the response.


## `dsp`: Dynamic Shrinkage Processes

`dsp` implements Bayesian trend filtering using dynamic shrinkage priors on successive differences in latent states.

**Key capabilities:**

- Shrinks adjacent differences toward zero while allowing sparse, large changes to escape shrinkage.
- Discovers localized trend shifts and heteroscedasticity without explicit changepoint parameterization.


## Other Structural Break and Time Series Packages on CRAN

- **`EnvCpt`**: Automated model selection comparing combinations of trend, mean, and AR(1)/AR(2) change points for climate and environmental series.
- **`mosum`**: Moving Sum statistics for testing and dating mean/variance shifts with confidence intervals.
- **`trendsegmentR`**: Linear wavelet-based detection for piecewise-linear trends and slope breaks with robust variance estimators.
- **`multibreakeR`**: Econometric testing and estimation of multiple structural breaks in multivariate time series.


# Application 3: Algorithmic Change Point Detection & Meta-Packages

Detection packages focus on identifying unknown shift locations in large sequences with high computational efficiency.

To illustrate high-throughput detection across sequences with many change points, we use the `coriell` genomic microarray dataset (Snijders et al. 2001, $N = 2,077$), which exhibits dozens of abrupt copy-number state transitions along chromosomes:

```{r, eval = requireNamespace("bcp", quietly = TRUE) && requireNamespace("changepoint", quietly = TRUE), message = FALSE, warning = FALSE}
data("coriell", package = "bcp")
df_dna = data.frame(
  pos = seq_along(na.omit(coriell$Coriell.13330)),
  y = as.numeric(na.omit(coriell$Coriell.13330))
)
```


## `changepoint`: High-Throughput Detection via PELT

`changepoint` implements fast dynamic programming and binary segmentation algorithms (PELT, BinSeg), segmenting thousands of observations in milliseconds:

```{r, eval = requireNamespace("changepoint", quietly = TRUE), message = FALSE, warning = FALSE}
library(changepoint)

# Detect mean and variance shifts in milliseconds using PELT
fit_cpt = cpt.meanvar(df_dna$y, method = "PELT", penalty = "Manual", pen.value = 15 * log(nrow(df_dna)))
plot(fit_cpt, xlab = "Genomic Index", ylab = "Log2 Ratio", main = "changepoint PELT on Coriell DNA (N = 2,077)")
```


## `bcp`: Bayesian Change Point Detection

`bcp` implements the Barry and Hartigan (1993) product partition model via Markov chain Monte Carlo, estimating the exact posterior probability profile of a change point at every observation:

```{r, eval = requireNamespace("bcp", quietly = TRUE), message = FALSE, warning = FALSE}
library(bcp)

fit_bcp = bcp(df_dna$y, p0 = 0.01, burnin = 100, mcmc = 500)
plot(fit_bcp, main = "bcp Posterior Probability Profile on Coriell DNA")
```


## `not`: Fast Detection of Slope Changes and Steps

`not` is a fast, univariate detection method for an unknown number of change points. Unlike `changepoint`, it includes piecewise-linear signals with either joined or discontinuous lines. This example contains both a slope change and an intercept jump at observation 100.

```{r}
set.seed(42)
df_not = data.frame(time = 1:200)
df_not$y = ifelse(
  df_not$time <= 100,
  5 + 0.03 * df_not$time,
  11 + 0.18 * (df_not$time - 100)
) + rnorm(200, sd = 0.8)
plot(y ~ time, data = df_not, pch = 16, col = "gray60")
abline(v = 100, lty = 2, col = "firebrick")
```

```{r, eval = requireNamespace("not", quietly = TRUE), message = FALSE, warning = FALSE}
library(not)

fit_not = not(df_not$y, contrast = "pcwsLinMean")
cp_not = features(fit_not, penalty = "sic")$cpt
cp_not
plot(fit_not, cpt = cp_not)
```

The matching `mcp` formula is a two-segment model with a new intercept and slope after the change point. Unlike the joined-slope syntax (`~ 0 + time`), `~ 1 + time` permits the discontinuity present here.

```{r, message = FALSE, warning = FALSE, results = 'hide'}
model_not = list(
  y ~ 1 + time,
  ~ 1 + time
)
fit_mcp_not = mcp(model_not, data = df_not, par_x = "time", seed = 42, adapt = 500, iter = 1000)
```

```{r}
summary(fit_mcp_not)
set.seed(42)
plot(fit_mcp_not)
```

Both methods recover a change near 100. `not` searches over the number of changes in one fast procedure, whereas `mcp` reports a posterior for the pre-specified single change point and estimates the two regression lines. For many unknown breaks, the computational advantage remains with `not`.


## `stepR`: Multiscale Inference for a Change in Mean and Variance

`stepR` estimates a piecewise-constant mean while allowing the residual standard deviation to change by segment through its HSMUCE family. This provides simultaneous frequentist confidence statements while `mcp` models the mean and log-SD as separate distributional parameters.

```{r}
set.seed(42)
df_step = data.frame(time = 1:160)
df_step$y = c(rnorm(80, mean = 0, sd = 0.5), rnorm(80, mean = 2, sd = 1.5))
plot(y ~ time, data = df_step, pch = 16, col = "gray60")
abline(v = 80, lty = 2, col = "firebrick")
```

```{r, eval = requireNamespace("stepR", quietly = TRUE), message = FALSE, warning = FALSE}
library(stepR)

fit_step = stepFit(
  df_step$y,
  x = df_step$time,
  family = "hsmuce",
  alpha = 0.1,
  jumpint = TRUE,
  confband = TRUE
)
plot(df_step$time, df_step$y, pch = 16, col = "gray60")
lines(fit_step, col = "blue", lwd = 2)
points(jumpint(fit_step), col = "blue", lwd = 2)
```

For `mcp`, the matching model instantiates both a new mean and a new log-SD at the same change point. The `sigma()` terms are explicit because the implicit Gaussian residual SD is shared across segments.

```{r, message = FALSE, warning = FALSE, results = 'hide'}
model_step = list(
  y ~ 1 + sigma(1),
  ~ 1 + sigma(1)
)
fit_mcp_step = mcp(model_step, data = df_step, par_x = "time", seed = 42, adapt = 500, iter = 1000)
```

```{r}
summary(fit_mcp_step)
set.seed(42)
plot(fit_mcp_step, q_predict = TRUE)
```

Both fits identify the break near 80. `stepR` is specialized to piecewise-constant signals and reports multiscale confidence intervals/bands; `mcp` gives posterior uncertainty for the location, means, and each segment's residual SD, and can extend the same model to slopes, covariates, or GLM likelihoods.


## `ecp`: Multivariate, Distribution-Free Detection

`ecp` detects changes in the joint distribution of multivariate observations, not merely a change in each marginal mean. The following signal changes both means and variances at observation 80. It has no cross-series correlation, but the simultaneous multivariate shift is beyond an ordinary univariate `mcp` likelihood.

```{r}
set.seed(42)
df_ecp = data.frame(time = 1:160)
df_ecp$y1 = c(rnorm(80, mean = 0, sd = 0.5), rnorm(80, mean = 2, sd = 1.5))
df_ecp$y2 = c(rnorm(80, mean = 1, sd = 0.5), rnorm(80, mean = -1, sd = 1.5))
matplot(df_ecp$time, df_ecp[, c("y1", "y2")], type = "p", pch = 16, col = c("steelblue", "darkorange"), xlab = "Time", ylab = "Response")
abline(v = 80, lty = 2, col = "firebrick")
legend("topleft", legend = c("y1", "y2"), col = c("steelblue", "darkorange"), pch = 16, bty = "n")
```

```{r, eval = requireNamespace("ecp", quietly = TRUE), message = FALSE, warning = FALSE}
library(ecp)

fit_ecp = e.divisive(
  as.matrix(df_ecp[, c("y1", "y2")]),
  sig.lvl = 0.05,
  R = 199,
  min.size = 20
)
fit_ecp$estimates
```

`mcp` has no multivariate joint-response likelihood, so the closest coherent comparison fits one response at a time. For `y1`, the model below allows the mean and residual SD to change at one common, unknown location.

```{r, message = FALSE, warning = FALSE, results = 'hide'}
model_ecp = list(
  y1 ~ 1 + sigma(1),
  ~ 1 + sigma(1)
)
fit_mcp_ecp = mcp(model_ecp, data = df_ecp, par_x = "time", seed = 42, adapt = 500, iter = 1000)
```

```{r}
summary(fit_mcp_ecp)
set.seed(42)
plot(fit_mcp_ecp, q_predict = TRUE)
```

`mcp` recovers the shared mean/variance break in `y1`; repeating the model for `y2` checks whether its location agrees. `ecp` instead uses both series jointly and can detect changes in a joint distribution even when a univariate mean/variance model is misspecified. It therefore complements rather than competes directly with `mcp`.


## `smoothbp_ss`: Spike-and-Slab Breakpoint Selection

For discovering whether candidate breakpoints exist within a single Bayesian piecewise model, `smoothbp` provides spike-and-slab variable selection (`smoothbp_ss`):

```{r, eval = requireNamespace("smoothbp", quietly = TRUE), message = FALSE, warning = FALSE}
library(smoothbp)

fit_ss = smoothbp_ss(
  formula = y ~ x,
  deltas = list(~ 1, ~ 1),
  omega = list(fixed(30), fixed(70)),
  rho = list(~ 1, ~ 1),
  data = df,
  iter = 1000,
  warmup = 500,
  chains = 2,
  seed = 42,
  .verbose = FALSE
)
pip(fit_ss)
```

The posterior inclusion probabilities (PIPs) evaluate the probability that each candidate breakpoint is active.


## When NOT to Use `mcp` or `smoothbp`

Neither package is designed for high-throughput detection across large sequences with many unknown change points ($\ge 10$). Specifying 10+ ordered change points in MCMC causes severe label switching, slow sampling, and poor mixing over $2^{10}$ permutations. For signals with dozens of unknown shifts across thousands of observations, use **`changepoint`**, **`not`**, or **`bcp`**.


## Other Algorithmic Detection Packages on CRAN

- **`ggchangepoint` / `tidychangepoint`**: Meta-packages providing standardized tidy interfaces and ggplot visualizations across multiple underlying detection packages.
- **`breakfast` & `IDetect`**: Isolate-Detect and Wild Binary Segmentation (WBS) algorithms for multiscale change point detection.
- **`cpm`**: Sequential and streaming change point detection with controlled false alarm rates.
- **`changepoints` / `changepointsVar` / `InspectChangepoint`**: Multivariate, high-dimensional, and VAR covariance change point detection.
- **`changepoint.geo`**: High-dimensional change point detection via geometric projection.
- **`ChangepointTesting`**: Sequential hypothesis testing for changes in variance and mean.


# Summary: Selecting a Package

- **Bayesian piecewise regression**: Use **`mcp`** for full generative modeling (arbitrary per-segment formulas, GLMs, group effects, ARMA, heteroscedasticity) or **`smoothbp`** for continuous smooth transitions and spike-and-slab selection.
- **Bayesian point-wise break detection**: Use **`bcp`** for observation-level posterior break probability profiles, or **`Rbeast`** for Bayesian trend and seasonal decomposition.
- **Known intervention with controls**: Use **`CausalImpact`** to estimate a causal counterfactual after a pre-specified intervention; a change point alone is not a causal effect estimate.
- **Frequentist continuous broken-line regression**: Use **`segmented`** when estimating joined slope changes on standard `lm()` / `glm()` fits with fast execution.
- **Frequentist econometric testing**: Use **`strucchange`** for formal hypothesis tests and dating of structural breaks in regression models.
- **High-throughput signal detection**: Use **`changepoint`**, **`not`**, or **`stepR`** for rapid univariate detection; use **`ecp`** when a joint multivariate distribution may change.
