---
title: "Poisson and negative-binomial change point analysis"
date: "`r format(file.mtime(knitr::current_input()), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson and negative-binomial change point analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: none
---

The Poisson and negative-binomial distributions model counts observed within comparable units of time, space, or exposure. The worked example below uses Poisson regression; a later section explains the additional `shape` parameter and priors in `negbinomial()`.

# Coal mining disasters

A dataset on coal mining disasters has grown very popular in the change point literature (available in `boot::coal`). It contains a timestamp of each coal mining disaster from 1851 to 1962. By binning the number of events within each year (fixed time frame), we have something very Poisson-friendly:

```{r, message=FALSE, warning=FALSE}
# Number of disasters by year
library(dplyr, warn.conflicts = FALSE)
df = round(boot::coal) %>% 
  group_by(date) %>% 
  count()

# See it
head(df)
```

The number of events (`n`) as a function of year (`date`) is typically modeled as a change between two intercepts. This is very simple to do in `mcp`:

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic
```

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
model = list(
  n ~ 1,  # intercept-only
  ~ 1  # intercept-only
)

fit = mcp(model, data = df, family = poisson(), par_x = "date", seed = 42)
```

Let us see the two intercepts (lambda in log-units) and the change point (in years):

```{r}
result = summary(fit)
```

We can see that the model ran well with good convergence and a large number of effective samples. At a first glance, the change point is estimated to lie between the years 1880 and 1895 (approximately).

Let us take a more direct look, using the default `mcp` plot:

```{r}
set.seed(42)
plot(fit)
```

It seems to fit the data well, but we can see that the change point probability "lumps" around particular data points. Years with a very low number of disasters abruptly increase the probability that the change to a lower disaster rate has taken place. The posterior distributions of change points regularly take these "weird" forms, i.e., not well-described by our toolbox of parameterized distributions.

We can see this more clearly if plotting the posteriors. We include a traceplot too, just to check convergence visually.

```{r}
set.seed(42)
plot_pars(fit)
```

# Priors for Poisson models

`poisson()` defaults to `link = 'log'`, meaning that we have to exponentiate the estimates to get the "raw" Poisson parameter $\lambda$. $\lambda$ has the nice property of being the mean number of events. So we see that the mean number of events in segment 1 is `exp(result$mean[2])` (`r exp(result$mean[2])`) and it is `exp(result$mean[3])` (`r exp(result$mean[3])`) for segment 2.

The intercept prior is a Student-t distribution centered on the rounded median of `log(pmax(n, 0.1))`, with scale equal to the maximum of 2.5 and its rounded MAD. This follows the robust calibration in `brms`; `pmax()` keeps zeros finite and the minimum scale avoids a narrow prior.

Categorical contrasts use scale 2.5; numeric coefficients are scaled to a representative predictor change. These are proper Student-t priors where `brms` is flat, allowing prior sampling and mild regularization of short segments.

```{r}
prior_summary(fit)
```

As always, the prior on the change point forces it to occur in the observed range. The coefficient priors are deliberately broad defaults, so update them with more informed priors for your particular case when possible, e.g.:

```{r, cache = TRUE, eval=FALSE}
prior = list(
  cp_1 = "dnorm(1900, 30) T(min(date), 1925)"
)
fit_with_prior = mcp(model, data = df, family = poisson(), prior = prior, par_x = "date")
```

# Negative-binomial extension

`negbinomial()` uses log links for both its conditional mean `mu` and overdispersion `shape`, where $\operatorname{Var}(y) = \mu + \mu^2 / \mathit{shape}$. Its mean priors are the same as for `poisson()`.

When no `shape()` formula is supplied, the response-scale shape has an inverse-gamma(0.4, 0.3) prior, matching `brms`. It covers both strong overdispersion and the large-shape Poisson limit; `fit$prior` shows its log-shape representation as `dloginvgamma(0.4, 0.3)`.

An explicit `shape()` formula models log-shape. Its intercept defaults to `dt(0, 2.5, 3)` and coefficients use the usual proper scaling. `brms` uses the same intercept prior but flat coefficients. Shape regression can be weakly identified, so informed priors may be useful.

# Modeling rates using `offset()`

When modeling event counts where exposure (e.g., person-years, area, or population size) varies across observations, you can model rates $\lambda = \text{Count} / \text{exposure}$ on the log link by supplying `offset(log(exposure))`:

```r
# Example: Event counts over varying observation durations (as here) or population
model = list(
  events ~ 1 + year + offset(log(exposure)),  # Segment 1
  ~ 1 + year                                  # Segment 2
)

fit = mcp(model, data = df, family = poisson())  # implicit link = "log"
```

Adding a categorical predictor to the model, you can also model rate ratios.

> **Note on `par_x` vs. cumulative exposure:**  
> The offset is evaluated per row as local exposure for that observation. If `par_x` represents continuous cumulative time from an origin (e.g., $[0, t_i]$), using `offset(log(par_x))` does **not** integrate piecewise rates across prior segments; instead, the active segment's rate is applied to the total duration. If your data consists of discrete time intervals (bins) of duration $\Delta t_i$, using `offset(log(delta_t))` is valid.

# Model comparison

Despite the popularity of this dataset, a question rarely asked is what the evidence is that there is a change point at all. Let us fit two no-changepoint models and use approximate leave-one-out cross-validation to see how the predictive performance of the two models compare.

A flat model and a one-decay model:

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
# Fit an intercept-only model
fit_flat = mcp(list(n ~ 1), data = df, family=poisson(), par_x = "date", seed = 42)
fit_decay = mcp(list(n ~ 1 + date), data = df, family = poisson(), seed = 42)


set.seed(42)
plot(fit_flat) + plot(fit_decay)
```

Not we compute and compare the LOO ELPDs:

```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit_loo = loo(fit)
fit_flat_loo = loo(fit_flat)
fit_decay_loo = loo(fit_decay)
```

```{r, results='hold'}
loo::loo_compare(fit_loo, fit_flat_loo, fit_decay_loo)
```

The change point model seems to be preferred with a ratio of around 1.9 over the decay model and 2.5 over the flat model. Another approach is to look at the model weights:

```{r, results='hold'}
loo_list = list(fit_loo, fit_flat_loo, fit_decay_loo)
loo::loo_model_weights(loo_list, method="pseudobma")
```

Again, unsurprisingly, the change point model is preferred and they show the same ranking as implied by `loo_compare`.

# JAGS code for the Poisson example

Here is the JAGS code for the full Poisson change-point model above.

```{r}
fit$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
