---
title: "Poisson change point analysis with mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson change point analysis with mcp}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Finding change points in Poisson models using mcp
The Poisson distribution models the number of events within similar-sized time frames. 

# Coal mining disasters
A dataset if coal mining disasters has grown very popular in the change point literature (available in `boot::coal`). It contains a timestamp of each coal mining disaster from 1851 to 1962. By binning the number of events within each year (fixed time frame), we have something very Poisson-friendly:

```{r}
library(tidyverse)
data = round(boot::coal) %>% 
  group_by(date) %>% 
  count()
```


The number of events (`n`) as a function of year (`date`) is typically modeled as a change between two intercepts. This is very simple to do in `mcp`:
```{r}
library(mcp)
segments = list(
  n ~ 1,
  1 ~ 1
)

fit = mcp(segments, data, family = poisson(), par_x = "date")
```

Let us see the two intercepts (lambda in log-units) and the change point (in years):

```{r}
summary(fit)
```

We can see that the model ran well with good convergence and a large number of effective samples. At a first glance, the change point is estimated to lie between the years 1880 and 1895.

Let us take a more direct look, using the default `mcp` plot:

```{r}
plot(fit)
```

It seems to fit the data well, but we can see that the change point probability "lumps" around particular data points. Years with a very low number of disasters abruptly increase the probability that the change has taken place.

We can see this more clearly if plotting the posteriors. We include a traceplot too, just to check convergence visually.

```{r}
plot(fit, "combo")
```

A question rarely asked is what the evidence is that there is a change point at all. Let us fit an intercept-only model and use approximate leave-one-out cross-validation to see how the predictive performance of the two models compare:

```{r}
# Fit an intercept-only model
segments_null = list(n ~ 1)
fit_null = mcp(segments_null, data, family=poisson(), par_x = "date")

# Compute and compare loos
fit_null$loo = loo(fit_null)
fit$loo = loo(fit)
loo::loo_compare(fit_null$loo, fit$loo)
```


