---
title: "Poisson change point analysis with mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson change point analysis with mcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The Poisson distribution models the number of events within similar-sized time frames. 


# Coal mining disasters
A dataset on coal mining disasters has grown very popular in the change point literature (available in `boot::coal`). It contains a timestamp of each coal mining disaster from 1851 to 1962. By binning the number of events within each year (fixed time frame), we have something very Poisson-friendly:

```{r, message=FALSE, warning=FALSE}
# Number of disasters by year
library(dplyr, warn.conflicts = FALSE)
df = round(boot::coal) %>% 
  group_by(date) %>% 
  count()

# See it
df
```


The number of events (`n`) as a function of year (`date`) is typically modeled as a change between two intercepts. This is very simple to do in `mcp`:

```{r}
library(mcp)
options(mc.cores = 3)  # Speed up sampling!
```

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
model = list(
  n ~ 1,  # intercept-only
  ~ 1  # intercept-only
)

fit = mcp(model, data = df, family = poisson(), par_x = "date")
```

Let us see the two intercepts (lambda in log-units) and the change point (in years):

```{r}
result = summary(fit)
```

We can see that the model ran well with good convergence and a large number of effective samples. At a first glance, the change point is estimated to lie between the years 1880 and 1895 (approximately).

Let us take a more direct look, using the default `mcp` plot:

```{r}
plot(fit)
```

It seems to fit the data well, but we can see that the change point probability "lumps" around particular data points. Years with a very low number of disasters abruptly increase the probability that the change to a lower disaster rate has taken place. The posterior distributions of change points regularly take these "weird" forms, i.e., not well-described by our toolbox of parameterized distributions.

We can see this more clearly if plotting the posteriors. We include a traceplot too, just to check convergence visually.

```{r}
plot_pars(fit)
```


# Priors
`poisson()` defaults to `link = 'log'`, meaning that we have to exponentiate the estimates to get the "raw" Poisson parameter $\lambda$. $\lambda$ has the nice property of being the mean number of events. So we see that the mean number of events in segment 1 is `exp(result$mean[2])` (`r exp(result$mean[2])`) and it is `exp(result$mean[3])` (`r exp(result$mean[3])`) for segment 2.

Default priors were used. They are normals with a standard deviation of 10. I.e. with 68% probability mass between `exp(10) = 22026` and `exp(-10) = 1 / 22026`:

```{r}
cbind(fit$prior)
```

As always, the prior on the change point forces it to occur in the observed range. These priors are very vague, so update with more informed priors for your particular case, e.g.:

```{r, cache = TRUE, eval=FALSE}
prior = list(
  cp_1 = "dnorm(1900, 30) T(MINX, 1925)"
)
fit_with_prior = mcp(model, data = df, prior, poisson(), par_x = "date")
```


# Model comparison

Despite the popularity of this dataset, a question rarely asked is what the evidence is that there is a change point at all. Let us fit two no-changepoint models and use approximate leave-one-out cross-validation to see how the predictive performance of the two models compare. 

A flat model and a one-decay model:

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
# Fit an intercept-only model
fit_flat = mcp(list(n ~ 1), data = df, family=poisson(), par_x = "date")
fit_decay = mcp(list(n ~ 1 + date), data = df, family = poisson())


plot(fit_flat) + plot(fit_decay)
```


Not we compute and compare the LOO ELPDs:

```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit$loo = loo(fit)
fit_flat$loo = loo(fit_flat)
fit_decay$loo = loo(fit_decay)
```
```{r}
loo::loo_compare(fit$loo, fit_flat$loo, fit_decay$loo)
```


The change point model seems to be preferred with a ratio of around 1.7 over the decay model and 2.5 over the flat model. Another approach is to look at the model weights:

```{r}
loo_list = list(fit$loo, fit_flat$loo, fit_decay$loo)
loo::loo_model_weights(loo_list, method="pseudobma")
```

Again, unsurprisingly, the change point model is preferred and they show the same ranking as implied by `loo_compare`.


# JAGS code
Here is the JAGS code for the full model above.

```{r}
cat(fit$jags_code)
```

