---
title: "Working with priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: none
---

*NOTE: the JAGS-specific prior interface below is likely to be deprecated in the future, when additional backends are added beyond JAGS.*

# Setting a prior

`mcp` takes priors in the form of a named list. The names are the parameter names, and the values are JAGS code. Here is a fairly complicated example, to demonstrate the various ways priors can be used:

```{r}
model = list(
  y ~ 1 + x,  # Intercept_1 + x_1
  ~ 1 + x,  # cp_1, Intercept_2, and x_2
  ~ 1 + x  # cp_2
)

prior = list(
  Intercept_1 = "dnorm(0, 5) T(, 10)",  # Intercept; less than 10
  x_1 = "dbeta(2, 5)",  # slope: beta with right skew
  cp_1 = "dunif(min(x), cp_2)",  # change point between smallest x and cp_2
  x_2 = "dt(0, 1, 3) T(x_1, )",  # slope 2 > slope 1 and t-distributed
  cp_2 = 80,  # A constant (set; not estimated)
  x_3 = "x_2"  # continue same slope
  # Intercept_2 and Intercept_3 not specified. Use default.
)
```

The values use mcp's JAGS-string syntax, so JAGS distributions are generally allowed. See the [JAGS user manual](https://altushost-swe.dl.sourceforge.net/project/mcmc-jags/Manuals/4.x/jags_user_manual.pdf) for more details. Prior strings use R-like conventional distribution scales even though JAGS has a different parameterization underneath: SD for `dnorm()`, scale for `dt()`, `ddexp()`, and `dlogis()`, and log-SD for `dlnorm()`. When generating JAGS code, `mcp` converts these to inverse variance for `dnorm()`, `dt()`, and `dlnorm()`, and inverse scale for `ddexp()` and `dlogis()`. You can inspect the translation by comparing `fit$prior` with `fit$jags_code`.

Other notes:

- Default population-level change-point priors are ordered. For user priors, `mcp` adds truncation (e.g., `T(cp_1, )`) only when the prior has neither explicit truncation nor an inherently bounded form such as `dunif()` or `dirichlet()`. User-specified population-level bounds are respected as written. When change points have group-level deviations, their realized locations are constrained to remain ordered within every group.

- `mcp` allows inserting data-dependent values: for example, `min(x)`, `max(x)`, `median(y)`, `mad(y)`, `max(x) - min(x)`, `n_segments()`, and `n_cp()`. `mcp` resolves these expressions before generating JAGS code. The older `mcp` v0.3.4 uppercase constants (`MINX`, `MAXX`, and so on) remain temporarily supported with a deprecation warning.

- You can fix any parameter to a specific value. Simply set it to a numerical value (as `cp_2` above). A constant is a 100% prior belief in that value, and it will therefore not be estimated.

- You can also equate one variable with another (`x_3 = "x_2"` above). You would usually do this to share parameters across segments, but you can be creative and do something like `x_3 = "x_2 + 5 - cp_1/10"` if you want. In any case, it will lead to one less parameter being estimated, i.e., one less free parameter.

Let us see the priors after running them through `mcp` and compare to the default priors:

```{r}
library(mcp)
df_dummy = data.frame(x = 1:100, y = 1:100)
empty_manual = mcp(model, data = df_dummy, prior = prior, sample = FALSE)
empty_default = mcp(model, data = df_dummy, sample = FALSE)

# Inspect resolved priors and bounds. Add rules, descriptions, sources, and
# kinds with prior_summary(empty_manual, verbose = TRUE).
prior_summary(empty_manual, verbose = TRUE)
prior_summary(empty_default, verbose = TRUE)
```

Now, let's simulate some data from `model`. The following priors are "at odds" with the actual data so as to show their effect.

```{r}
df = data.frame(x = runif(200, 0, 100), y = 1)  # 200 datapoints between 0 and 100
set.seed(42)
df$y = empty_default$simulate(empty_default, df, 
    Intercept_1 = 20, Intercept_2 = 22, Intercept_3 = 30,  # intercepts
    x_1 = -0.5, x_2 = 0.5, x_3 = 0,  # slopes
    cp_1 = 35, cp_2 = 70,  # change points
    sigma = 5)

head(df)
```

Sample the prior and posterior. We give the manual fit a longer warmup since it is harder to find the right posterior under these weird prior constraints (priors will usually improve sampling efficiency).

```{r, cache = TRUE, message=FALSE, warning=FALSE, results=FALSE}
future::plan(future::multisession, workers = 3)
fit_manual = mcp(model, data = df, sample = "both", warmup = 10000, prior = prior, seed = 42)
fit_default = mcp(model, data = df, sample = "both", warmup = 10000, seed = 42)
```

First, let's look at the priors side by side. Notice the use of `prior = TRUE` to show prior samples. This works in `plot()`, `plot_pars()`, and `summary()` among others.

```{r, fig.height=12, fig.width=6}
library(ggplot2)
set.seed(42)
pp_default = plot_pars(fit_default, type = "dens_overlay", prior = TRUE) + 
  ggtitle("Default priors")

set.seed(42)
pp_manual = plot_pars(fit_manual, type = "dens_overlay", prior = TRUE) +
  ggtitle("Manual priors")

pp_default + pp_manual
```

Here are the resulting posterior fits:

```{r}
set.seed(42)
plot_default = plot(fit_default) + ggtitle("Default priors")

set.seed(42)
plot_manual = plot(fit_manual) + ggtitle("Manual priors")

plot_default + plot_manual
```

We see the effects of the priors:

- The intercept `Intercept_1` was truncated to be below 10.
- The slope `x_1` is bound to be non-negative (because `dbeta`).
- The slopes `x_2` and `x_3` were forced to be identical.
- The change point `cp_2` was a constant, so there is no uncertainty there.

This is a contrived example. Usually setting priors manually aims to sample the "correct" posterior.

# Default priors on change points {#cp_prior}

Change point locations must lie within the observed range `[min(x), max(x)]` and maintain strict order `cp_1 < cp_2 < ... < cp_n`. Without this rule, change point locations would be unidentifiable. If you inspect `fit$jags_code`, you will see that `mcp` inserts boundary markers at `cp_0 = min(x)` and `cp_{n+1} = max(x)` to ensure the ordering constraint.

For any given range, introducing more change points makes the prior for each change point more informative (as shown in the figure below). `mcp` provides two built-in population-level prior families:

- **One change point:** Both specifications default to `cp_1 = uniform(min(x), max(x))`. Its location is uniform over the observed x-range.
- **The t-tail prior (default for 2+ change points):** Sequentially truncated Student-t distributions centered at `min(x)`. Specifically, `cp_i = student_t(df = n_cp() - 1, location = min(x), scale = (max(x) - min(x)) / n_cp()) T(cp_{i-1}, max(x))`.
- **The Dirichlet prior:** A joint prior on segment spacings specified via `prior = list(cp_1 = "dirichlet(1)", cp_2 = "dirichlet(1)", ...)`.

```{r, echo = FALSE, results=FALSE, warning=FALSE, message=FALSE}
df_dummy = data.frame(x = c(0, 0.5, 1), y = rnorm(3))

# Sample priors for two default change points
model_two = list(y ~ x, ~ 1, ~1)
fit_two = mcp(model_two, data = df_dummy, sample = "prior", chains = 1, iter = 300000, seed = 42)

# Sample priors for five default change points
model_five = list(y ~ x, ~ 1, ~1, ~1, ~1, ~1)
fit_five = mcp(model_five, data = df_dummy, sample = "prior", chains = 1, iter = 300000, seed = 42)
```

```{r, echo = FALSE, fig.height=6, fig.width=7}
par(mfrow = c(2,2))
MINX = min(df_dummy$x)
MAXX = max(df_dummy$x)


###############
# DIRICH TWO #
###############

# Initiate for cp_1
curve(dbeta(x, 1, 2) / 2, 
      ylab = "density",
      xlab = NA,
      main = "Dirichlet two")

text(0.1, 0.7, labels = "cp_1")

# Add others
curve(dbeta(x, 2, 1) / 2, add = T); text(0.9, 0.7, labels = "cp_2")

# The sum
curve(dunif(x, 0, 1), lty = 2, add = T); text(0.5, 0.95, labels = "sum")

# Add extra x-axis
axis(1, at = seq(0, 1, by = 0.2), line=3)
mtext("Beta", 1, line = 1, at = -0.2)
mtext("Data", 1, line = 3, at = -0.2)


###############
# DIRICH FIVE #
###############

# Initiate for cp_1
curve(dbeta(x, 1, 5) / 5, 
      ylab = "density",
      xlab = NA,
      main = "Dirichlet five")

text(0.1, 0.8, labels = "cp_1")

# Add others
curve(dbeta(x, 2, 4) / 5, add = T); text(0.25, 0.48, labels = "cp_2")
curve(dbeta(x, 3, 3) / 5, add = T); text(0.5, 0.43, labels = "cp_3") 
curve(dbeta(x, 4, 2) / 5, add = T); text(0.75, 0.48, labels = "cp_4") 
curve(dbeta(x, 5, 1) / 5, add = T); text(0.9, 0.8, labels = "cp_5") 

# The sum
curve(dunif(x, 0, 1), lty = 2, add = T); text(0.5, 0.95, labels = "sum")

# Add extra x-axis
axis(1, at = seq(0, 1, by = 0.2), line=3)
mtext("Beta", 1, line = 1, at = -0.2)
mtext("Data", 1, line = 3, at = -0.2)


###############
# DEFAULT TWO #
###############
draws_two = as_draws_df(fit_two)
y1 = density(draws_two$cp_1, from = MINX, to = MAXX)
y2 = density(draws_two$cp_2, from = MINX, to = MAXX)
ysum = (y1$y + y2$y) / 2
plot(y1$x, ysum, 
     type = "l", lty = 2, 
     xlab = NA, ylab = "Density", main = "t-tail two",
     ylim = c(0, max(ysum))
)

lines(y1$x, y1$y / 2)
lines(y2$x, y2$y / 2)

text(0.5, 0.9, labels = "sum")
text(0.1, 0.7, labels = "cp_1")
text(0.9, 0.6, labels = "cp_2")



################
# DEFAULT FIVE #
################
draws_five = as_draws_df(fit_five)
z1 = density(draws_five$cp_1)
z2 = density(draws_five$cp_2)
z3 = density(draws_five$cp_3)
z4 = density(draws_five$cp_4)
z5 = density(draws_five$cp_5)
zsum = (z1$y + z2$y + z3$y + z4$y + z5$y) / 4

plot(z1$x, zsum, 
     type = "l", lty = 2, 
     xlab = NA, ylab = "Density", main = "t-tail five",
     ylim = c(0, max(zsum))); 
text(0.3, 1.3, labels = "sum")

lines(z5$x, z5$y / 5); text(0.9, 0.8, labels = "cp_5")
lines(z4$x, z4$y / 5); text(0.65, 0.45, labels = "cp_4")
lines(z3$x, z3$y / 5); text(0.45, 0.5, labels = "cp_3")
lines(z3$x, z2$y / 5); text(0.25, 0.6, labels = "cp_2")
lines(z2$x, z1$y / 5); text(0.1, 0.8, labels = "cp_1")
par(mfrow = c(1,1))
```

## The t-tail prior on 2+ change points (default)

For 2+ change points, the default rule (on all change points) is `cp_i = student_t(df = n_cp() - 1, location = min(x), scale = (max(x) - min(x)) / n_cp())`, bounded below by the preceding change point and above by `max(x)`.

- **Degrees of freedom:** It is t-distributed with $N - 1$ degrees of freedom (`n_cp() - 1`). Together with the decreasing scale, this makes the underlying distributions more concentrated as the number of change points increases.
- **Scale:** The scale is `(max(x) - min(x)) / n_cp()`. This is a calibration choice, not the equal-segment spacing (which would divide by `n_cp() + 1`).
- **Location and truncation:** The location parameter is always the lowest observed `x`. Thus `cp_1` is a bounded half-t and `cp_2+` use sequentially truncated right tails of the same t (hence "t-tail prior"). Their means are shifted by truncation and are not equal to `min(x)`.
- **Properties:** The sequential truncation does not produce exact uniform order statistics; the middle segments tend to get narrower than the first/last segments. For non-small datasets, the effect on the posterior is negligible, though point-`hypothesis()` tests via Savage-Dickey density ratios would be strongly impacted. Its main advantage is sampling speed: JAGS samples this formulation much faster than the Dirichlet prior.

## Dirichlet-based prior on change points

The [Dirichlet distribution](https://en.wikipedia.org/wiki/Dirichlet_distribution) places a joint prior on positive segment spacings that sum to the observed x-range. Consequently, the derived population-level change points are naturally ordered and bounded. This construction also underlies monotonic effects in `brms` ([Bürkner & Charpentier (2019)](https://psyarxiv.com/9qkhj/)).

The Dirichlet distribution is a simplex of beta distributions (they always sum to 1), where the individual betas represent inter-changepoint distances. For $N$ change points, the distribution for change point $i$ is $\text{Beta}(i, N + 1 - i)$, so `cp_1 ~ Beta(1, N)` and `cp_N ~ Beta(N, 1)`. Scaling from $[0, 1]$ to the observed data range (`0 --> min(x)`, `1 --> max(x)`) gives the desired bounding.

To use the Dirichlet prior, specify it for all change points:

```{r, eval=FALSE}
prior_dirichlet = list(
  cp_1 = "dirichlet(1)",
  cp_2 = "dirichlet(1)",
  cp_3 = "dirichlet(1)"
)
```

The number in parentheses is a common $\alpha$ concentration parameter (all change points must share the same $\alpha$). Values below 1 favor uneven spacings, `dirichlet(1)` treats spacings exchangeably (giving the ordered-uniform distribution), and values above 1 favor more evenly spaced change points.

## Manual priors on change points

You can customize change point priors as shown in the initial example. Beware that the nature of the priors changes when truncation is applied; use `plot_pars(fit, prior = TRUE)` to inspect the realized prior distributions.

For an otherwise unbounded user prior such as `cp_2 = "dnorm(40, 10)"`, `mcp` automatically adds default bounds equivalent to `cp_2 = "dnorm(40, 10) T(cp_1, max(x))"`. An explicit truncation, including `T(,)`, or a `dunif()` prior is used as written. Relaxing the order can cause label-switching between segments and make sampling difficult, so inspect the resulting prior carefully.

# Default priors on linear predictors

The defaults borrow the robust calibration used by `brms`, adapted to the local parameterization of `mcp`. You can inspect all resolved priors for a model using `prior_summary(fit, verbose = TRUE)`.

- **Intercepts:** Unlike the improper flat default coefficient priors in `brms`, `mcp` uses proper Student-t priors because JAGS does not support flat priors. With the identity link, the Gaussian intercept prior is centered on `round(median(y), 1)` with scale `max(2.5, round(mad(y), 1))`. With the log link, location and scale are derived from $\log(y)$ (with zeros replaced by 0.1).
- **Slopes:** Numeric coefficient priors use the corresponding family-level scale divided by a representative change in the predictor: its observed range `max(x) - min(x)` for `par_x` and binary variables, and two standard deviations ($2\,\text{SD}$) otherwise. This follows the scaling convention proposed by [Gelman (2008)](https://doi.org/10.1002/sim.3107) and ensures that slope priors remain identical and comparable across models regardless of the number of change points.
- **Unspecified SD model (`sigma`):** For Gaussian models without an explicit `sigma()` formula, `sigma_1` defaults to `dt(0, max(2.5, round(mad(y), 1)), 3) T(0, )`, matching `brms`. This remains true for `gaussian(link = "log")`: that link constrains the conditional mean, not the observations, so non-positive responses are valid and are not passed through `log()` merely to construct a sigma prior.
- **Modeled SD (`sigma(1 + x)`):** An explicit `sigma()` formula models log-SD, as in `brms`. Its intercept prior is `dt(0, 2.5, 3)`, and its contrast and numeric-coefficient priors use the same reference-change scaling described above.

# Default priors on group-level effects

Each group-level effect has a population-level SD parameter. Predictor-side deviations use a zero-mean normal distribution governed by that SD, as in ordinary multilevel regression. With a multi-coefficient `||` term, every independent coefficient has its own SD. The SD receives a positive, weakly informative prior from the family and link specification. For distributional parameters, the deviations and their SD are on that parameter's link scale; for example, effects inside an explicit `sigma()` formula are on the log-SD scale.

Group-level change-point deviations are exactly zero-centered so that each population-level change point is the arithmetic mean of its group-specific locations. Their default priors use sequential bounds, and the model constrains the resulting locations to remain ordered within each group. All group-level change points in one model must use the same grouping factor. Predictor-side deviations are hierarchically mean-zero but are neither exactly centered nor ordered. See [group-level effects with mcp](../articles/group_effects.html).

# Prior predictive checks

Prior predictive checks are a great way to ensure that the priors are meaningful. Simply set `sample = "prior"`. Let us do it for the two sets of priors defined previously in this article, to see their different prior predictive space.

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
# Sample priors 
fit_pp_manual = mcp(model, data = df, prior = prior, sample = "prior", seed = 42)
fit_pp_default = mcp(model, data = df, sample = "prior", seed = 42)

# Plot it
set.seed(42)
plot_pp_manual = plot(fit_pp_manual, lines = 100) + ylim(c(-400, 400)) + ggtitle("Manual prior")
set.seed(42)
plot_pp_default = plot(fit_pp_default, lines = 100) + ylim(c(-400, 400)) + ggtitle("Default prior")
plot_pp_manual +  plot_pp_default  # using patchwork
```

You can see how the manual priors are more dense to the left, and the "concerted" change at x = 80.

# JAGS code

Here is the JAGS code for `fit_manual`:

```{r}
fit_manual$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
