---
title: "Working with priors"
date: "`r format(file.mtime(knitr::current_input()), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: none
---

*NOTE: the JAGS-specific prior interface below is likely to be deprecated in the future, when additional backends are added beyond JAGS.*

# Setting a prior

`mcp` takes priors in the form of a named list. The names are the parameter names, and the values are JAGS code. Here is a fairly complicated example, to demonstrate the various ways priors can be used:

```{r}
model = list(
  y ~ 1 + x,  # Intercept_1 + x_1
  ~ 1 + x,  # cp_1, Intercept_2, and x_2
  ~ 1 + x  # cp_2
)

prior = list(
  Intercept_1 = "dnorm(0, 5) T(, 10)",  # Intercept; less than 10
  x_1 = "dbeta(2, 5)",  # slope: beta with right skew
  cp_1 = "dunif(min(x), cp_2)",  # change point between smallest x and cp_2
  x_2 = "dt(0, 1, 3) T(x_1, )",  # slope 2 > slope 1 and t-distributed
  cp_2 = 80,  # A constant (set; not estimated)
  x_3 = "x_2"  # continue same slope
  # Intercept_2 and Intercept_3 not specified. Use default.
)
```

The values use mcp's JAGS-string syntax, so JAGS distributions are generally allowed. See the [JAGS user manual](https://altushost-swe.dl.sourceforge.net/project/mcmc-jags/Manuals/4.x/jags_user_manual.pdf) for more details. Prior strings use R-like conventional distribution scales even though JAGS has a different parameterization underneath: SD for `dnorm()`, scale for `dt()`, `ddexp()`, and `dlogis()`, and log-SD for `dlnorm()`. When generating JAGS code, `mcp` converts these to inverse variance for `dnorm()`, `dt()`, and `dlnorm()`, and inverse scale for `ddexp()` and `dlogis()`. You can inspect the translation by comparing `fit$prior` with `fit$jags_code`.

Other notes:

- Default population-level change-point priors are ordered. For user priors, `mcp` adds truncation (e.g., `T(cp_1, )`) only when the prior has neither explicit truncation nor an inherently bounded form such as `dunif()` or `dirichlet()`. User-specified population-level bounds are respected as written. When change points have group-level deviations, their realized locations are constrained to remain ordered within every group.

- `mcp` allows inserting data-dependent values: for example, `min(x)`, `max(x)`, `median(y)`, `mad(y)`, `max(x) - min(x)`, `n_segments()`, and `n_cp()`. `mcp` resolves these expressions before generating JAGS code. The older `mcp` v0.3.4 uppercase constants (`MINX`, `MAXX`, and so on) remain temporarily supported with a deprecation warning.

- You can fix any parameter to a specific value. Simply set it to a numerical value (as `cp_2` above). A constant is a 100% prior belief in that value, and it will therefore not be estimated.

- You can also equate one variable with another (`x_3 = "x_2"` above). You would usually do this to share parameters across segments, but you can be creative and do something like `x_3 = "x_2 + 5 - cp_1/10"` if you want. In any case, it will lead to one less parameter being estimated, i.e., one less free parameter.

Let us see the priors after running them through `mcp` and compare to the default priors:

```{r, R.options = list(width = 9999)}
library(mcp)
df_dummy = data.frame(x = 1:100, y = 1:100)
empty_manual = mcp(model, data = df_dummy, prior = prior, sample = FALSE)
empty_default = mcp(model, data = df_dummy, sample = FALSE)

# Inspect resolved priors and bounds
prior_summary(empty_manual)  # For more details: add verbose = TRUE
prior_summary(empty_default)  # For more details: add verbose = TRUE
```

Now, let's simulate some data from `model`. The following priors are "at odds" with the actual data so as to show their effect.

```{r}
df = data.frame(x = runif(200, 0, 100), y = 1)  # 200 datapoints between 0 and 100
set.seed(42)
df$y = empty_default$simulate(empty_default, df, 
    Intercept_1 = 20, Intercept_2 = 22, Intercept_3 = 30,  # intercepts
    x_1 = -0.5, x_2 = 0.5, x_3 = 0,  # slopes
    cp_1 = 35, cp_2 = 70,  # change points
    sigma = 5)

head(df)
```

Sample the prior and posterior. We give the manual fit a longer warmup since it is harder to find the right posterior under these weird prior constraints (priors will usually improve sampling efficiency).

```{r, cache = TRUE, message=FALSE, warning=FALSE, results=FALSE}
future::plan(future::multisession, workers = 3)
fit_manual = mcp(model, data = df, sample = "both", warmup = 10000, prior = prior, seed = 42)
fit_default = mcp(model, data = df, sample = "both", warmup = 10000, seed = 42)
```

First, let's look at the priors side by side. Notice the use of `prior = TRUE` to show prior samples. This works in `plot()`, `plot_pars()`, and `summary()` among others.

```{r, fig.height=12, fig.width=6}
library(ggplot2)
set.seed(42)
pp_default = plot_pars(fit_default, type = "dens_overlay", prior = TRUE, nvariables = NULL) + 
  ggtitle("Default priors")

set.seed(42)
pp_manual = plot_pars(fit_manual, type = "dens_overlay", prior = TRUE, nvariables = NULL) +
  ggtitle("Manual priors")

pp_default + pp_manual
```

Here are the resulting posterior fits:

```{r}
set.seed(42)
plot_default = plot(fit_default) + ggtitle("Default priors")

set.seed(42)
plot_manual = plot(fit_manual) + ggtitle("Manual priors")

plot_default + plot_manual
```

We see the effects of the priors:

- The intercept `Intercept_1` was truncated to be below 10.
- The slope `x_1` is bound to be non-negative (because `dbeta`).
- The slopes `x_2` and `x_3` were forced to be identical.
- The change point `cp_2` was a constant, so there is no uncertainty there.

Adding `verbose = TRUE` reveals full details including the rule descriptions, parameter classifications (`kind`), and whether each prior came from the user or data-calibrated defaults:

```{r, R.options = list(width = 9999)}
prior_summary(fit_manual, verbose = TRUE)
```

This is a contrived example. Usually setting priors manually aims to sample the "correct" posterior.

# Default priors on change points {#cp_prior}

Change point locations must lie within the observed range `[min(x), max(x)]` and maintain strict order `cp_1 < cp_2 < ... < cp_n`. Without this rule, change point locations would be unidentifiable and label switching occur.

The default prior on all change points in `mcp` is **uniform order statistics** (`dirichlet(1)`), which treats any set of ordered change points as equally likely anywhere in the observed interval `[min(x), max(x)]`. See below for a visualization. Some properties of this prior:

- **Single change point:** For 1 change point, this reduces to the uniform prior in the observed range: $\text{cp}_1 \sim \text{Beta}(1, 1) \equiv \text{Uniform}(\min(x), \max(x))$.
- **Multiple change points:** For $N \ge 2$, change points are evenly spaced on average: $$\mathbb{E}[\text{cp}_j] = \min(x) + \frac{j}{N + 1} (\max(x) - \min(x))$$ For example, with 2 change points, prior means are at 33% and 67% of the data range. With 5 change points, they are at 16.7%, 33.3%, 50%, 66.7%, and 83.3%.
- **Symmetry and flat sum:** The prior is symmetric under reflection (flipping $x$ leaves prior probabilities unchanged), and the sum of prior densities across all change points is uniform across the entire data range.
- **Increasing informativeness with segment count:** For any given range, introducing more change points makes the prior for each change point more effectively informative because they are "squeezed" more together (as shown in the figure below).

The plot below compares these theoretical Dirichlet distributions with the sampled prior distributions for 2 and 5 change points:

```{r, echo = FALSE, results=FALSE, warning=FALSE, message=FALSE}
df_dummy = data.frame(x = c(0, 0.5, 1), y = rnorm(3))

# Sample priors for two default change points
model_two = list(y ~ x, ~ 1, ~ 1)
fit_two = mcp(model_two, data = df_dummy, sample = "prior", chains = 1, iter = 50000, seed = 42)

# Sample priors for five default change points
model_five = list(y ~ x, ~ 1, ~ 1, ~ 1, ~ 1, ~ 1)
fit_five = mcp(model_five, data = df_dummy, sample = "prior", chains = 1, iter = 50000, seed = 42)
```

```{r, echo = FALSE, fig.height=3.5, fig.width=7}
par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
MINX = min(df_dummy$x)
MAXX = max(df_dummy$x)
col_emp = rgb(1, 0, 0)

# Panel 1: Two change points
curve(dbeta(x, 1, 2) / 2, from = 0, to = 1, ylim = c(0, 1.15),
      ylab = "Density", xlab = "x", main = "Two change points",
      col = "black", lwd = 1.5)
curve(dbeta(x, 2, 1) / 2, add = TRUE, col = "black", lwd = 2)
curve(dunif(x, 0, 1), lty = 2, add = TRUE, col = "black", lwd = 2)

draws_two = as_draws_df(fit_two)
y1 = density(draws_two$cp_1, from = MINX, to = MAXX)
y2 = density(draws_two$cp_2, from = MINX, to = MAXX)
ysum = (y1$y + y2$y) / 2

lines(y1$x, y1$y / 2, col = col_emp)
lines(y2$x, y2$y / 2, col = col_emp)
lines(y1$x, ysum, col = col_emp, lty = 2)

text(0.12, 0.75, labels = "cp_1")
text(0.88, 0.75, labels = "cp_2")
text(0.5, 1.05, labels = "sum")

# Panel 2: Five change points
curve(dbeta(x, 1, 5) / 5, from = 0, to = 1, ylim = c(0, 1.15),
      ylab = "Density", xlab = "x", main = "Five change points",
      col = "black", lwd = 1.5)
curve(dbeta(x, 2, 4) / 5, add = TRUE, col = "black", lwd = 2)
curve(dbeta(x, 3, 3) / 5, add = TRUE, col = "black", lwd = 2)
curve(dbeta(x, 4, 2) / 5, add = TRUE, col = "black", lwd = 2)
curve(dbeta(x, 5, 1) / 5, add = TRUE, col = "black", lwd = 2)
curve(dunif(x, 0, 1), lty = 2, add = TRUE, col = "black", lwd = 2)

draws_five = as_draws_df(fit_five)
z1 = density(draws_five$cp_1, from = MINX, to = MAXX)
z2 = density(draws_five$cp_2, from = MINX, to = MAXX)
z3 = density(draws_five$cp_3, from = MINX, to = MAXX)
z4 = density(draws_five$cp_4, from = MINX, to = MAXX)
z5 = density(draws_five$cp_5, from = MINX, to = MAXX)
zsum = (z1$y + z2$y + z3$y + z4$y + z5$y) / 5

lines(z1$x, z1$y / 5, col = col_emp)
lines(z2$x, z2$y / 5, col = col_emp)
lines(z3$x, z3$y / 5, col = col_emp)
lines(z4$x, z4$y / 5, col = col_emp)
lines(z5$x, z5$y / 5, col = col_emp)
lines(z1$x, zsum, col = col_emp, lty = 2)

text(0.1, 0.85, labels = "cp_1")
text(0.25, 0.53, labels = "cp_2")
text(0.5, 0.48, labels = "cp_3")
text(0.75, 0.53, labels = "cp_4")
text(0.9, 0.85, labels = "cp_5")
text(0.5, 1.05, labels = "sum")

par(mfrow = c(1, 1))
```

Note on implementation: Under the hood, `mcp` parameterizes this using a sequential stick-breaking Beta chain ($z_j \sim \text{Beta}(1, N - j + 1)$, $\text{cp}_j = \text{cp}_{j-1} + z_j(\max(x) - \text{cp}_{j-1})$). This allows JAGS to sample the model much more efficiently.

The [Dirichlet distribution](https://en.wikipedia.org/wiki/Dirichlet_distribution) places a joint prior on positive segment spacings that sum to the observed x-range. This construction also underlies monotonic effects in `brms` ([Bürkner & Charpentier (2019)](https://psyarxiv.com/9qkhj/)).

## Regularizing change points: `dirichlet(alpha)`

To adjust the spacing regularity of change points, specify `dirichlet(alpha)` for all change points:

```{r, eval=FALSE}
prior = list(
  cp_1 = "dirichlet(2)",
  cp_2 = "dirichlet(2)"
)
```

The number in parentheses is a common $\alpha$ concentration parameter (all change points must share the same $\alpha$):

- $\alpha = 1$ (default): completely flat uniform order statistics as shown above.
- $\alpha > 1$ (e.g. 2, 5): penalizes change points from occurring close to each other or the boundaries, favoring more evenly spaced segments.
- $\alpha < 1$ (e.g. 0.5): favors clustering and permits change points to occur close together.

## Manual priors on change points

You can customize change point priors as shown in the initial example. Beware that the nature of the priors changes when truncation is applied; use `plot_pars(fit, prior = TRUE)` to inspect the realized prior distributions.

For an otherwise unbounded user prior such as `cp_2 = "dnorm(40, 10)"`, `mcp` automatically adds default bounds equivalent to `cp_2 = "dnorm(40, 10) T(cp_1, max(x))"`. An explicit truncation, including `T(,)`, or a `dunif()` prior is used as written. Relaxing the order can cause label-switching between segments and make sampling difficult, so inspect the resulting prior carefully.

# Default priors on linear predictors

The defaults borrow the robust calibration used by `brms`, adapted to the local parameterization of `mcp`. You can inspect all resolved priors for a model using `prior_summary(fit, verbose = TRUE)`.

- **Intercepts:** Unlike the improper flat default coefficient priors in `brms`, `mcp` uses proper Student-t priors because JAGS does not support flat priors. With the identity link, the Gaussian intercept prior is centered on `round(median(y), 1)` with scale `max(2.5, round(mad(y), 1))`. With the log link, location and scale are derived from $\log(y)$ (with zeros replaced by 0.1).
- **Slopes:** Numeric coefficient priors use the corresponding family-level scale divided by a representative change in the predictor: its observed range `max(x) - min(x)` for `par_x` and binary variables, and two standard deviations ($2\,\text{SD}$) otherwise. This follows the scaling convention proposed by [Gelman (2008)](https://doi.org/10.1002/sim.3107) and ensures that slope priors remain identical and comparable across models regardless of the number of change points.
- **Unspecified SD model (`sigma`):** For Gaussian models without an explicit `sigma()` formula, `sigma_1` defaults to `dt(0, max(2.5, round(mad(y), 1)), 3) T(0, )`, matching `brms`. This remains true for `gaussian(link = "log")`: that link constrains the conditional mean, not the observations, so non-positive responses are valid and are not passed through `log()` merely to construct a sigma prior.
- **Modeled SD (`sigma(1 + x)`):** An explicit `sigma()` formula models log-SD, as in `brms`. Its intercept prior is `dt(0, 2.5, 3)`, and its contrast and numeric-coefficient priors use the same reference-change scaling described above.

# Default priors on group-level effects

Each group-level effect has a population-level SD parameter. Predictor-side deviations use a zero-mean normal distribution governed by that SD, as in ordinary multilevel regression. With a multi-coefficient `||` term, every independent coefficient has its own SD. The SD receives a positive, weakly informative prior from the family and link specification. For distributional parameters, the deviations and their SD are on that parameter's link scale; for example, effects inside an explicit `sigma()` formula are on the log-SD scale.

Group-level change-point deviations are exactly zero-centered so that each population-level change point is the arithmetic mean of its group-specific locations. Their default priors use sequential bounds, and the model constrains the resulting locations to remain ordered within each group; `cp_i_sd` estimates the sample standard deviation ($N - 1$ degrees of freedom) across these group locations. All group-level change points in one model must use the same grouping factor. Predictor-side deviations are hierarchically mean-zero but are neither exactly centered nor ordered. See [group-level effects with mcp](../articles/group_effects.html).

# Prior predictive checks

Prior predictive checks are a great way to ensure that the priors are meaningful. Simply set `sample = "prior"`. Let us do it for the two sets of priors defined previously in this article, to see their different prior predictive space.

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
# Sample priors 
fit_pp_manual = mcp(model, data = df, prior = prior, sample = "prior", seed = 42)
fit_pp_default = mcp(model, data = df, sample = "prior", seed = 42)

# Plot it
set.seed(42)
plot_pp_manual = plot(fit_pp_manual, lines = 100) + ylim(c(-400, 400)) + ggtitle("Manual prior")
set.seed(42)
plot_pp_default = plot(fit_pp_default, lines = 100) + ylim(c(-400, 400)) + ggtitle("Default prior")
plot_pp_manual +  plot_pp_default  # using patchwork
```

You can see how the manual priors are more dense to the left, and the "concerted" change at x = 80.

# JAGS code

Here is the JAGS code for `fit_manual`:

```{r}
fit_manual$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
