---
title: "Working with priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Setting a prior
`mcp` takes priors in the form of a named list. The names are the parameter names, and the values are JAGS code. Here is a fairly complicated example, just to get enough priors to demonstrate the various ways priors can be used:

```{r}
model = list(
  y ~ 1 + x,  # Intercept_1 + x_1
  ~ 1 + x,  # cp_1, Intercept_2, and x_2
  ~ 1 + x  # cp_2
)

prior = list(
  Intercept_1 = "dnorm(0, 5) T(, 10)",  # Intercept; less than 10
  x_1 = "dbeta(2, 5)",  # slope: beta with right skew
  cp_1 = "dunif(min(x), cp_2)",  # change point between smallest x and cp_2
  x_2 = "dt(0, 1, 3) T(x_1, )",  # slope 2 > slope 1 and t-distributed
  cp_2 = 80,  # A constant (set; not estimated)
  x_3 = "x_2"  # continue same slope
  # Intercept_2 and Intercept_3 not specified. Use default.
)
```

The values are JAGS code, so all JAGS distributions are allowed. These also include `gamma`, `dt`, `cauchy`, and many others. See the [JAGS user manual](https://web.sgh.waw.pl/~atoroj/ekonometria_bayesowska/jags_user_manual.pdf) for more details. The parameterization of the distributions are identical to standard R. Use SD when you specify priors for `dnorm`, `dt`, `dlogis`, etc. `mcp` converts to precision for JAGS under the hood via the `sd_to_prec()` function (`prec = 1 / sd^2`), so you don't have to worry about it. You can see the effects of this conversion by inspecting the difference between `fit$prior` (using SD) and `fit$jags_code` (using precision).

Other notes:

* Default population-level change-point priors are ordered. For user priors, `mcp` adds truncation (e.g., `T(cp_1, )`) only when the prior has neither explicit truncation nor an inherently bounded form such as `dunif()` or `dirichlet()`. User-specified population-level bounds are respected as written. When change points have group-level deviations, their realized locations are constrained to remain ordered within every group.

* Data-dependent values can be written directly in priors: for example, `min(x)`, `max(x)`, `median(y)`, `mad(y)`, `max(x) - min(x)`, `segment_width(x)`, `n_segments()`, and `n_cp()`. `mcp` resolves these expressions before generating JAGS code. The older uppercase constants (`MINX`, `MAXX`, and so on) remain temporarily supported with a deprecation warning.

* You can fix any parameter to a specific value. Simply set it to a numerical value (as `cp_2` above). A constant is a 100% prior belief in that value, and it will therefore not be estimated.

* You can also equate one variable with another (`x_3 = "x_2"` above). You would usually do this to share parameters across segments, but you can be creative and do something like `x_3 = "x_2 + 5 - cp_1/10"` if you want. In any case, it will lead to one less parameter being estimated, i.e., one less free parameter.

Let us see the priors after running them through `mcp` and compare to the default priors:

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
df_dummy = data.frame(x = 1:100, y = 1:100)
empty_manual = mcp(model, data = df_dummy, prior = prior, sample = FALSE)
empty_default = mcp(model, data = df_dummy, sample = FALSE)

# Inspect resolved priors and bounds. Add rules, descriptions, sources, and
# kinds with prior_summary(empty_manual, verbose = TRUE).
prior_summary(empty_manual)
prior_summary(empty_default)
```

Now, let's simulate some data that from the `model`. The following priors are "at odds" with the actual data so as to show their effect.

```{r}
df = data.frame(x = runif(200, 0, 100), y = 1)  # 200 datapoints between 0 and 100
df$y = empty_default$simulate(empty_default, df, 
    Intercept_1 = 20, Intercept_2 = 30, Intercept_3 = 30,  # intercepts
    x_1 = -0.5, x_2 = 0.5, x_3 = 0,  # slopes
    cp_1 = 35, cp_2 = 70,  # change points
    sigma = 5)

head(df)
```


Sample the prior and posterior. We give the manual fit a longer warmup, since it is harder to find the right posterior under these weird prior constraints (priors will usually improve sampling efficiency).

```{r, cache = TRUE, message=FALSE, warning=FALSE, results=FALSE}
fit_manual = mcp(model, data = df, sample = "both", warmup = 10000, prior = prior)
fit_default = mcp(model, data = df, sample = "both", warmup = 10000)
```

First, let's look at the priors side by side. Notice the use of `prior = TRUE` to show prior samples. This works in `plot()`, `plot_pars()`, and `summary()` among others.

```{r, fig.height=12, fig.width=6}
library(ggplot2)
pp_default = plot_pars(fit_default, type = "dens_overlay", prior = TRUE) + 
  ggtitle("Default priors")

pp_manual = plot_pars(fit_manual, type = "dens_overlay", prior = TRUE) +
  ggtitle("Manual priors")

pp_default + pp_manual
```

Here is the resulting posterior fits:

```{r}
plot_default = plot(fit_default) + ggtitle("Default priors")
plot_manual = plot(fit_manual) + ggtitle("Manual priors")

plot_default + plot_manual
```

We see the effects of the priors.

 * The intercept `Intercept_1` was truncated to be below 10.
 * The slope `x_1` is bound to be non-negative (because `dbeta`).
 * The slopes `x_2` and `x_3` were forced to be identical.
 * The change point `cp_2` was a constant, so there is no uncertainty there.

Of course, it will usually be the other way around: setting priors manually will often serve to sample the "correct" posterior.


# Default priors on change points {#cp_prior}
The following are treated more formally in the [mcp paper](https://osf.io/preprints/fzqxv/).

The default priors order the population-level change points from left (`cp_1`) to right (`cp_2+`). User-supplied bounds are used as written, but all sampled population- and group-level change points are checked for ordering; population-level change points are also checked against the observed x-range. `mcp` offers two built-in population-level prior families with different tradeoffs:

* **Regularizing t-tail prior:** The default for multiple change points uses sequentially truncated Student-t distributions centered at the minimum x. It samples efficiently, and is reasonably close to the dirichlet priors (see below), but its segment spacings are not exchangeable and it can regularize the middle spacings downward.

* **Symmetric segment spacings:** `dirichlet(1)` is equivalent to sorting independent uniforms and treats the spacings exchangeably. This is a transparent substantive prior, not an uninformative one, but JAGS samples it much less efficiently than the t-tail prior. It follows [Bürkner & Charpentier (2019)](https://psyarxiv.com/9qkhj/).

With one change point, both specifications imply a uniform location within the observed range, though the direct uniform default samples more efficiently.


```{r, echo = FALSE, results=FALSE, warning=FALSE, message=FALSE}
df_dummy = data.frame(x = c(50, 100, 150), y = rnorm(3))

# Sample priors for two default change points
model_two = list(y ~ x, ~ 1, ~1)
fit_two = mcp(model_two, data = df_dummy, sample = "prior", chains = 1, iter = 300000)

# Sample priors for five default change points
model_five = list(y ~ x, ~ 1, ~1, ~1, ~1, ~1)
fit_five = mcp(model_five, data = df_dummy, sample = "prior", chains = 1, iter = 300000)
```

```{r, echo = FALSE, fig.height=6, fig.width=7}
par(mfrow = c(2,2))
MINX = min(df_dummy$x)
MAXX = max(df_dummy$x)


###############
# DIRICH TWO #
###############

# Initiate for cp_1
curve(dbeta(x, 1, 2) / 2, 
      ylab = "density",
      xlab = NA,
      main = "Dirichlet two")

text(0.1, 0.7, labels = "cp_1")

# Add others
curve(dbeta(x, 2, 1) / 2, add = T); text(0.9, 0.7, labels = "cp_2")

# The sum
curve(dunif(x, 0, 1), lty = 2, add = T); text(0.5, 0.95, labels = "sum")

# Add extra x-axis
axis(1, at = seq(0, 1, by = 0.2), labels = seq(MINX, MAXX, by = 20), line=3)
mtext("Beta", 1, line = 1, at = -0.2)
mtext("Data", 1, line = 3, at = -0.2)


###############
# DIRICH FIVE #
###############

# Initiate for cp_1
curve(dbeta(x, 1, 5) / 5, 
      ylab = "density",
      xlab = NA,
      main = "Dirichlet five")

text(0.1, 0.8, labels = "cp_1")

# Add others
curve(dbeta(x, 2, 4) / 5, add = T); text(0.25, 0.48, labels = "cp_2")
curve(dbeta(x, 3, 3) / 5, add = T); text(0.5, 0.43, labels = "cp_3") 
curve(dbeta(x, 4, 2) / 5, add = T); text(0.75, 0.48, labels = "cp_4") 
curve(dbeta(x, 5, 1) / 5, add = T); text(0.9, 0.8, labels = "cp_5") 

# The sum
curve(dunif(x, 0, 1), lty = 2, add = T); text(0.5, 0.95, labels = "sum")

# Add extra x-axis
axis(1, at = seq(0, 1, by = 0.2), labels = seq(MINX, MAXX, by = 20), line=3)
mtext("Beta", 1, line = 1, at = -0.2)
mtext("Data", 1, line = 3, at = -0.2)


###############
# DEFAULT TWO #
###############
draws_two = as_draws_df(fit_two)
y1 = density(draws_two$cp_1, from = MINX, to = MAXX)
y2 = density(draws_two$cp_2, from = MINX, to = MAXX)
plot(y1$x, y1$y + y2$y, 
     type = "l", lty = 2, 
     xlab = NA, ylab = "Density", main = "t-tail two",
     xlim = c(MINX, MAXX), ylim = c(0, 0.024))

lines(y1)
lines(y2)

text(100, 0.018, labels = "sum")
text(65, 0.011, labels = "cp_1")
text(130, 0.011, labels = "cp_2")



################
# DEFAULT FIVE #
################
draws_five = as_draws_df(fit_five)
z1 = density(draws_five$cp_1, from = MINX, to = MAXX)
z2 = density(draws_five$cp_2, from = MINX, to = MAXX)
z3 = density(draws_five$cp_3, from = MINX, to = MAXX)
z4 = density(draws_five$cp_4, from = MINX, to = MAXX)
z5 = density(draws_five$cp_5, from = MINX, to = MAXX)

plot(z1$x, (z1$y + z2$y + z3$y + z4$y + z5$y), 
     type = "l", lty = 2, 
     xlab = NA, ylab = "Density", main = "t-tail five",
     xlim = c(MINX, MAXX), ylim = c(0, 0.07)); 
text(100, 0.043, labels = "sum")

lines(z5); text(137, 0.03, labels = "cp_5")
lines(z4); text(115, 0.020, labels = "cp_4")
lines(z3); text(94, 0.023, labels = "cp_3")
lines(z3); text(75, 0.027, labels = "cp_2")
lines(z2); text(62, 0.041, labels = "cp_1")
lines(z1)

par(mfrow = c(1,1))
```



## The t-tail prior on 2+ change points (default)

With exactly one change point, the default rule is `cp_1 = uniform(min = min(x), max = max(x))`. Thus its location is uniform over the observed x-range. This is equivalent to the one-change-point Dirichlet specification.

For 2+ change points, the default rule (on *all* change points) is `cp_i = student_t(df = n_cp() - 1, location = min(x), scale = (max(x) - min(x)) / n_cp())`, bounded below by the preceding change point and above by `max(x)`. This is not as complicated as it looks, so let me unpack it.

 * It is t-distributed with $N - 1$ degrees of freedom (`n_cp() - 1`). Together with the decreasing scale, this makes the underlying distributions more concentrated as the number of change points increases.
 * Under the defaults, it is bounded below by the preceding population-level `cp` and above by `max(x)`. For example, `cp_3` is bounded below by `cp_2` and above by `max(x)`, keeping the default population-level change points ordered within the observed range.
 * The scale is `(max(x) - min(x)) / n_cp()`. This is a calibration choice, not the equal-segment spacing (which would divide by `n_cp() + 1`) and not the Student-t standard deviation; for low degrees of freedom the untruncated variance is not finite.
 * The location parameter of the underlying Student-t is always the lowest observed `x`. Thus `cp_1` is a bounded half-t and `cp_2+` use sequentially truncated right tails of the same t, hence the name "t-tail prior". Their means are shifted by truncation and are not equal to `min(x)`.

The sequential truncation does not produce uniform order statistics. The middle segments tend get narrower than the first/last segments. For non-small datasets, the effect on the posterior is negligible, though point-`hypothesis()` tests via Savage-Dickey would be strongly impacted.


## Dirichlet-based prior on change points
The [Dirichlet distribution](https://en.wikipedia.org/wiki/Dirichlet_distribution) places a joint prior on positive segment spacings that sum to the observed x-range. Consequently, the derived population-level change points are bounded and ordered. Equal concentration parameters make the spacings exchangeable, but do not make the prior uninformative: they encode a specific joint distribution over segment widths. This construction also underlies monotonic effects in `brms` ([Bürkner & Charpentier (2019)](https://psyarxiv.com/9qkhj/)).

To use the Dirichlet prior, you need to specify it for all or none of the change points. E.g.,

```{r, eval=FALSE}
prior_dirichlet = list(
  cp_1 = "dirichlet(1)",
  cp_2 = "dirichlet(1)",
  cp_3 = "dirichlet(1)"
)
```

The number in the parenthesis is the $\alpha$ parameter, so you could also specify `cp_1 = "dirichlet(3)` if you want to push credence for that and later change points more to the rightwards while pushing earlier priors leftwards.


## Manual priors on change points
You can easily change or modify change point priors, just as we did in the initial example. But beware that the nature of the priors change when truncation is applied. Use `plot_pars(fit, prior = TRUE)` to check the resulting prior.

For an otherwise unbounded user prior such as `cp_2 = "dnorm(40, 10)"`, `mcp` adds default bounds equivalent to `cp_2 = "dnorm(40, 10) T(cp_1, max(x))"`. An explicit truncation, including `T(,)`, or a `dunif()` prior is used as written. Relaxing the order can make segment labels ambiguous and sampling difficult, so inspect the resulting prior carefully.


# Default priors on linear predictors
The defaults borrow the robust calibration used by `brms`, adapted to the local parameterization of `mcp`. Intercepts use Student-t priors centered on a robust, link-appropriate location. Data-calibrated scales are at least 2.5, avoiding implausibly narrow defaults when the response happens to have little variation. Unlike the flat default coefficient priors in `brms`, `mcp` uses proper Student-t priors because change-point models need regularization for stable sampling and prior prediction.

Numeric coefficient priors use the corresponding family-level scale divided by a representative change in the model-matrix column: its observed range when it has two values, and two standard deviations otherwise. This is the scaling convention proposed by [Gelman (2008)](https://doi.org/10.1002/sim.3107). Transformations and interactions are treated as model-matrix columns in their own right. Treatment-coded categorical contrasts keep the family-level scale.

For terms involving local `par_x`, the representative change also includes the expected segment width, `(max(x) - min(x)) / n_segments()`, or the corresponding power for polynomial terms. Thus the prior concerns the amount of change across a typical segment rather than depending arbitrarily on the units or number of change points.

## Gaussian mean priors

With the identity link, the Gaussian intercept prior is centered on `round(median(y), 1)` and uses scale `max(2.5, round(mad(y), 1))`. Contrasts and numeric coefficients use the same scale and the reference-change scaling above.

With the log link, this calibration uses the log response, replacing zero by 0.1 and using finite fallbacks. Both intercept rules follow `brms`; `mcp` adds proper scaled coefficient priors for prior sampling and mild segment regularization.

## Gaussian residual-SD priors

For Gaussian models without an explicit `sigma()` formula, the residual-SD prior is calibrated from `mad(y)` on the response scale. This remains true for `gaussian(link = "log")`: that link constrains the conditional mean, not the observations, so non-positive responses are valid and are never passed through `log()` merely to construct a sigma prior.

An explicit `sigma()` formula instead models log-SD, as in `brms`. Its intercept prior is `dt(0, 2.5, 3)`, and its proper contrast and numeric-coefficient priors use the same reference-change scaling described above. These proper coefficient priors intentionally differ from the improper flat defaults in `brms` so that prior-only sampling remains defined.

See the family-specific articles for more information about the priors for other families:

 * `vignette("binomial")` - also relevant for `bernoulli`
 * `vignette("poisson")` - also relevant for negative-binomial `shape`
 * `vignette("variance")` for explicit `sigma()` models
 * `vignette("arma")` for AR/MA regularization and root conditions

The general principles belong here: parameterization, data-based scaling, proper regularization, and prior predictive checking. Exact defaults and constraints that arise from a particular likelihood or model component are documented in its own article. `prior_summary(fit, verbose = TRUE)` is the authoritative model-specific inventory, so the articles do not need to duplicate every generated prior.


# Default priors on group-level effects

Each group-level effect has a population-level SD parameter. Predictor-side deviations use a zero-mean normal distribution governed by that SD, as in ordinary multilevel regression. With a multi-coefficient `||` term, every independent coefficient has its own SD. The SD receives a positive, weakly informative prior from the family and link specification. For distributional parameters, the deviations and their SD are on that parameter's link scale; for example, effects inside an explicit `sigma()` formula are on the log-SD scale.

Group-level change-point deviations are exactly zero-centered so that each population-level change point is the arithmetic mean of its group-specific locations. Their default priors use sequential bounds, and the model constrains the resulting locations to remain ordered within each group. All group-level change points in one model must use the same grouping factor. Predictor-side deviations are hierarchically mean-zero but are neither exactly centered nor ordered. See [group-level effects with mcp](https://lindeloev.github.io/mcp/articles/varying.html).


# Prior predictive checks
Prior predictive checks is a great way to ensure that the priors are meaningful. Simply set `sample = "prior"`. Let us do it for the two sets of priors defined previously in this article, to see their different prior predictive space.

```{r, cache = TRUE, results= FALSE, message=FALSE, warning=FALSE}
# Sample priors 
fit_pp_manual = mcp(model, data = df, prior = prior, sample = "prior")
fit_pp_default = mcp(model, data = df, sample = "prior")

# Plot it
plot_pp_manual = plot(fit_pp_manual, lines = 100) + ylim(c(-400, 400)) + ggtitle("Manual prior")
plot_pp_default = plot(fit_pp_default, lines = 100) + ylim(c(-400, 400)) + ggtitle("Default prior")
plot_pp_manual +  plot_pp_default  # using patchwork
```

You can see how the manual priors are more dense to the left, and the "concerted" change at x = 80.


# JAGS code
Here is the JAGS code for `fit_manual`:

```{r}
fit_manual$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
