---
title: "Modeling variance and variance changes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling variance and variance changes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For GLM families with a variance parameter (`sigma`), you can model this explicitly. For example, if you want a flat mean (a plateau) with increasing variance, you can do `y ~ 1 + sigma(1 + x)`. In general, all formula syntax that is allowed outside of `sigma()` (where it applies to the mean) also works inside it (applying to the variance). For example, if you go all-out, you can do `~ 1 + sigma(rel(1) + sin(x) + I(x^2))`. Read more about mcp formulas [here](../articles/formulas.html). 

```{r}
library(mcp)
options(mc.cores = 3)  # Speed up sampling
set.seed(42)  # Make the script deterministic
```


# Simple example: a change in variance
Let us model a simple change in variance on a plateau model. First, we specify the model:

```{r}
model = list(
  y ~ 1,  # sigma(1) is implicit in the first segment
  ~ 0 + sigma(1)  # a new intercept on sigma, but not on the mean
)
```

We can simulate some data, starting with low variance and an abrupt change to a high variance at $x = 50$:

```{r}
empty = mcp(model, sample = FALSE, par_x = "x")
set.seed(40)
df = data.frame(x = 1:100)
df$y = empty$simulate(
  df$x, 
  cp_1 = 50, int_1 = 20, 
  sigma_1 = 5, sigma_2 = 20)

head(df)
```


Now we fit the model to the simulated data.

```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, par_x = "x")
```
We plot the results with the prediction interval to show the effect of the variance, since it won't be immediately obvious on the default plot of the fitted mean predictions:

```{r}
plot(fit, q_predict = TRUE)
```


We can see all parameters are well recovered (compare `sim` to `mean`). Like the other parameters, the `sigma`s are named after the segment where they were instantiated. There will always be a `sigma_1`.

```{r}
summary(fit)
```


# Advanced example
We can model changes in `sigma` alone or in combination with changes in the mean. In the following, I define a needlessly complex model, just to demonstrate the flexibility of modeling variance:

```{r}
model = list(
  # Increasing variance.
  y ~ 1 + sigma(1 + x),
  
  # Abrupt change in mean and variance.
  ~ 1 + sigma(1),
  
  # Joined slope on mean. variance changes as 2nd order poly.
  ~ 0 + x + sigma(0 + x + I(x^2)),
  
  # Continue slope on mean, but plateau variance (no sigma() tern).
  ~ 0 + x
)

# The slope in segment 4 is just a continuation of 
# the slope in segment 3, as if there was no change point.
prior = list(
  x_4 = "x_3"
)
```

Notice a few things here:

 * Segment 3 and 4: I changed the variance on a continuous slope. You can do this using priors to define that the slope is shared between segment 3 and 4, effectively canceling the change point on the mean (more about using priors in mcp [here](../articles/priors.html)).
 * Segment 4: By not specifying `sigma()`, segment 4 (and later segments) just inherits the variance from the state it was left in in the previous segment. 

In general, the variance parameters are named `sigma_[normalname]`, where "normalname" is the usual parameter names in mcp (see more [here](../articles/formulas.html)). For example, the variance slope on `x` in segment 3 is `sigma_x_3`. However, `sigma_int_i` is just too verbose, so variance intercepts are simply called `sigma_i`, where i is the segment number.



## Simulate data
We simulate some data from this model, setting all parameters. As always, we can fit an empty model to get `fit$simulate`, which is useful for simulation and predictions from this model.

```{r}
empty = mcp(model, sample = FALSE)
set.seed(40)
df = data.frame(x = 1:200)
df$y = empty$simulate(
  df$x,
  cp_1 = 50, cp_2 = 100, cp_3 = 150,
  int_1 = -20, int_2 = 0,
  sigma_1 = 3, sigma_x_1 = 0.5,
  sigma_2 = 10, 
  sigma_x_3 = -0.5,
  sigma_x_3_E2 = 0.02,
  x_3 = 1, x_4 = 1)
```


## Fit it and inspect results
Fit it in parallel, to speed things up:

```{r, cache = TRUE, results = FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df, prior = prior)
```

Plotting the prediction interval is an intuitive way to to see how the variance is estimated:

```{r}
plot(fit, q_predict = TRUE)
```


We can also plot the `sigma_` parameters directly. Now the y-axis is `sigma`:

```{r}
plot_dpar(fit, dpar = "sigma", q_fit = TRUE)
```


`summary()` show that the parameters are well recovered (compare `sim` to `mean`). The last change point is estimated with greater uncertainty than the others. This is expected, given that the only "signal" of this change point is a stop in variance growth.

```{r}
summary(fit)
```

The effective sample size (`n.eff`) is fairly low, indicating poor mixing for these parameters. `Rhat` is acceptable at < 1.1, indicating good convergence between chains. Let us verify this by taking a look at the posteriors and trace. For now, we just look at the sigmas:

```{r, fig.height=7, fig.width = 6}
plot_pars(fit, regex_pars = "sigma_")
```

This confirms the impression from `Rhat` and `n.eff`. Setting `mcp(..., iter = 10000)` would be advisable to increase the effective sample size. Read more about [tips, tricks, and debugging](../articles/tips.html).




# Varying change points and variance
The variance model applies to varying change points as well. For example, here we do a spin on the example in [the article on varying change points](../articles/varying.html), and add a by-person change in `sigma`. We model two joined slopes, varying by `id`. The second slope is also characterized by a different variance. This means that the model has more information about when the change point occurs, so it should be easier to estimate (require fewer data).

```{r}
model = list(
  # intercept + slope
  y ~ 1 + x,
  
  # joined slope and increase in variance, varying by id.
  1 + (1|id) ~ 0 + x + sigma(1)
)
```

Simulate data:
```{r}
empty = mcp(model, sample = FALSE)
set.seed(40)
df = data.frame(
  x = 1:180,
  id = rep(1:6, times = 30)
)
df$y = empty$simulate(
  df$x, 
  cp_1 = 70, cp_1_id = 15 * (df$id - mean(df$id)),
  int_1 = 20, x_1 = 1, x_2 = -0.5,
  sigma_1 = 10, sigma_2 = 25)
```

Fit it:
```{r, cache = TRUE, results=FALSE, warning=FALSE, message=FALSE}
fit = mcp(model, data = df)
```

Plot it:
```{r}
plot(fit, facet_by = "id")
```


As usual, we can get the individual change points:
```{r}
ranef(fit)
```

