---
title: "Modeling variance and variance changes in mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling variance and variance changes in mcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For GLM families with a variance parameter (`sigma`), you can model this explicitly. For example, if you want a flat mean (a plateau) with increasing variance, you can do `y ~ 1 + sigma(1 + x)`. In general, all formula syntax that is allowed outside of `sigma()` (where it applies to the mean) also works inside it (applying to the variance). For example, if you go all-out, you can do `~ 1 + sigma(rel(1) + sin(x) + I(x^2))`. Read more about mcp formulas [here](../articles/formulas.html). 


# Simple example: a change in variance
Let us model a simple change in variance on a plateau model. First, we specify the model:

```{r}
segments = list(
  y ~ 1,  # sigma(1) is implicit in the first segment
  ~ 0 + sigma(1)  # a new intercept on sigma, but not on the mean
)
```

We can simulate some data, starting with low variance and an abrupt change to a high variance at $x = 50$:

```{r}
library(mcp)
empty = mcp(segments, sample = FALSE, par_x = "x")
set.seed(40)
data = tibble::tibble(
  x = 1:100,
  y = empty$func_y(x, cp_1 = 50, int_1 = 20, 
                   sigma_1 = 5, sigma_2 = 20)
)
```


Now we fit the model to the simulated data.

```{r, cache = TRUE, results=FALSE}
fit = mcp(segments, data, par_x = "x")
```
We plot the results with the prediction interval to show the effect of the variance, since it won't be immediately obvious on the default plot of the fitted mean predictions:

```{r}
plot(fit, quantiles = TRUE, predict = "predict", lines = 0)
```


We can see all parameters are well recovered. Like the other parameters, the `sigma`s are named after the segment where they were instantiated. There will always be a `sigma_1`.

```{r}
summary(fit)
```


# Advanced example
We can model changes in `sigma` alone or in combination with changes in the mean. In the following, I define a needlessly complex model, just to demonstrate the flexibility of modeling variance:

```{r}
segments = list(
  # Increasing variance,
  y ~ 1 + sigma(1 + x),
  
  # Abrupt change in mean and variance (relative to last segment)
  ~ 1 + sigma(1),
  
  # Joined slope on mean. variance changes quadratically
  ~ 0 + x + sigma(0 + x + I(x^2)),
  
  ~ 0 + x  # Continue slope, but plateau variance
)

# The slope in segment 4 is just a continuation of 
# the slope in segment 3, as if there was no change point.
prior = list(
  x_4 = "x_3"
)
```

Notice a few things here:

 * I changed the variance on a mean-slope. You can do this using priors to define that the slope is shared between segment 3 and 4, effectively cancelling the change point on the mean (more about using priors in mcp [here](../articles/priors.html)). 
 * By not specifying `sigma()`, segment 4 (and later segments) just inherits the variance from the state it was left in in the previous segment. 
 * There is an exponential growth of variance in segment 3, and at the change point to segment 4, this just stops and plateaus for the remainder of segment 4. If we had specified a different variance in segment 4, this would, of course, override it, just like specifying an intercept on the mean would.

In general, the variance parameters are named `sigma_[normalname]`, where "normalname" is the usual parameter names in mcp (see more [here](../articles/formulas.html)). For example, the variance slope on `x` in segment 3 is `sigma_x_3`. However, `sigma_int_i` is just too verbose, so variance intercepts are simply called `sigma_i`, where i is the segment number.

## Simulate data
We simulate some data, setting all parameters:

```{r}
empty = mcp(segments, sample = FALSE)
set.seed(40)
data = tibble::tibble(
  x = 1:200,
  y = empty$func_y(x,
    cp_1 = 50, cp_2 = 100, cp_3 = 150,
    int_1 = -20, int_2 = 0,
    sigma_1 = 3, sigma_x_1 = 0.5,
    sigma_2 = 10, 
    sigma_x_3 = -0.5,
    sigma_xE2_3 = 0.02,
    x_3 = 1, x_4 = 1)
)
```


## Fit it and inspect results
Fit it in parallel, to speed things up:

```{r, cache = TRUE, results = FALSE}
fit = mcp(segments, data, prior, cores = 3)
```

Again, it is easiest to see the effect of variance changes by plotting the prediction intervals:

```{r}
plot(fit, quantiles = TRUE, predict = "predict", lines = 0)
```

The parameters are well recovered. There is some uncertainty about the last change point, but this is expected, given that the only "sigmal" that a change happened is that variance stopped growing.

```{r}
summary(fit)
```

The effective sample size (`n.eff`) is fairly low for some of the parameters. Let us take a look at the posteriors and trace plots for a more intuitive inspection whether there is a convergence problem. For now, we just look at the sigmas:

```{r, fig.height=7, fig.width = 6}
plot(fit, "combo", regex_pars = "sigma_")
```

This confirms that sampling is ineffective (low `n.eff` = poor mixing) but that convergence is good (`Rhat` < 1.1). Setting `mcp(..., iter = 10000)` would be advisable to increase the effective sample size. Read more about [tips, tricks, and debugging](../articles/debug.html).
