---
title: "Varying change points with mcp"
author: "Jonas Kristoffer Lindeløv"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Varying change points with mcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A unique feature of `mcp` is modeling change points as varying effects (sometimes called "random effects"). This has the advantage that you can let the change point vary by a factor while keeping other parameters common across varying factor levels.

This article in brief:

 * How to simulate varying change points
 * Get posteriors using `ranef(fit)`
 * Plot using `plot(fit, facet_by="my_group")` and `plot_pars(fit, pars = "varying", type = "dens_overlay", ncol = 3)`.
 * The default priors restrict varying change points to lie between the two adjacent change points.
 * The article on modeling variance via `sigma()` contains [an example on varying change points](../articles/variance.html) as well.



# Specifying varying change points

You specify varying effects using the classical [`lmer`](https://www.rdocumentation.org/packages/lme4/versions/1.1-21/topics/lmer) syntax `(1|group)`. Currently (v. 0.1) `mcp` only support varying intercepts. For example, here we model a varying change point between a plateau and a joined slope:

```{r}
model = list(
  y ~ 1,  # int_1
  1 + (1|id) ~ 0 + x  # cp_1, cp_1_sd, cp_1_id[i]
)
```

You can have multiple varying change points with multiple groupings:

```{r}
model = list(
  y ~ 1,  # int_1
  1 + (1|id) ~ 0 + x,  # cp_1, cp_1_sd, cp_1_id[i]
  1 + (1|species) ~ 0,  # cp_2, cp_2_sd, cp_2_species[i]
  (1|id) ~ 1  # cp_3 (implicit), cp_3_sd, cp_3_id[i]
)
```

Here are some properties of the change point varying effects:

**Zero centered:** The varying effects are zero-centered around the associated group-level change point. In other words, the sum of all varying effects are exactly zero. This constraint is necessary for the parameters to be identifiable. 

**Hierarchical:** Consider the first change point, `cp_1`, and it's associated varying effects, `cp_1_id`. By default, it is modeled as sampled from (nested within) the group-level change point, `cp_1`, as well as a spread, `cp_1_sd`.

**Constraints:** The varying effects are constrained to lie (1) in the observed range of the x-axis, and/or (2) between the two adjacent change points. That is, all `cp_1_id` are between `min(x)` and `cp_2`. All `cp_2_species` are between `cp_1` and `cp_3` and all `cp_3_id` are between `cp_2` and `max(x)`. These constraints are enforced through truncation of the default prior (`fit$prior`) and you can override them by specifying a manual prior (see vignette("priors")).



# Simulating varying effects
Let us do a worked example, simulating the varying change point between a plateau and a slope: 

```{r}
model = list(
  y ~ 1,  # int_1
  1 + (1|id) ~ 0 + x  # cp_1, cp_1_sd, cp_1_id[i]
)
```

It is quite similar to simulating non-varying data, except that we need to simulate some varying offsets before passing all parameters to `empty$simulate`:

```{r, message=FALSE, warning=FALSE}
library(mcp)
empty = mcp(model, sample = FALSE)

library(dplyr, warn.conflicts = FALSE)
varying = c("Clark", "Louis", "Batman", "Batgirl", "Spiderman", "Jane")
df = data.frame(
  x = runif(length(varying) * 30, 0, 100),  # 30 data points for each
  id = rep(varying, each = 30)  # the group names
)
df$id_numeric = as.numeric(as.factor(df$id))  # to positive integers
df$y = empty$simulate(df$x,
  # Population-level:
  int_1 = 20, x_2 = 0.5, cp_1 = 50, sigma = 2,
  
  # Varying: zero-centered and 10 between each level
  cp_1_id = 10 * (df$id_numeric - mean(df$id_numeric)))

head(df)
```

Here, we "translated" the `id` to an offset on the x-axis by multiplying with 10. We subtracted the mean to make the varying effects zero-centered around `cp_1`. The result:

```{r}
library(ggplot2)
ggplot(df, aes(x=x, y=y)) + 
  geom_point() +
  facet_wrap(~id)
```


# Summarise and plot varying effects.
Fitting the model is simple:

```{r, cache = TRUE, message = FALSE, warning=FALSE, results=FALSE}
fit = mcp(model, data = df)
```

If we just use `plot(fit)`, we would see all points in one plot. We want to facet by `id`, so:

```{r}
plot(fit, facet_by="id", ncol = 3)
```

It seems that `mcp` did a good job of recovering the change points. There is a lot of information in this data, since the intercept and the slope on each side of the (varying) change point is shared between participants here.

If you use `summary(fit)` (or `fixef(fit)`) you will get the posteriors for the population-level effects. To get the random effects, do:

```{r}
ranef(fit)
```

Inspecting the `sim` and `match` columns, we see that they recovered the simulation parameters well.

Good convergence is not always as obvious as in this example. While `plot_pars(fit)` show population-level parameters only, you can do this to get varying effects only:

```{r}
plot_pars(fit, pars = "varying", type = "trace", ncol=3)
```

Notice the use of the `ncol` argument to set the number of columns. You will often have *many* levels on your varying effect, so this is useful to get a good view of all of them. Naturally, you can do this for almost all kinds of plots.

Using `pars = "varying"` will plot all varying effects. This may be too much if you have multiple varying effects. To select just one, use regular expression in `regex_pars`. Two very handy operators are "^" (begins with) and "$" (ends with). Just to show that this "faceting" works for almost all of the many plot types, we now do two columns of `"dens_overlay`:

```{r}
plot_pars(fit, regex_pars = "^cp_1_id", type = "dens_overlay", ncol = 2)
```




# Priors for varying effects
You can see the priors of the model like this:

```{r}
cbind(fit$prior)
```

The priors `cp_1_sd` is the population-level standard deviation of `cp_1_id`, the latter of which is applied to all levels of `id`. This is also apparent if you inspect the JAGS code for this model. The truncation of varying effects is quite contrived, but just keeps them between the two adjacent (population-level) change points.


# JAGS code
Here is the JAGS code for the model used in this article:

```{r}
cat(fit$jags_code)
```

