---
title: "Group-level effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group-level effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: none
---

`mcp` supports group-level effects (also called random effects) in both the predictor and change-point parts of a formula. The syntax follows `lme4` and `brms`: `(1|group)` specifies a group-level intercept, `(factor||group)` specifies independent group coefficients, `fixef()` reports population-level effects, and `ranef()` reports group-level deviations.

This article in brief:

 * Predictor and change-point group-level effects
 * How group-level effects persist across segments
 * How to simulate group-level change-point deviations
 * Get posteriors using `ranef(fit)`
 * Plot using `plot(fit, facet_by="my_group")` and `plot_pars(fit, pars = "group", type = "dens_overlay", ncol = 3)`.
 * The default priors bound group-specific change points relative to adjacent population-level change points.
 * The article on modeling standard deviation via `sigma()` contains [another group-level change-point example](../articles/dpar.html).

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic
```


# Specifying group-level change points

You specify group-level effects using the familiar [`lmer`](https://www.rdocumentation.org/packages/lme4/versions/1.1-21/topics/lmer) and `brms` syntax `(1|group)`. In the cp part, this models a group-specific deviation from the population-level change point. For example:

```{r}
model = list(
  y ~ 1,  # Intercept_1
  1 + (1|id) ~ 0 + x  # cp_1, cp_1_sd, cp_1_id[i]
)
```

You can have multiple group-level change points, but they must use the same grouping factor so their realized locations can be ordered within each group:

```{r}
model = list(
  y ~ 1,  # Intercept_1
  1 + (1|id) ~ 0 + x,  # cp_1, cp_1_sd, cp_1_id[i]
  1 + (1|id) ~ 0,      # cp_2, cp_2_sd, cp_2_id[i]
  (1|id) ~ 1           # cp_3 (implicit), cp_3_sd, cp_3_id[i]
)
```

Here are some properties of group-level change-point deviations:

**Exactly zero centered:** The group-level deviations sum to zero around the associated population-level change point. Besides making `cp_i` the arithmetic mean of the group-specific change points, this avoids a poorly identified translation between `cp_i` and the average group deviation.

**Hierarchical:** Consider the population-level change point `cp_1` and its associated group-level deviations `cp_1_id`. Their spread is governed by the population-level parameter `cp_1_sd`.

**Constraint:** Realized group-specific change points must remain strictly ordered within each group.

Unlike predictor group-level effects, a change-point group effect applies only to the boundary where it is written. There is no carry-over as there is for group-level effects on RHS.

# Predictor group-level effects

The same syntax in the predictor part gives each group a deviation from the population-level intercept:

```{r}
model = list(
  y ~ 1 + (1|id),
  ~ 0 + x,
  ~ 1 + (0|id)
)
```

Here, the group-level intercept introduced in segment 1 also applies in segment 2. A later `(1|id)` would replace it from that segment onward, while `(0|id)` turns it off, as in segment 3 above. Persistence is tracked separately for each grouping factor and distributional parameter.

Use `||` for independent group-level slopes and factor coefficients:

```{r}
model = list(
  y ~ 1 + condition + (condition||id),
  ~ 0 + x
)
```

With the default treatment coding, `(condition||id)` contains a group-level intercept and one group-level deviation for each non-reference contrast of `condition`. `(0 + condition||id)` instead gives each factor level its own group-level coefficient. Similarly, `(1 + z||id)` specifies independent group-level intercepts and slopes on a numeric predictor `z`. Each coefficient has its own population-level SD.

The population-level and group-level formulas need not contain the same coefficients. For example, `y ~ 1 + (0 + condition||id)` has a population-level intercept but only group-specific condition coefficients.

Group-level intercepts also work inside distributional formulas:

```{r}
model = list(
  y ~ 1 + (1|id) + sigma(1 + (1|id))
)
```

This model has group-level deviations in both the conditional mean and log-SD. The `||` syntax works inside distributional formulas too, for example `sigma(1 + (condition||id))`.

A later predictor group-level term for the same distributional parameter and grouping factor replaces the entire earlier block. Thus, if `(condition||id)` is followed in a later segment by `(1||id)`, only the new intercept deviation applies from that segment onward. `(0|id)` turns the block off.

Multi-coefficient terms with `|`, such as `(1 + x|id)`, would imply correlated group coefficients and are not yet supported; use `(1 + x||id)` for independent coefficients. Group-level terms inside `ar()` and `ma()` are also not supported.

Unlike change-point deviations, predictor deviations are not constrained to sum exactly to zero. They have an ordinary mean-zero hierarchical normal distribution, matching multilevel regression in `lme4` and `brms`. For example, `(1|id)` in segment 1 creates the deviation vector `Intercept_1_id` and its population-level SD parameter `Intercept_1_id_sd`. `(condition||id)` additionally creates names such as `conditionB_1_id` and `conditionB_1_id_sd`; inside `sigma()`, the corresponding names start with `sigma`.



# Simulating group-level change points
Let us simulate group-specific deviations in the change point between a plateau and a slope:

```{r}
model = list(
  y ~ 1,  # Intercept_1
  1 + (1|id) ~ 0 + x  # cp_1, cp_1_sd, cp_1_id[i]
)
```

This follows the same interface as predictor group-level effects: supply the population-level SD and `fit$simulate()` draws one deviation per group and centers them exactly on zero.

```{r, message=FALSE, warning=FALSE}
library(dplyr, warn.conflicts = FALSE)
group_levels = c("Clark", "Louis", "Batman", "Batgirl", "Spiderman", "Jane")
df = data.frame(
  x = runif(length(group_levels) * 30, 0, 100),  # 30 data points for each
  id = rep(group_levels, each = 30),  # the group names
  y = 1
)
empty = mcp(model, data = df, sample = FALSE)

set.seed(42)
df$y = empty$simulate(empty, df,
  # Population-level:
  Intercept_1 = 20, x_2 = 0.5, cp_1 = 50, sigma = 2,
  
  # Generate exactly centered group-level deviations with this SD
  cp_1_sd = 15)

head(df)
```

For models with multiple group-level change points, all deviations are drawn once and then checked for ordering. An unordered draw produces an error; `fit$simulate()` does not redraw, sort, or clip the change points. The result:

```{r}
library(ggplot2)
ggplot(df, aes(x=x, y=y)) + 
  geom_point() +
  facet_wrap(~id)
```


# Summarise and plot group-level effects
Fitting the model is simple:

```{r, cache = TRUE, message = FALSE, warning=FALSE, results=FALSE}
fit = mcp(model, data = df, seed = 42)
```

If we just use `plot(fit)`, we would see all points in one plot. We want to facet by `id`, so:

```{r}
set.seed(42)
plot(fit, facet_by = "id")
```

It seems that `mcp` recovered the group-specific change points well. There is a lot of information in these data because the population-level intercept and slopes on either side of the change point are shared across participants (`id`).

`summary(fit)` (or `fixef(fit)`) returns posterior summaries for the population-level effects. To get the group-level deviations (random effects), use:

```{r}
mcp::ranef(fit)
```

Inspecting the `sim` and `match` columns, we see that they recovered the simulation parameters well.

Prediction methods include all group-level effects by default. Set `varying = FALSE` for population-only predictions, `varying = "cp"` for all group-level effects in the cp part, `varying = "predictor"` for predictor-side group-level effects, or supply an exact name such as `varying = "cp_1_id"`. `ranef(fit)` deliberately remains simple and returns all group-level effects.

Good convergence is not always as obvious as in this example. While `plot_pars(fit)` shows population-level parameters only, you can select the group-level deviations with `"group"`:

```{r}
set.seed(42)
plot_pars(fit, pars = "group", type = "trace", ncol = 3)
```

The `ncol` argument controls the number of columns. Group-level effects often have many levels, so this is useful for viewing all deviations.

Using `pars = "group"` plots all group-level deviations. To select one group-level effect, use a regular expression in `regex_pars`; for example, `^` anchors the start of a parameter name:

```{r}
set.seed(42)
plot_pars(fit, regex_pars = "^cp_1_id", type = "dens_overlay", ncol = 2)
```

You can also do posterior predictive checking with facets. I think that for the relatively univariate models supported as of `mcp` 0.3, this does not add much new information over and above `plot(fit, facet_by = "id")`, but it's a standard assessment that many will be acquainted with:

```{r}
set.seed(42)
pp_check(fit, facet_by = "id")
```




# Priors for group-level effects
You can see the priors of the model like this:

```{r}
cbind(fit$prior)
```

The prior `cp_1_sd` is the population-level standard deviation governing the `cp_1_id` deviations across levels of `id`. Change-point deviations are exactly zero-centered, while predictor group-level effects use ordinary hierarchical mean-zero deviations. Their realized locations are constrained to remain ordered within each group.


# JAGS code
Here is the JAGS code for the model used in this article:

```{r}
fit$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
