---
title: "Group-level effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group-level effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`mcp` supports group-level effects (also called varying or random effects) in both the predictor and change-point parts of a formula. The syntax and terminology follow `lme4` and `brms`: `(1|group)` specifies a group-level intercept, `(factor||group)` specifies independent group coefficients, `fixef()` reports population-level coefficients, and `ranef()` reports group-level deviations.

This article in brief:

 * Predictor and change-point group-level effects
 * How group-level effects persist across segments
 * How to simulate varying change points
 * Get posteriors using `ranef(fit)`
 * Plot using `plot(fit, facet_by="my_group")` and `plot_pars(fit, pars = "varying", type = "dens_overlay", ncol = 3)`.
 * The default priors restrict varying change points to lie between the two adjacent change points.
 * The article on modeling variance via `sigma()` contains [an example on varying change points](../articles/variance.html) as well.

```{r}
library(mcp)
future::plan(future::multisession, workers = 3)
set.seed(42)  # Make the script deterministic
```


# Specifying varying change points

You specify group-level effects using the familiar [`lmer`](https://www.rdocumentation.org/packages/lme4/versions/1.1-21/topics/lmer) and `brms` syntax `(1|group)`. In the cp part, this models a varying change point. For example, here we model a varying change point between a plateau and a joined slope:

```{r}
model = list(
  y ~ 1,  # int_1
  1 + (1|id) ~ 0 + x  # cp_1, cp_1_sd, cp_1_id[i]
)
```

You can have multiple varying change points with multiple groupings:

```{r}
model = list(
  y ~ 1,  # int_1
  1 + (1|id) ~ 0 + x,  # cp_1, cp_1_sd, cp_1_id[i]
  1 + (1|species) ~ 0,  # cp_2, cp_2_sd, cp_2_species[i]
  (1|id) ~ 1  # cp_3 (implicit), cp_3_sd, cp_3_id[i]
)
```

Here are some properties of the change point varying effects:

**Zero centered:** The group-level deviations are zero-centered around the associated population-level change point. In other words, the sum of all varying effects is exactly zero. This constraint is necessary for the parameters to be identifiable.

**Hierarchical:** Consider the population-level change point `cp_1` and its associated group-level deviations `cp_1_id`. Their spread is governed by the population-level parameter `cp_1_sd`.

**Constraints:** The varying effects are constrained to lie (1) in the observed range of the x-axis, and/or (2) between the two adjacent change points. That is, all `cp_1_id` are between `min(x)` and `cp_2`. All `cp_2_species` are between `cp_1` and `cp_3` and all `cp_3_id` are between `cp_2` and `max(x)`. These constraints are enforced through truncation of the default prior (`fit$prior`) and you can override them by specifying a manual prior (see vignette("priors")).

# Predictor group-level effects

The same syntax in the predictor part gives each group a deviation from the population intercept:

```{r}
model = list(
  y ~ 1 + (1|id),
  ~ 0 + x,
  ~ 1 + (0|id)
)
```

Here, the group-level intercept introduced in segment 1 also applies in segment 2. A later `(1|id)` would replace it from that segment onward, while `(0|id)` turns it off, as in segment 3 above. Persistence is tracked separately for each grouping factor and distributional parameter.

Use `||` for independent group-level slopes and factor coefficients:

```{r}
model = list(
  y ~ 1 + condition + (condition||id),
  ~ 0 + x
)
```

With the default treatment coding, `(condition||id)` contains a group-level intercept and one group-level deviation for each non-reference contrast of `condition`. `(0 + condition||id)` instead gives each factor level its own group-level coefficient. Similarly, `(1 + z||id)` specifies independent group-level intercepts and slopes on a numeric predictor `z`. Each coefficient has its own population-level SD.

The population-level and group-level formulas need not contain the same coefficients. For example, `y ~ 1 + (0 + condition||id)` has a population intercept but only group-specific condition coefficients.

Group-level intercepts also work inside distributional formulas:

```{r}
model = list(
  y ~ 1 + (1|id) + sigma(1 + (1|id))
)
```

This model has group-level deviations in both the conditional mean and log-SD. The `||` syntax works inside distributional formulas too, for example `sigma(1 + (condition||id))`.

A later predictor group-level term for the same distributional parameter and grouping factor replaces the entire earlier block. Thus, if `(condition||id)` is followed in a later segment by `(1||id)`, only the new intercept deviation applies from that segment onward. `(0|id)` turns the block off.

Multi-coefficient terms with `|`, such as `(1 + x|id)`, would imply correlated group coefficients and are not yet supported; use `(1 + x||id)` for independent coefficients. Group-level terms inside `ar()` and `ma()` are also not supported.

Unlike change-point deviations, predictor deviations are not constrained to sum exactly to zero. They have an ordinary mean-zero hierarchical normal distribution, matching multilevel regression in `lme4` and `brms`. For example, `(1|id)` in segment 1 creates the deviation vector `Intercept_1_id` and its population-level SD parameter `Intercept_1_id_sd`. `(condition||id)` additionally creates names such as `conditionB_1_id` and `conditionB_1_id_sd`; inside `sigma()`, the corresponding names start with `sigma`.



# Simulating varying change points
Let us do a worked example, simulating the varying change point between a plateau and a slope: 

```{r}
model = list(
  y ~ 1,  # int_1
  1 + (1|id) ~ 0 + x  # cp_1, cp_1_sd, cp_1_id[i]
)
```

It is quite similar to simulating non-varying data, except that we need to simulate some varying offsets before passing all parameters to `empty$simulate`:

```{r, message=FALSE, warning=FALSE}
empty = mcp(model, sample = FALSE)

library(dplyr, warn.conflicts = FALSE)
varying = c("Clark", "Louis", "Batman", "Batgirl", "Spiderman", "Jane")
df = data.frame(
  x = runif(length(varying) * 30, 0, 100),  # 30 data points for each
  id = rep(varying, each = 30)  # the group names
)
df$id_numeric = as.numeric(as.factor(df$id))  # to positive integers
df$y = empty$simulate(df$x,
  # Population-level:
  int_1 = 20, x_2 = 0.5, cp_1 = 50, sigma = 2,
  
  # Varying: zero-centered and 10 between each level
  cp_1_id = 10 * (df$id_numeric - mean(df$id_numeric)))

head(df)
```

Here, we "translated" the `id` to an offset on the x-axis by multiplying with 10. We subtracted the mean to make the varying effects zero-centered around `cp_1`. The result:

```{r}
library(ggplot2)
ggplot(df, aes(x=x, y=y)) + 
  geom_point() +
  facet_wrap(~id)
```


# Summarise and plot varying effects.
Fitting the model is simple:

```{r, cache = TRUE, message = FALSE, warning=FALSE, results=FALSE}
fit = mcp(model, data = df)
```

If we just use `plot(fit)`, we would see all points in one plot. We want to facet by `id`, so:

```{r}
plot(fit, facet_by = "id")
```

It seems that `mcp` did a good job of recovering the change points. There is a lot of information in this data, since the intercept and the slope on each side of the (varying) change point is shared between participants here.

If you use `summary(fit)` (or `fixef(fit)`) you will get the posteriors for the population-level effects. To get the random effects, do:

```{r}
ranef(fit)
```

Inspecting the `sim` and `match` columns, we see that they recovered the simulation parameters well.

Prediction methods include all group-level effects by default. Set
`varying = FALSE` for population-only predictions, `varying = "cp"` for all
group-level effects in the cp part, `varying = "predictor"` for predictor-side
group-level effects, or supply an exact name such as `varying = "cp_1_id"`.
`ranef(fit)` deliberately remains simple and returns all group-level effects.

Good convergence is not always as obvious as in this example. While `plot_pars(fit)` show population-level parameters only, you can do this to get varying effects only:

```{r}
plot_pars(fit, pars = "varying", type = "trace", ncol = 3)
```

Notice the use of the `ncol` argument to set the number of columns. You will often have *many* levels on your varying effect, so this is useful to get a good view of all of them. Naturally, you can do this for almost all kinds of plots.

Using `pars = "varying"` will plot all varying effects. This may be too much if you have multiple varying effects. To select just one, use regular expression in `regex_pars`. Two very handy operators are "^" (begins with) and "$" (ends with). Just to show that this "faceting" works for almost all of the many plot types, we now do two columns of `"dens_overlay`:

```{r}
plot_pars(fit, regex_pars = "^cp_1_id", type = "dens_overlay", ncol = 2)
```

You can also do posterior predictive checking with facets. I think that for the relatively univariate models supported as of `mcp` 0.3, this does not add much new information over and above `plot(fit, facet_by = "id")`, but it's a standard assessment that many will be acquainted with:

```{r}
pp_check(fit, facet_by = "id")
```




# Priors for varying effects
You can see the priors of the model like this:

```{r}
cbind(fit$prior)
```

The prior `cp_1_sd` is the population-level standard deviation of `cp_1_id`, the latter of which is applied to all levels of `id`. Predictor group-level effects similarly receive a population-level SD parameter, but use ordinary mean-zero normal deviations without exact zero-centering. This is also apparent if you inspect the JAGS code for a model. The truncation of change-point deviations keeps them between the two adjacent population-level change points.


# JAGS code
Here is the JAGS code for the model used in this article:

```{r}
fit$jags_code
```

```{r cleanup, include=FALSE}
future::plan(future::sequential)
```
